---
title: "What the Sharpe Ratio Can't See"
subtitle: "Stress-testing our NFL Draft model — and finding where it falls short"
author: "Nick Gurol — [X/Twitter](https://x.com/run_the_sims) | [LinkedIn](https://www.linkedin.com/in/nick-gurol-87738a1b7/)"
output:
  html_document:
    toc: true
    toc_float: true
    self_contained: true
knit: (function(input, ...) rmarkdown::render(input, output_dir = dirname(input), ...))
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, warning = FALSE, message = FALSE,
  fig.width = 8, fig.height = 6, dpi = 150,
  fig.path = "charts/v2_",
  out.width = "100%"
)
library(tidyverse)
library(arrow)
library(scales)
library(gt)
library(nflplotR)
library(nflreadr)

pos_colors <- c(
  "QB" = "#E41A1C", "RB" = "#377EB8", "WR" = "#4DAF4A", "TE" = "#984EA3",
  "OT" = "#FF7F00", "IOL" = "#FFFF33", "EDGE" = "#A65628", "DL" = "#F781BF",
  "LB" = "#999999", "CB" = "#66C2A5", "S" = "#FC8D62"
)

tier_order <- c("Top 10", "Late 1st", "Round 2", "Round 3", "Rounds 4-7")
key_positions <- c("QB", "RB", "WR", "TE", "OT", "IOL", "EDGE", "DL", "LB", "CB", "S")
```

```{r load-data}
# Article 1 data
df <- read_parquet("data/draft_sharpe_analysis.parquet")
sharpe <- read_csv("output/sharpe_ratios.csv", show_col_types = FALSE)
player_lookup <- read_csv("output/player_lookup.csv", show_col_types = FALSE)

sharpe <- sharpe %>% mutate(tier = factor(tier, levels = tier_order))
df <- df %>% mutate(tier = factor(tier, levels = tier_order))
player_lookup <- player_lookup %>% mutate(tier = factor(tier, levels = tier_order))
df_eligible <- df %>% filter(season <= 2022)

# V2 data
class_depth <- read_csv("output/v2/class_depth.csv", show_col_types = FALSE)
blue_chip <- read_csv("output/v2/blue_chip.csv", show_col_types = FALSE)
roster_needs <- read_csv("output/v2/roster_needs.csv", show_col_types = FALSE)
team_efficiency <- read_csv("output/v2/team_efficiency_adjusted.csv", show_col_types = FALSE)
r1_picks <- read_csv("output/v2/r1_picks_need_adjusted.csv", show_col_types = FALSE)
market_efficiency <- read_csv("output/v2/market_efficiency.csv", show_col_types = FALSE)
ngs_prospects <- read_csv("output/v2/ngs_prospects_2025.csv", show_col_types = FALSE)
ngs_historical <- read_csv("output/v2/ngs_historical_matched.csv", show_col_types = FALSE)
ngs_class_tiers <- read_csv("output/v2/ngs_class_tiers.csv", show_col_types = FALSE)

# Factor levels
blue_chip <- blue_chip %>% mutate(tier = factor(tier, levels = tier_order))
r1_picks <- r1_picks %>% mutate(tier = factor(tier, levels = tier_order))
```

## TL;DR

In [Part 1](https://stranger9977.github.io/draft-sharpe-analysis/), we built a model that said **never draft an RB in round 1**. That was the takeaway from our initial analysis. But if we break down the key assumptions and approach this from a different angle, does it make sense to draft an RB early? We stress-test three assumptions: that roster context doesn't matter, that all draft classes are equal, and that snap-based quality tells the whole story. The answer: it depends on the team, the year, and the prospect.

---

## The Model Was Right. And Incomplete.

In [Part 1](https://stranger9977.github.io/draft-sharpe-analysis/), we built a quasi-Sharpe ratio for NFL draft picks. The math was clean. The conclusion was clear: **RB shows negative returns across all tiers -- never draft one in round 1.**

That was the takeaway from our initial analysis. But if we break down some key assumptions and approach this from a different angle, does it make sense to draft an RB early?

GMs don't draft on average. They draft in specific years, with specific rosters, facing specific prospects. A model that can't account for context isn't wrong -- it's incomplete. This is Part 2: an honest stress test that defines where Article 1's advice falls short.

The gap between the model's advice and what teams actually do isn't always inefficiency -- sometimes it's intelligence. We tested three assumptions behind Part 1:

```{r assumptions-table, results='asis'}
cat("
| Assumption | What Could Break It |
|---|---|
| Teams should always draft the highest-Sharpe position | Roster needs make some positions irrelevant for some teams |
| All draft classes are interchangeable | Class depth varies wildly year to year |
| Snaps measure quality — and all prospects are fungible | Career value tells a different story, and blue-chip prospects break the averages |
")
```

---

## Quick Recap: What Part 1 Found

```{r recap-sharpe, results='asis'}
rb_top10 <- sharpe %>% filter(pos_group == "RB", tier == "Top 10")
qb_top10 <- sharpe %>% filter(pos_group == "QB", tier == "Top 10")
best_late1 <- sharpe %>% filter(tier == "Late 1st", !is.na(sharpe_elite)) %>%
  slice_max(sharpe_elite, n = 1)

cat(paste0(
  "The quasi-Sharpe ratio combined three things: rookie contract surplus (how much cheaper ",
  "a drafted starter is than a free agent), second-contract value (the market's verdict), ",
  "and elite probability (the chance of a top-decile outcome). We measured risk-adjusted ",
  "returns by position and draft tier from 2010--2022.\n\n",
  "**Key findings:** QB leads all positions in top-10 Sharpe (",
  round(qb_top10$sharpe_elite, 2), "). OT and EDGE peak in the late first round (",
  round(best_late1$sharpe_elite, 2), " for ", best_late1$pos_group,
  "). RB was the worst first-round investment: Sharpe of **",
  round(rb_top10$sharpe_elite, 2), "** in the top 10, negative across every tier.\n\n",
  "The advice was simple: if you want a running back, wait until Day 2 or sign a free agent. ",
  "The data said so.\n\n",
  "But data without context is just math. Let's break down the assumptions and see where they fall short."
))
```

### How We Measure Quality (and Why It Matters)

This section addresses the third assumption from our table: **snaps measure quality**. Part 1 and Part 2 measure "quality" differently, and that's deliberate -- it's one of the key places where the model's assumptions can be challenged.

**Part 1's "hit"** is snap-based: a player who plays above the positional snap threshold (75% of baseline snaps) earns starter status. Snaps tell you who was on the field. This is how Article 1 builds its Sharpe ratio.

**Part 2's "elite outcome"** uses [career Approximate Value](https://www.pro-football-reference.com/about/approximate_value.htm) (AV) from Pro Football Reference. Career AV is a cumulative, season-by-season measure of a player's total contribution -- it accounts for playing time, statistical production, and team success. A player who starts for 10 years accumulates more AV than one who starts for 3, even if both cleared the snap threshold. We define an elite outcome as finishing in the **top 10% of career AV at their position** (90th percentile). This is a much stricter bar than Part 1's snap-based hit.

Why the distinction? Snaps tell you who became a starter. Career AV tells you who became *great*. A player can clear the snap threshold and still have a mediocre career. When we're asking whether a generational prospect justifies breaking the model, we need the higher bar.

---

## Assumption 1: Teams Should Always Draft the Highest-Sharpe Position

Article 1 implicitly assumed that every team should always draft the highest-Sharpe position available. But what if you already have a starter?

The Bills are the perfect test case. From 2014 through 2022, Buffalo always had someone playing quarterback -- EJ Manuel, Kyle Orton, Tyrod Taylor, Matt Barkley, then Josh Allen. A snap-only model would say "QB is filled every year." But having a warm body at quarterback isn't the same as having a good one.

We added a quality gate: a position is only "filled" when the starter meets three criteria -- plays above the snap threshold, earned a second contract above the free-agent replacement cost for the position, and isn't on an expiring deal. A bad QB who plays a lot doesn't count. Tyrod Taylor started 46 games for the Bills and cleared every snap threshold, but he never earned a QB-level second contract -- the market's way of saying he wasn't the long-term answer.

```{r bills-qb-timeline, fig.height=4, fig.width=8}
bills_qb <- roster_needs %>%
  filter(team == "BUF", pos_group == "QB") %>%
  mutate(
    starter_name = case_when(
      pfr_player_id == "ManuEJ00" ~ "EJ Manuel",
      pfr_player_id == "OrtoKy00" ~ "Kyle Orton",
      pfr_player_id == "TaylTy00" ~ "Tyrod Taylor",
      pfr_player_id == "BarkMa00" ~ "Matt Barkley",
      pfr_player_id == "AlleJo02" ~ "Josh Allen",
      TRUE ~ pfr_player_id
    ),
    fill_label = if_else(is_filled, "Filled", "Need"),
    quality_label = case_when(
      starter_is_good ~ "Quality Starter",
      starter_is_hit & !starter_is_good ~ "Playing, Not Good Enough",
      TRUE ~ "No Starter"
    )
  )

ggplot(bills_qb, aes(x = draft_season, y = 1)) +
  geom_segment(aes(xend = draft_season, y = 0.6, yend = 1), color = "gray70", linewidth = 0.8) +
  geom_point(aes(color = is_filled), size = 6) +
  geom_text(aes(label = starter_name), vjust = -1.5, size = 3.5, fontface = "bold") +
  geom_text(aes(label = fill_label), vjust = 2.5, size = 3, color = "gray50") +
  scale_color_manual(values = c("TRUE" = "#2e7d32", "FALSE" = "#c62828"),
                     labels = c("TRUE" = "Position Filled", "FALSE" = "Position Need")) +
  scale_x_continuous(breaks = 2014:2022) +
  coord_cartesian(ylim = c(0.3, 1.7)) +
  labs(
    title = "Buffalo Bills: QB Need vs Filled (2014-2022, Quality Gate)",
    subtitle = "Red = starter lacked quality despite playing | Green = quality starter in place",
    x = NULL, y = NULL, color = NULL
  ) +
  theme_minimal(base_size = 14) +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "top"
  )
```

This is the interesting part. The model says QB has the best top-10 Sharpe. The Bills had a QB on the roster every year. But the quality gate reveals the truth: the model would have been **right** to tell them to draft QB in years where EJ Manuel or Matt Barkley were the answer -- and right to leave the position alone once Josh Allen arrived. The model's general advice ("draft QB") was correct in spirit, even when the roster looked superficially filled.

### How Need-Adjustment Changes the Rankings

In Part 1, we ranked teams by "draft efficiency" -- the average Sharpe ratio of their first-round picks. A team that spent top-10 picks on high-Sharpe positions (QB, OT, EDGE) scored well; a team that spent them on low-Sharpe positions (RB, S) scored poorly. It's a clean metric, but it ignores one obvious thing: whether the team actually needed a player at that position.

The need-adjusted version gives partial credit to teams that drafted low-Sharpe positions to fill genuine roster holes, and penalizes teams that drafted high-Sharpe positions they already had covered. The chart shows every team's original rank vs their need-adjusted rank.

```{r team-efficiency-slope, fig.height=10, fig.width=8}
te_plot <- team_efficiency %>%
  mutate(
    direction = case_when(
      rank_change > 2 ~ "Improved",
      rank_change < -2 ~ "Declined",
      TRUE ~ "Stable"
    )
  )

ggplot(te_plot, aes(y = reorder(team, -rank_adjusted))) +
  geom_segment(
    aes(x = rank_original, xend = rank_adjusted,
        yend = reorder(team, -rank_adjusted),
        color = direction),
    linewidth = 1.2, alpha = 0.7
  ) +
  geom_nfl_logos(aes(x = rank_adjusted, team_abbr = team), width = 0.04) +
  geom_point(aes(x = rank_original), shape = 1, size = 3, color = "gray50") +
  scale_color_manual(
    values = c("Improved" = "#2e7d32", "Declined" = "#c62828", "Stable" = "gray60"),
    name = "After Need Adjustment"
  ) +
  scale_x_reverse(breaks = seq(0, 32, 4)) +
  labs(
    title = "Draft Efficiency: Original Rank vs Need-Adjusted Rank",
    subtitle = "Open circle = original rank | Logo = need-adjusted rank | Left is better",
    x = "Rank (lower = better)", y = NULL
  ) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "top")
```

```{r top-movers, results='asis'}
biggest_up <- team_efficiency %>% slice_max(rank_change, n = 1) %>% slice_head(n = 1)
biggest_down <- team_efficiency %>% slice_min(rank_change, n = 1) %>% slice_head(n = 1)

cat(paste0(
  "**Biggest climber:** ", biggest_up$team, " moved up ", biggest_up$rank_change,
  " spots — they addressed real needs with their first-round picks. ",
  "**Biggest faller:** ", biggest_down$team, " dropped ", abs(biggest_down$rank_change),
  " spots — they kept drafting positions they already had filled."
))
```

### Need-Based Picks vs Filled-Position Picks

Here's the intuition: if a team drafts a position they actually need, the player steps into a real role. If they draft a position that's already filled, the player competes for scraps. Does that show up in the Sharpe numbers?

```{r need-vs-filled-box, fig.height=5, fig.width=8}
r1_context <- r1_picks %>%
  mutate(
    context = if_else(is_need, "Addressed a Need", "Position Already Filled")
  )

ggplot(r1_context, aes(x = sharpe_adjusted, y = context, fill = context)) +
  geom_boxplot(alpha = 0.7, outlier.shape = 21, outlier.alpha = 0.5, width = 0.5) +
  geom_jitter(height = 0.15, alpha = 0.3, size = 2) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
  scale_fill_manual(values = c("Addressed a Need" = "#2196F3", "Position Already Filled" = "#FF9800")) +
  labs(
    title = "Need-Adjusted Sharpe: Picks That Filled Needs vs Filled Positions",
    subtitle = "Dashed line = break-even | Each dot is a first-round pick (2014-2022)",
    x = "Need-Adjusted Sharpe Ratio", y = NULL, fill = NULL
  ) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none")
```

```{r need-vs-filled-stats, results='asis'}
need_stats <- r1_context %>%
  group_by(context) %>%
  summarise(
    mean_adj = mean(sharpe_adjusted, na.rm = TRUE),
    median_adj = median(sharpe_adjusted, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  )
need_row <- need_stats %>% filter(context == "Addressed a Need")
filled_row <- need_stats %>% filter(context == "Position Already Filled")

cat(paste0(
  "Picks addressing real needs average a need-adjusted Sharpe of **",
  round(need_row$mean_adj, 3), "** (n=", need_row$n,
  "), while picks at already-filled positions average **",
  round(filled_row$mean_adj, 3), "** (n=", filled_row$n,
  "). The gap isn't just theoretical — teams that ignore roster context ",
  "systematically underperform their draft capital."
))
```

**The takeaway:** Picks that fill genuine roster needs perform better. It makes sense — a player drafted into opportunity has a clearer path to snaps, production, and a second contract. The Sharpe ratio tells you which positions have the best *average* returns. But if you already have a franchise QB, that average is irrelevant to you. Roster context turns a general truth into specific advice.

---

## Assumption 2: All Draft Classes Are Interchangeable

The Sharpe ratio averages across 13 draft classes (2010-2022). But class quality varies wildly. If a GM could know whether this year's class is loaded or barren *before* the draft, the model's one-size-fits-all advice starts to crack.

We actually can measure this -- at least partially. NFL Next Gen Stats publishes composite prospect scores before the draft, combining athletic testing, college production, and size. These scores exist before a single snap is played in the NFL.

### Prospective Class Quality: What GMs Can Actually See

NGS uses four tiers: **Below Average** (50-60), **Average** (60-70), **Good** (70-90), and **Elite** (90+). Instead of looking at mean scores (which compress into a narrow band), we count how many prospects in each class grade as Good or better. A class with five Good+ prospects is fundamentally different from one with two, even if their averages are similar.

We'll start with RB -- the position at the center of the Sharpe debate -- then see how the pattern plays out across every position group.

```{r ngs-rb-depth-big, fig.height=5, fig.width=8}
ngs_rb_tiers <- ngs_class_tiers %>%
  filter(pos_group == "RB", season >= 2010, season <= 2025,
         ngs_tier %in% c("Elite", "Good")) %>%
  mutate(ngs_tier = factor(ngs_tier, levels = c("Elite", "Good")))

ggplot(ngs_rb_tiers, aes(x = factor(season), y = n, fill = ngs_tier)) +
  geom_col(width = 0.7, alpha = 0.85) +
  scale_fill_manual(
    values = c("Elite" = "#1565C0", "Good" = "#66BB6A"),
    name = "NGS Tier"
  ) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  labs(
    title = "RB Draft Class Depth: Good and Elite Prospects by Year",
    subtitle = "Only prospects scoring Good (70-90) or Elite (90+) on NGS Combined Score",
    x = NULL, y = "Number of Good+ Prospects"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "top"
  )
```

The variation is striking. Some RB classes are loaded with Good+ talent; others barely register. A GM drafting RB in a deep year is playing a fundamentally different game than one drafting in a thin year -- but the Sharpe ratio treats them identically.

Now the same view across every position:

```{r ngs-all-depth-small, fig.height=12, fig.width=8}
ngs_other_tiers <- ngs_class_tiers %>%
  filter(pos_group %in% key_positions, season >= 2010, season <= 2025,
         ngs_tier %in% c("Elite", "Good")) %>%
  mutate(
    ngs_tier = factor(ngs_tier, levels = c("Elite", "Good")),
    pos_group = factor(pos_group, levels = key_positions)
  )

ggplot(ngs_other_tiers, aes(x = factor(season), y = n, fill = ngs_tier)) +
  geom_col(width = 0.7, alpha = 0.85) +
  scale_fill_manual(
    values = c("Elite" = "#1565C0", "Good" = "#66BB6A"),
    name = "NGS Tier"
  ) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  facet_wrap(~pos_group, scales = "free_y", ncol = 3) +
  labs(
    title = "Draft Class Depth Across All Positions",
    subtitle = "Good (70-90) and Elite (90+) NGS prospects by year | Each position tells a different story",
    x = NULL, y = "Number of Good+ Prospects"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 7),
    legend.position = "top",
    strip.text = element_text(face = "bold", size = 12)
  )
```

```{r ngs-class-narrative, results='asis'}
ngs_rb_class <- class_depth %>%
  filter(pos_group == "RB", !is.na(ngs_n_good_plus))
best_class <- ngs_rb_class %>% slice_max(ngs_n_good_plus, n = 1)
worst_class <- ngs_rb_class %>% slice_min(ngs_n_good_plus, n = 1)

cat(paste0(
  "The **", best_class$season,
  "** RB class had **", best_class$ngs_n_good_plus,
  "** prospects grading Good or better — the deepest pool of pre-draft talent in our sample. ",
  "The **", worst_class$season, "** class had just **", worst_class$ngs_n_good_plus,
  "**. That's the difference between a class where you can find value in any round and one ",
  "where you're chasing a single standout.\n\n",
  "The same pattern plays out at every position. Some years WR is loaded and EDGE is thin. ",
  "Some years the OT class is deep while TE is barren. QB class depth is notoriously spiky -- ",
  "one or two Elite prospects can define the entire class. The Sharpe ratio can't see any of this — ",
  "it averages across all years as if every class is the same. GMs know better."
))
```

```{r class-2025-commentary, results='asis'}
class_2025 <- ngs_class_tiers %>%
  filter(season == 2025, ngs_tier %in% c("Elite", "Good")) %>%
  group_by(pos_group) %>%
  summarise(
    n_good_plus = sum(n),
    n_elite = sum(n[ngs_tier == "Elite"]),
    .groups = "drop"
  ) %>%
  arrange(desc(n_good_plus))

deep_2025 <- class_2025 %>% slice_head(n = 3)
thin_2025 <- class_2025 %>% slice_tail(n = 3)

rb_2025 <- class_2025 %>% filter(pos_group == "RB")
qb_2025 <- class_2025 %>% filter(pos_group == "QB")

cat(paste0(
  "### The 2025 Class at a Glance\n\n",
  "Applying the same framework to the 2025 draft class: **",
  deep_2025$pos_group[1], "** (",  deep_2025$n_good_plus[1], " Good+), **",
  deep_2025$pos_group[2], "** (", deep_2025$n_good_plus[2], "), and **",
  deep_2025$pos_group[3], "** (", deep_2025$n_good_plus[3],
  ") are the deepest classes this year. The thinnest: **RB** (",
  if(nrow(rb_2025) > 0) rb_2025$n_good_plus else "few", " Good+) and **QB** (",
  if(nrow(qb_2025) > 0) qb_2025$n_good_plus else "few", " Good+).\n\n",
  "The RB class stands out as especially thin. Only ",
  if(nrow(rb_2025) > 0) rb_2025$n_good_plus else "a handful of",
  " RB prospects grade as Good or better — and Love is the only one in the Elite tier. ",
  "That 17-point gap to RB2 isn't just a Love story; it's a class story. If you pass on Love, ",
  "the next RB option is a significant downgrade.\n\n",
  "QB is thin too, but for QB the bar is different: Good isn't good enough. ",
  "QB is the one position where only blue-chip prospects (Elite tier) reliably return value in the top 10 — ",
  "a Good QB prospect at pick 5 is a much bigger gamble than a Good OT or EDGE at the same spot. ",
  "For most other positions, Good is deep enough to find value.\n\n",
  "On the other end, **IOL** is loaded — ",
  if(nrow(deep_2025) > 0 && deep_2025$pos_group[1] == "IOL") paste0(deep_2025$n_good_plus[1], " Good+ prospects") else "the deepest position group",
  " give teams plenty of quality options throughout the draft. ",
  "If you need an interior lineman in 2025, the class will come to you.\n\n",
  "This is exactly the kind of information the Sharpe ratio can't incorporate — ",
  "the best position to draft depends on which classes are loaded *this year*."
))
```

### Did the Pre-Draft Signal Match Reality?

The NGS scores tell us what the class looked like before the draft. But did the "deep" classes actually produce more hits? Here's the RB timeline with retrospective hit rates -- bar color is the pre-draft NGS classification, bar height is what actually happened.

```{r rb-class-timeline-big, fig.height=6, fig.width=8}
rb_classes <- class_depth %>%
  filter(pos_group == "RB")

rb_notable <- tribble(
  ~season, ~label,
  2010, "0 hits",
  2011, "Murray",
  2012, "Martin, Morris",
  2013, "Bell, Lacy",
  2014, "Hill, Hyde",
  2015, "Gordon, Gurley,\nDavid Johnson",
  2016, "Zeke, Henry",
  2017, "CMC, Kamara,\nCook, Hunt,\nMixon, Jones",
  2018, "Saquon, Chubb",
  2019, "Jacobs, Sanders",
  2020, "Taylor, Swift",
  2021, "Harris, Javonte",
  2022, "Walker, Hall"
)

rb_classes <- rb_classes %>%
  left_join(rb_notable, by = "season")

ggplot(rb_classes, aes(x = factor(season), y = hit_rate, fill = ngs_class_quality)) +
  geom_col(width = 0.7, alpha = 0.85) +
  geom_text(
    aes(label = label),
    vjust = -0.3, size = 3, lineheight = 0.85
  ) +
  scale_fill_manual(
    values = c("Deep" = "#2e7d32", "Average" = "#FFA726", "Thin" = "#c62828"),
    na.value = "gray50"
  ) +
  scale_y_continuous(
    labels = percent_format(),
    expand = expansion(mult = c(0, 0.25))
  ) +
  labs(
    title = "RB Draft Class Quality: A Rollercoaster",
    subtitle = "Retrospective hit rate by draft year | Bar color = pre-draft NGS class quality",
    x = NULL, y = "Hit Rate", fill = "Class Quality"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "top"
  )
```

```{r rb-timeline-narrative, results='asis'}
ngs_rb_class <- class_depth %>%
  filter(pos_group == "RB", !is.na(ngs_n_good_plus))
ngs_rb_cor <- cor(ngs_rb_class$ngs_n_good_plus, ngs_rb_class$hit_rate, use = "complete.obs")

cat(paste0(
  "The colors (pre-draft class depth) and heights (what actually happened) track each other. ",
  "The correlation between a class's Good+ prospect count and its eventual hit rate is **",
  round(ngs_rb_cor, 2), "**.\n\n",
  "The mismatches are revealing. 2010 had the highest mean NGS score but produced zero hits — ",
  "athletic profiles alone don't guarantee production. Meanwhile, 2017 was the best actual class ",
  "(23.3% hit rate) because six different backs all found NFL roles.\n\n",
  "The model says \"never draft RB.\" The class data says \"it depends on the year.\" ",
  "And the NGS tier breakdown gives GMs a real window into which kind of year it is — ",
  "before the first pick is made."
))
```

The same signal-vs-reality check across other positions:

```{r other-pos-timeline-small, fig.height=12, fig.width=8}
other_classes <- class_depth %>%
  filter(pos_group %in% key_positions, pos_group != "RB",
         !is.na(hit_rate)) %>%
  mutate(ngs_class_quality = replace_na(ngs_class_quality, "No NGS Data"))

ggplot(other_classes, aes(x = factor(season), y = hit_rate, fill = ngs_class_quality)) +
  geom_col(width = 0.7, alpha = 0.85) +
  scale_fill_manual(
    values = c("Deep" = "#2e7d32", "Average" = "#FFA726", "Thin" = "#c62828", "No NGS Data" = "gray50"),
    name = "Pre-Draft Class Quality"
  ) +
  scale_y_continuous(labels = percent_format(), expand = expansion(mult = c(0, 0.15))) +
  facet_wrap(~pos_group, scales = "free_y", ncol = 2) +
  labs(
    title = "Pre-Draft Signal vs Reality: All Positions",
    subtitle = "Retrospective hit rate by year | Bar color = pre-draft NGS class quality (deep/average/thin)",
    x = NULL, y = "Hit Rate"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 7),
    legend.position = "top",
    strip.text = element_text(face = "bold", size = 12)
  )
```

```{r other-timeline-narrative, results='asis'}
# Compute correlation for each position
pos_cors <- class_depth %>%
  filter(pos_group %in% key_positions, !is.na(ngs_n_good_plus), !is.na(hit_rate)) %>%
  group_by(pos_group) %>%
  summarise(
    cor = cor(ngs_n_good_plus, hit_rate, use = "complete.obs"),
    .groups = "drop"
  ) %>%
  arrange(desc(cor))

best_cor <- pos_cors %>% slice_head(n = 1)
worst_cor <- pos_cors %>% slice_tail(n = 1)

cat(paste0(
  "The pre-draft signal isn't equally predictive everywhere. **", best_cor$pos_group,
  "** shows the strongest correlation (r = ", round(best_cor$cor, 2),
  ") between Good+ prospect count and eventual hit rate, while **", worst_cor$pos_group,
  "** shows the weakest (r = ", round(worst_cor$cor, 2),
  "). Positions where athleticism translates more directly — like RB and EDGE — tend to have stronger ",
  "pre-draft signals. Positions where scheme fit and development matter more — like IOL and S — ",
  "show weaker connections between prospect scores and career outcomes."
))
```

---

## Assumption 3: Snaps Measure Quality, and All Prospects Are Fungible

The Sharpe ratio averages over all RB prospects. But we've already seen that not all classes are equal. The same logic applies to individual prospects: the model can't distinguish between an average back and a generational one.

The challenge is defining "blue chip" without hindsight. [Career Approximate Value](https://www.pro-football-reference.com/about/approximate_value.htm) (AV) -- our retrospective measure -- tells us *who turned out* to be elite, but a GM can't use it on draft day. NGS Combined Scores can. They're published before the draft, combining athletic testing, college production, and size into a single number. It's the closest thing to a prospective blue-chip identifier that exists.

**Definitions for this section:**

- **Pre-draft blue chip:** Top 10% of NGS Combined Score at the position (knowable before the draft)
- **Career blue chip:** Top 10% of [career Approximate Value](https://www.pro-football-reference.com/about/approximate_value.htm) at the position (only knowable in hindsight)
- **Hit (outcome measure):** Article 1's snap-based starter threshold — a player who plays above 75% of baseline snaps at their position. This is our outcome metric for comparing blue-chip categories because it's independent of career AV (which defines the career blue-chip group).

### NGS Scores and Career Outcomes

```{r ngs-blue-chip-scatter, fig.height=6, fig.width=8}
ngs_with_outcomes <- blue_chip %>%
  filter(!is.na(cmbn_score), !is.na(is_hit), !is.na(car_av),
         pos_group %in% key_positions, season <= 2022)

ggplot(ngs_with_outcomes, aes(x = cmbn_score, y = car_av, color = pos_group)) +
  geom_point(alpha = 0.4, size = 2) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1, alpha = 0.7) +
  scale_color_manual(values = pos_colors) +
  labs(
    title = "Do Pre-Draft NGS Scores Predict Career Value?",
    subtitle = "Each dot = a drafted player with both NGS score and career value | By position",
    x = "NGS Combined Score (pre-draft)", y = "Career Approximate Value",
    color = "Position"
  ) +
  theme_minimal(base_size = 14)
```

```{r ngs-outcome-narrative, results='asis'}
rb_ngs_cor <- ngs_with_outcomes %>%
  filter(pos_group == "RB") %>%
  summarise(r = cor(cmbn_score, car_av, use = "complete.obs"), n = n())

all_ngs_cor <- ngs_with_outcomes %>%
  summarise(r = cor(cmbn_score, car_av, use = "complete.obs"), n = n())

cat(paste0(
  "For RBs, the correlation between pre-draft NGS Combined Score and career value is **",
  round(rb_ngs_cor$r, 2), "** (n=", rb_ngs_cor$n,
  "). Across all positions it's **", round(all_ngs_cor$r, 2),
  "** (n=", all_ngs_cor$n,
  "). The relationship isn't deterministic -- plenty of high-scoring prospects bust, and ",
  "late-round gems score modestly -- but the signal is there. Higher NGS scores map to ",
  "better career outcomes at a rate that's hard to dismiss as noise."
))
```

### Pre-Draft Blue Chips vs Career Blue Chips

We can define "blue chip" two ways: retrospectively (top 10% of career AV at the position) and prospectively (top 10% of NGS Combined Score at the position). How much do they overlap -- and does it differ by position?

If high NGS scores actually predict career success, the top-scoring prospects should cluster toward the top of the career value axis. Here's every RB with an NGS score — the pre-draft blue chips (top 10% NGS) are highlighted:

```{r ngs-bc-scatter-rb, fig.height=6, fig.width=8}
rb_scatter <- ngs_with_outcomes %>%
  filter(pos_group == "RB") %>%
  mutate(is_bc = is_ngs_blue_chip)

# Label top NGS prospects
rb_labeled <- rb_scatter %>% filter(is_bc | car_av >= quantile(car_av, 0.9))

ggplot(rb_scatter, aes(x = cmbn_score, y = car_av)) +
  geom_point(data = rb_scatter %>% filter(!is_bc), color = "gray70", alpha = 0.4, size = 2) +
  geom_point(data = rb_scatter %>% filter(is_bc), color = "#377EB8", size = 3.5, alpha = 0.8) +
  geom_text(data = rb_labeled,
            aes(label = pfr_player_name, color = is_bc),
            vjust = -1, size = 2.8, fontface = "bold", show.legend = FALSE) +
  geom_smooth(method = "lm", se = TRUE, color = "#377EB8", linewidth = 0.8, alpha = 0.2) +
  scale_color_manual(values = c("TRUE" = "#377EB8", "FALSE" = "gray50")) +
  labs(
    title = "RB: Do Pre-Draft Scores Predict Career Value?",
    subtitle = "Blue = pre-draft blue chip (top 10% NGS) | Gray = everyone else | Higher NGS → higher career AV?",
    x = "NGS Combined Score (pre-draft)", y = "Career Approximate Value"
  ) +
  theme_minimal(base_size = 14)
```

```{r ngs-bc-rb-narrative, results='asis'}
rb_bc <- rb_scatter %>% filter(is_bc)
rb_nonbc <- rb_scatter %>% filter(!is_bc)

cat(paste0(
  "The pre-draft blue chips (blue dots) cluster toward the upper right — high NGS scores ",
  "mapping to high career value. Pre-draft blue chip RBs averaged **",
  round(mean(rb_bc$car_av, na.rm = TRUE), 1),
  "** career AV, vs **", round(mean(rb_nonbc$car_av, na.rm = TRUE), 1),
  "** for everyone else. They became starters at **",
  round(mean(rb_bc$is_hit, na.rm = TRUE) * 100, 1),
  "%** vs **", round(mean(rb_nonbc$is_hit, na.rm = TRUE) * 100, 1),
  "%**.\n\n",
  "Not every pre-draft blue chip panned out — Jahvid Best and LaMichael James scored elite on NGS but busted. ",
  "But the signal is clear: the top of the NGS distribution produces starters at dramatically higher rates."
))
```

The same view across all positions:

```{r ngs-bc-scatter-all, fig.height=14, fig.width=8}
all_scatter <- ngs_with_outcomes %>%
  filter(pos_group %in% key_positions) %>%
  mutate(
    is_bc = is_ngs_blue_chip,
    pos_group = factor(pos_group, levels = key_positions)
  )

# Label the top blue chips in each position
bc_labels <- all_scatter %>%
  filter(is_bc) %>%
  group_by(pos_group) %>%
  slice_max(car_av, n = 3) %>%
  ungroup()

# Label interesting gray dots: non-blue-chips with top career AV (the model misses)
gray_labels <- all_scatter %>%
  filter(!is_bc) %>%
  group_by(pos_group) %>%
  slice_max(car_av, n = 2) %>%
  ungroup()

ggplot(all_scatter, aes(x = cmbn_score, y = car_av)) +
  geom_point(data = all_scatter %>% filter(!is_bc), color = "gray70", alpha = 0.3, size = 1.5) +
  geom_point(data = all_scatter %>% filter(is_bc), color = "#377EB8", size = 2.5, alpha = 0.7) +
  geom_text(data = bc_labels, aes(label = pfr_player_name),
            vjust = -0.8, size = 2.2, color = "#377EB8", fontface = "bold") +
  geom_text(data = gray_labels, aes(label = pfr_player_name),
            vjust = -0.8, size = 2, color = "gray40", fontface = "italic") +
  geom_smooth(method = "lm", se = FALSE, color = "#377EB8", linewidth = 0.7, alpha = 0.5) +
  facet_wrap(~pos_group, scales = "free", ncol = 3) +
  labs(
    title = "Pre-Draft Score vs Career Value: All Positions",
    subtitle = "Blue = pre-draft blue chip (top 10% NGS) | Top 3 blue chips labeled per position",
    x = "NGS Combined Score", y = "Career AV"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    strip.text = element_text(face = "bold", size = 11)
  )
```

```{r ngs-bc-all-narrative, results='asis'}
bc_all <- all_scatter %>% filter(is_bc)
nonbc_all <- all_scatter %>% filter(!is_bc)

cat(paste0(
  "Across all positions, pre-draft blue chips became starters at **",
  round(mean(bc_all$is_hit, na.rm = TRUE) * 100, 1),
  "%** vs **", round(mean(nonbc_all$is_hit, na.rm = TRUE) * 100, 1),
  "%** for everyone else. The trend line slopes upward at every position — the pre-draft signal ",
  "isn't perfect, but it meaningfully separates prospects who become starters from those who don't."
))
```

### The Historical NGS Elite

Who were the highest-graded pre-draft prospects at each position, and did they deliver? Here we use the stricter bar: **top 10% of career AV** at their position — not just a starter, but a *great* player. Starting with RB:

```{r ngs-elite-rb-big, fig.height=5, fig.width=8}
# Compute career AV blue chip threshold per position
av_thresholds <- ngs_with_outcomes %>%
  group_by(pos_group) %>%
  summarise(av_90 = quantile(car_av, 0.9, na.rm = TRUE), .groups = "drop")

ngs_with_av_bc <- ngs_with_outcomes %>%
  left_join(av_thresholds, by = "pos_group") %>%
  mutate(is_career_bc = car_av >= av_90)

top_ngs_rb <- ngs_with_av_bc %>%
  filter(pos_group == "RB") %>%
  arrange(desc(cmbn_score)) %>%
  slice_head(n = 8) %>%
  mutate(
    player_label = paste0(pfr_player_name, " ('", str_sub(season, 3, 4), ")"),
    player_label = fct_reorder(player_label, cmbn_score),
    outcome = if_else(is_career_bc, "Top 10% Career AV", "Below Threshold")
  )

ggplot(top_ngs_rb, aes(x = cmbn_score, y = player_label, fill = outcome)) +
  geom_col(width = 0.7, alpha = 0.85) +
  geom_text(aes(label = round(cmbn_score, 1)), hjust = -0.15, size = 4, fontface = "bold") +
  scale_fill_manual(values = c("Top 10% Career AV" = "#1565C0", "Below Threshold" = "#EF5350")) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(
    title = "Top 8 RB Prospects by NGS Combined Score (2010-2022)",
    subtitle = "Pre-draft NGS score vs career outcome | Blue = top 10% career AV at position",
    x = "NGS Combined Score", y = NULL, fill = "Career\nOutcome"
  ) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "top")
```

```{r ngs-elite-rb-narrative, results='asis'}
rb_elite_stats <- top_ngs_rb %>%
  summarise(elite_rate = mean(is_career_bc, na.rm = TRUE), n = n())

cat(paste0(
  "The top 8 RB prospects by pre-draft NGS score finished in the top 10% of career AV at **",
  round(rb_elite_stats$elite_rate * 100, 1),
  "%** — far above the position's overall average. The names are what you'd expect: ",
  "Jonathan Taylor, Saquon Barkley, Derrick Henry, Ezekiel Elliott. These aren't random outcomes — ",
  "the pre-draft signal identified them before a single NFL snap.\n\n",
  "Now the same view across all positions:"
))
```

```{r ngs-elite-all-small, fig.height=14, fig.width=8}
top_ngs_all <- ngs_with_av_bc %>%
  filter(pos_group %in% key_positions) %>%
  group_by(pos_group) %>%
  arrange(desc(cmbn_score)) %>%
  slice_head(n = 6) %>%
  ungroup() %>%
  mutate(
    player_label = paste0(pfr_player_name, " ('", str_sub(season, 3, 4), ")"),
    player_label_unique = paste0(pos_group, "___", player_label),
    player_label_unique = fct_reorder(player_label_unique, cmbn_score),
    outcome = if_else(is_career_bc, "Top 10% Career AV", "Below Threshold"),
    pos_group = factor(pos_group, levels = key_positions)
  )

ggplot(top_ngs_all, aes(x = cmbn_score, y = player_label_unique, fill = outcome)) +
  geom_col(width = 0.7, alpha = 0.85) +
  geom_text(aes(label = round(cmbn_score, 1)), hjust = -0.15, size = 2.8, fontface = "bold") +
  scale_fill_manual(values = c("Top 10% Career AV" = "#1565C0", "Below Threshold" = "#EF5350")) +
  scale_y_discrete(labels = function(x) str_remove(x, "^.*___")) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.12))) +
  facet_wrap(~pos_group, scales = "free", ncol = 2) +
  labs(
    title = "Top 6 Historical Players by NGS Combined Score: All Positions",
    subtitle = "Pre-draft NGS score vs career outcome | Blue = top 10% career AV",
    x = "NGS Combined Score", y = NULL, fill = "Career\nOutcome"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "top",
    strip.text = element_text(face = "bold", size = 11)
  )
```

```{r ngs-elite-all-narrative, results='asis'}
elite_stats <- ngs_with_av_bc %>%
  filter(pos_group %in% key_positions) %>%
  group_by(pos_group) %>%
  arrange(desc(cmbn_score)) %>%
  slice_head(n = 6) %>%
  summarise(
    elite_rate = mean(is_career_bc, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  ) %>%
  arrange(desc(elite_rate))

best_elite <- elite_stats %>% slice_head(n = 1)
worst_elite <- elite_stats %>% slice_tail(n = 1)

cat(paste0(
  "**", best_elite$pos_group, "** leads the pack — the top 6 NGS scorers reached top-10% career AV at ",
  round(best_elite$elite_rate * 100, 1),
  "%. **", worst_elite$pos_group, "** has the weakest pre-draft-to-career translation (",
  round(worst_elite$elite_rate * 100, 1),
  "%). The model's \"never draft RB\" ",
  "is really \"never draft an *average* RB.\" When the pre-draft signal says a prospect is ",
  "separated from the class, they become *great* at dramatically higher rates."
))
```

---

## The Love Question Revisited

In Part 1, we asked: *Should you draft Jeremiah Love at 10?*

The model said no. RB has a deeply negative Sharpe in the top 10. But we've now shown that the answer depends on things the model can't see: roster needs, class depth, and prospect separation. Love is the case that brings all three together.

### The 2025 RB Class

We showed above that historical RBs with elite pre-draft NGS scores become great at dramatically higher rates than the position average. Where does Love's 2025 class fit in?

```{r ngs-love-bar-big, fig.height=7, fig.width=8}
ngs_rb <- ngs_prospects %>%
  filter(pos_group == "RB", !is.na(cmbn_score)) %>%
  arrange(desc(cmbn_score)) %>%
  mutate(
    is_love = str_detect(player, "(?i)love"),
    ngs_tier = case_when(
      cmbn_score >= 90 ~ "Elite",
      cmbn_score >= 70 ~ "Good",
      cmbn_score >= 60 ~ "Average",
      TRUE ~ "Below Avg"
    ),
    ngs_tier = factor(ngs_tier, levels = c("Elite", "Good", "Average", "Below Avg")),
    player_label = paste0(player, " (", college, ")"),
    player_label = fct_reorder(player_label, cmbn_score)
  )

ggplot(ngs_rb, aes(x = cmbn_score, y = player_label, fill = ngs_tier)) +
  geom_col(width = 0.7, alpha = 0.85) +
  geom_text(aes(label = round(cmbn_score, 1)), hjust = -0.15, size = 3, fontface = "bold") +
  scale_fill_manual(
    values = c("Elite" = "#1565C0", "Good" = "#66BB6A", "Average" = "#FFA726", "Below Avg" = "#EF5350"),
    name = "NGS Tier"
  ) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.12))) +
  coord_cartesian(xlim = c(50, NA)) +
  labs(
    title = "2025 RB Prospects: NGS Combined Score",
    subtitle = "Love's 94.3 is tied for #1 overall (all positions) with Sonny Styles | ProdScr: 95.9 (highest of any prospect)",
    x = "NGS Combined Score", y = NULL
  ) +
  theme_minimal(base_size = 14)
```

```{r ngs-love-narrative, results='asis'}
love_row <- ngs_rb %>% filter(is_love) %>% slice(1)
rb2_row <- ngs_rb %>% filter(!is_love) %>% slice_max(cmbn_score, n = 1)
gap <- love_row$cmbn_score - rb2_row$cmbn_score

cat(paste0(
  "Love's NGS Combined Score of **", round(love_row$cmbn_score, 1),
  "** is tied for #1 overall across all positions. His Production Score of **95.9** is the highest ",
  "of any prospect in the 2025 class. The gap to RB2 (", rb2_row$player, " at ",
  round(rb2_row$cmbn_score, 1), ") is **", round(gap, 1), " points** -- a chasm by NGS standards."
))
```

### Where Love Ranks in the Historical Record

How does Love's 94.3 compare to every RB prospect since 2010? Here's every RB with an NGS Combined Score, 2010-2025 — Love highlighted:

```{r love-historical-rank, fig.height=5, fig.width=8}
# Combine historical and 2025 RB NGS scores
historical_rb_ngs <- ngs_with_outcomes %>%
  filter(pos_group == "RB") %>%
  select(player_name = pfr_player_name, season, cmbn_score) %>%
  mutate(source = "Historical (2010-2022)")

prospect_rb_ngs <- ngs_prospects %>%
  filter(pos_group == "RB", !is.na(cmbn_score)) %>%
  select(player_name = player, cmbn_score) %>%
  mutate(season = 2025, source = "2025 Prospect")

all_rb_ngs <- bind_rows(historical_rb_ngs, prospect_rb_ngs) %>%
  arrange(desc(cmbn_score)) %>%
  mutate(
    rank = row_number(),
    is_love = str_detect(player_name, "(?i)love") & source == "2025 Prospect"
  )

love_rank <- all_rb_ngs %>% filter(is_love) %>% pull(rank)
total_rb <- nrow(all_rb_ngs)
love_pctile <- round((1 - love_rank / total_rb) * 100, 1)

# Show top 20 as a horizontal bar
top20 <- all_rb_ngs %>% slice_head(n = 20) %>%
  mutate(
    label = paste0(player_name, " ('", str_sub(season, 3, 4), ")"),
    label = fct_reorder(label, cmbn_score)
  )

ggplot(top20, aes(x = cmbn_score, y = label, fill = is_love)) +
  geom_col(width = 0.7, alpha = 0.85) +
  geom_text(aes(label = round(cmbn_score, 1)), hjust = -0.15, size = 3.5, fontface = "bold") +
  scale_fill_manual(values = c("TRUE" = "#1565C0", "FALSE" = "gray60"), guide = "none") +
  scale_x_continuous(expand = expansion(mult = c(0, 0.12))) +
  coord_cartesian(xlim = c(75, NA)) +
  labs(
    title = "Top 20 RB Prospects by NGS Score (2010-2025)",
    subtitle = paste0("Love ranks #", love_rank, " out of ", total_rb, " RBs — ", love_pctile, "th percentile"),
    x = "NGS Combined Score", y = NULL
  ) +
  theme_minimal(base_size = 13)
```

```{r love-historical-narrative, results='asis'}
cat(paste0(
  "Love's 94.3 ranks **#", love_rank, "** among all ", total_rb,
  " RB prospects in the NGS era. That puts him in the **", love_pctile,
  "th percentile** — in the company of Jonathan Taylor, Saquon Barkley, and Breece Hall. ",
  "The pre-draft signal is as strong as it gets.\n\n",
  "But Love isn't the only outlier in 2025. Here's a look at the top prospect at every position:"
))
```

```{r ngs-2025-all-positions, fig.height=14, fig.width=8}
# Top 5 prospects by position in 2025 class
top_2025_all <- ngs_prospects %>%
  filter(!is.na(cmbn_score), pos_group %in% key_positions) %>%
  group_by(pos_group) %>%
  arrange(desc(cmbn_score)) %>%
  slice_head(n = 5) %>%
  ungroup() %>%
  mutate(
    ngs_tier = case_when(
      cmbn_score >= 90 ~ "Elite",
      cmbn_score >= 70 ~ "Good",
      cmbn_score >= 60 ~ "Average",
      TRUE ~ "Below Avg"
    ),
    ngs_tier = factor(ngs_tier, levels = c("Elite", "Good", "Average", "Below Avg")),
    player_label = paste0(player, " (", college, ")"),
    player_label_unique = paste0(pos_group, "___", player_label),
    player_label_unique = fct_reorder(player_label_unique, cmbn_score),
    pos_group = factor(pos_group, levels = key_positions)
  )

ggplot(top_2025_all, aes(x = cmbn_score, y = player_label_unique, color = ngs_tier)) +
  geom_point(size = 4) +
  geom_text(aes(label = round(cmbn_score, 1)), hjust = -0.4, size = 2.8, fontface = "bold",
            show.legend = FALSE) +
  scale_color_manual(
    values = c("Elite" = "#1565C0", "Good" = "#66BB6A", "Average" = "#FFA726", "Below Avg" = "#EF5350"),
    name = "NGS Tier"
  ) +
  scale_y_discrete(labels = function(x) str_remove(x, "^.*___")) +
  scale_x_continuous(limits = c(58, 100), breaks = seq(60, 100, 10)) +
  facet_wrap(~pos_group, scales = "free_y", ncol = 3) +
  labs(
    title = "2025 Draft Class: Top 5 Prospects by NGS Combined Score",
    subtitle = "Elite (90+) | Good (70-90) | Average (60-70) | Below Avg (<60)",
    x = "NGS Combined Score", y = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "top",
    strip.text = element_text(face = "bold", size = 11),
    panel.grid.major.y = element_blank()
  )
```

```{r ngs-2025-all-narrative, results='asis'}
# Find positions with biggest separation
pos_sep <- ngs_prospects %>%
  filter(!is.na(cmbn_score), pos_group %in% key_positions) %>%
  group_by(pos_group) %>%
  arrange(desc(cmbn_score)) %>%
  summarise(
    top_player = first(player),
    top_score = first(cmbn_score),
    second_score = nth(cmbn_score, 2),
    gap = first(cmbn_score) - nth(cmbn_score, 2),
    .groups = "drop"
  ) %>%
  arrange(desc(gap))

biggest_gap <- pos_sep %>% slice_head(n = 1)
smallest_gap <- pos_sep %>% slice_tail(n = 1)

cat(paste0(
  "The 2025 class has fascinating separation patterns. **", biggest_gap$pos_group,
  "** has the biggest gap between the top prospect (",
  biggest_gap$top_player, " at ", round(biggest_gap$top_score, 1),
  ") and #2 — a **", round(biggest_gap$gap, 1),
  "-point** chasm. Meanwhile, **", smallest_gap$pos_group,
  "** is the tightest race, with just a **", round(smallest_gap$gap, 1),
  "-point** gap at the top. Where a prospect is clearly separated, the blue-chip argument ",
  "gets stronger. Where the class is tightly bunched, the model's average-case advice holds."
))
```

### How Good Does Love Need to Be?

This is the practical question. Article 1 gave us the Sharpe ratio for RB in the top 10: **-0.615**. That's the average across all RBs drafted in the top 10 since 2010. But averages hide a massive range. Here's every first-round RB in our data, plotted against the opportunity cost of not drafting the best available non-RB at the same pick.

Each player's individual Sharpe is computed the same way as Article 1's positional Sharpe, but for a single player: their rookie contract surplus, second-contract value, and whether they became a starter (hit = 75% of positional snap baseline). The **circle** is the RB's individual Sharpe. The **yellow X** is the best non-RB taken in the next 5 picks — a realistic proxy for who the team could have drafted instead. Green means the RB bet paid off; red means the nearby alternative would have been better.

```{r rb-breakeven, fig.height=9, fig.width=8}
# All first-round picks with individual Sharpe
all_r1 <- player_lookup %>%
  filter(pick <= 32, pos_group %in% key_positions)

# For each RB first rounder, find the BEST non-RB still available (highest Sharpe in rest of round 1)
rb_r1 <- all_r1 %>% filter(pos_group == "RB")

rb_with_alts <- map_dfr(1:nrow(rb_r1), function(i) {
  row <- rb_r1[i, ]
  candidate <- all_r1 %>%
    filter(season == row$season, pos_group != "RB", pick > row$pick, pick <= row$pick + 5) %>%
    slice_max(player_sharpe, n = 1)
  if (nrow(candidate) > 0) {
    row %>% mutate(
      alt_name = candidate$pfr_player_name,
      alt_pos = candidate$pos_group,
      alt_pick = candidate$pick,
      alt_sharpe = candidate$player_sharpe
    )
  } else {
    row %>% mutate(
      alt_name = NA_character_, alt_pos = NA_character_,
      alt_pick = NA_real_, alt_sharpe = NA_real_
    )
  }
})

rb_with_alts <- rb_with_alts %>%
  filter(!is.na(alt_sharpe)) %>%
  mutate(
    beat_alt = player_sharpe > alt_sharpe,
    rb_label = paste0(pfr_player_name, " (#", pick, ", '", str_sub(season, 3, 4), ")"),
    alt_label = paste0(alt_name, " (", alt_pos, ")"),
    row_label = paste0(rb_label, "  vs  ", alt_label),
    row_label = fct_reorder(row_label, pick)
  )

ggplot(rb_with_alts, aes(y = row_label)) +
  geom_vline(xintercept = 0, color = "gray80", linewidth = 0.5) +
  # Connecting segment
  geom_segment(aes(x = player_sharpe, xend = alt_sharpe,
                   yend = row_label, color = beat_alt),
               linewidth = 1.5, alpha = 0.4) +
  # RB dot
  geom_point(aes(x = player_sharpe, color = beat_alt), size = 4) +
  # Alternative dot
  geom_point(aes(x = alt_sharpe), shape = 4, size = 3, color = "#FF9800", stroke = 1.5) +
  scale_color_manual(values = c("TRUE" = "#2e7d32", "FALSE" = "#c62828"),
                     labels = c("TRUE" = "RB beat the alternative", "FALSE" = "Alternative was better"),
                     name = NULL) +
  labs(
    title = "The RB Gamble: Who They Took\nvs Who They Could Have Had",
    subtitle = "Circle = RB | X = best non-RB in next 5 picks | Green = RB won the bet",
    x = "Individual Player Sharpe", y = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "top",
    axis.text.y = element_text(size = 9)
  )
```

```{r rb-breakeven-narrative, results='asis'}
rb_top10_sharpe <- sharpe %>% filter(pos_group == "RB", tier == "Top 10") %>% pull(sharpe_elite)
cmc <- rb_with_alts %>% filter(pfr_player_name == "Christian McCaffrey")
gurley <- rb_with_alts %>% filter(pfr_player_name == "Todd Gurley")
trich <- rb_with_alts %>% filter(pfr_player_name == "Trent Richardson")
rb_beat_pct <- mean(rb_with_alts$beat_alt, na.rm = TRUE)

cat(paste0(
  "Article 1's tier-level Sharpe for RB in the top 10 is **", round(rb_top10_sharpe, 3),
  "** — the worst of any position. Now you can see why: most RBs lost the bet against the best nearby alternative.\n\n",
  "**The wins:** CMC (Sharpe ", round(cmc$player_sharpe, 2),
  ") beat ", cmc$alt_name, " (", cmc$alt_pos, ", ",
  round(cmc$alt_sharpe, 2), "). Gurley (",
  round(gurley$player_sharpe, 2), ") beat ", gurley$alt_name, " (", gurley$alt_pos, ").\n\n",
  "**The losses:** Trent Richardson at pick 3 (Sharpe ",
  round(trich$player_sharpe, 2), ") — the team could have had **", trich$alt_name,
  "** (", trich$alt_pos, ", Sharpe ", round(trich$alt_sharpe, 2), ").\n\n",
  "Only **", round(rb_beat_pct * 100), "%** of first-round RBs beat the best non-RB taken in the next 5 picks. ",
  "Love needs to be in the CMC/Gurley cluster, not the Richardson cluster, to justify the pick."
))
```

### Sharpe at Pick 3: The Steepest Cost

```{r sharpe-at-3, results='asis'}
# What do other positions return at pick 3 specifically?
pick_3_5 <- player_lookup %>%
  filter(pick >= 2, pick <= 5) %>%
  group_by(pos_group) %>%
  summarise(
    mean_sharpe = mean(player_sharpe, na.rm = TRUE),
    mean_return = mean(player_return, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  ) %>%
  arrange(desc(mean_sharpe))

best_at_3 <- pick_3_5 %>% slice_head(n = 3)
rb_at_3 <- pick_3_5 %>% filter(pos_group == "RB")

cat(paste0(
  "At picks 2-5 specifically, the opportunity cost is highest. The best-returning positions: **",
  best_at_3$pos_group[1], "** (avg Sharpe ", round(best_at_3$mean_sharpe[1], 2),
  "), **", best_at_3$pos_group[2], "** (", round(best_at_3$mean_sharpe[2], 2),
  "), **", best_at_3$pos_group[3], "** (", round(best_at_3$mean_sharpe[3], 2),
  "). RB at picks 2-5 averages just **", round(rb_at_3$mean_sharpe, 2),
  "** — that's Saquon Barkley (#2), Trent Richardson (#3), Zeke (#4), and Fournette (#4).\n\n",
  "At pick 3, you're giving up the most draft capital. The only RB ever drafted at 3 in our window ",
  "was Trent Richardson, who returned almost nothing. Taking Love at 3 would require him to be ",
  "a **CMC-level outcome** just to match what an average EDGE or QB pick would have returned there.\n\n",
  "This is where Article 1's advice is strongest. The higher the pick, the more the RB needs to ",
  "overcome. The case for Love gets better as you slide down the board."
))
```

### Where in the First Round Does Love Make Sense?

Here's the real question: at what pick does an RB with Love's profile start to justify the cost? We can compute this by comparing the CMC/Gurley outcome (the realistic ceiling for an elite RB) against the average non-RB return at each pick range.

```{r love-pick-range, fig.height=5, fig.width=8}
# Compare CMC/Gurley-level return vs opportunity cost across the first round
pick_ranges <- player_lookup %>%
  filter(pick <= 32) %>%
  mutate(pick_range = case_when(
    pick <= 5 ~ "1-5",
    pick <= 10 ~ "6-10",
    pick <= 16 ~ "11-16",
    pick <= 24 ~ "17-24",
    pick <= 32 ~ "25-32"
  )) %>%
  mutate(pick_range = factor(pick_range, levels = c("1-5", "6-10", "11-16", "17-24", "25-32")))

# Average non-RB return by pick range
opp_by_range <- pick_ranges %>%
  filter(pos_group != "RB") %>%
  group_by(pick_range) %>%
  summarise(mean_sharpe = mean(player_sharpe, na.rm = TRUE), .groups = "drop") %>%
  mutate(type = "Avg Non-RB Pick")

# RB hits (CMC/Gurley tier — Sharpe > 2.0) by pick range
rb_ceiling <- pick_ranges %>%
  filter(pos_group == "RB", player_sharpe > 2.0) %>%
  group_by(pick_range) %>%
  summarise(mean_sharpe = mean(player_sharpe, na.rm = TRUE), .groups = "drop") %>%
  mutate(type = "Elite RB (CMC/Gurley tier)")

# All RB average
rb_avg <- pick_ranges %>%
  filter(pos_group == "RB") %>%
  group_by(pick_range) %>%
  summarise(mean_sharpe = mean(player_sharpe, na.rm = TRUE), .groups = "drop") %>%
  mutate(type = "Avg RB Pick")

compare <- bind_rows(opp_by_range, rb_ceiling, rb_avg) %>%
  mutate(type = factor(type, levels = c("Elite RB (CMC/Gurley tier)", "Avg Non-RB Pick", "Avg RB Pick")))

ggplot(compare, aes(x = pick_range, y = mean_sharpe, fill = type)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6, alpha = 0.85) +
  geom_hline(yintercept = 0, color = "gray60", linewidth = 0.5) +
  scale_fill_manual(values = c(
    "Elite RB (CMC/Gurley tier)" = "#377EB8",
    "Avg Non-RB Pick" = "#FF9800",
    "Avg RB Pick" = "#c62828"
  )) +
  labs(
    title = "Where Does an Elite RB Justify the Pick?",
    subtitle = "CMC/Gurley-tier RB Sharpe vs average non-RB alternative | By first-round pick range",
    x = "Pick Range", y = "Average Individual Sharpe", fill = NULL
  ) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "top")
```

```{r love-pick-narrative, results='asis'}
cat(paste0(
  "The blue bars are the CMC/Gurley tier — first-round RBs who hit at elite levels. ",
  "The orange bars are what you'd expect from the average non-RB pick in the same range. ",
  "The red bars are the average RB — the number Article 1 is built on.\n\n",
  "At picks 1-5, even an elite RB barely matches the average non-RB alternative. The opportunity cost is simply too high. ",
  "At picks 6-10, the elite RB **clears** the bar — this is where CMC (pick 8) and Gurley (pick 10) ",
  "actually lived, and both delivered positive returns above what the average alternative position offered. ",
  "In the late first, the math gets more favorable, but elite RB talent is rarely available that late."
))
```

### Love's Fair Value: What the Data Says

The chart above tells us what happens when an elite RB *hits*. But you don't know if Love will be CMC or Fournette before the draft. The real question: given the full range of outcomes for blue-chip RB prospects, at what pick does the expected value justify the selection?

```{r love-fair-value, fig.height=6, fig.width=8}
# Blue-chip RB expected Sharpe (all NGS top-10% RBs)
bc_rb_sharpes <- blue_chip %>%
  filter(pos_group == "RB", is_ngs_blue_chip == TRUE, !is.na(player_sharpe))

bc_mean <- mean(bc_rb_sharpes$player_sharpe)
bc_median <- median(bc_rb_sharpes$player_sharpe)

# Non-RB average Sharpe by individual pick, smoothed
non_rb_by_pick <- player_lookup %>%
  filter(pick <= 32, pos_group != "RB") %>%
  group_by(pick) %>%
  summarise(avg_sharpe = mean(player_sharpe, na.rm = TRUE), n = n(), .groups = "drop") %>%
  arrange(pick)

# 5-pick rolling mean (manual, no zoo dependency)
smooth_vals <- numeric(32)
for(i in 1:32) {
  window <- max(1, i-2):min(32, i+2)
  smooth_vals[i] <- mean(non_rb_by_pick$avg_sharpe[window])
}
non_rb_by_pick$smooth_sharpe <- smooth_vals

ggplot(non_rb_by_pick, aes(x = pick)) +
  geom_line(aes(y = smooth_sharpe), color = "#FF9800", linewidth = 1.2) +
  geom_point(aes(y = avg_sharpe), color = "#FF9800", alpha = 0.4, size = 2) +
  geom_hline(yintercept = bc_mean, color = "#377EB8", linewidth = 1.2, linetype = "solid") +
  geom_hline(yintercept = bc_median, color = "#377EB8", linewidth = 1, linetype = "dashed") +
  annotate("text", x = 2, y = bc_mean + 0.06, label = paste0("BC RB expected value (mean: ", round(bc_mean, 2), ")"),
           color = "#377EB8", hjust = 0, size = 3.5, fontface = "bold") +
  annotate("text", x = 2, y = bc_median + 0.06, label = paste0("BC RB median: ", round(bc_median, 2)),
           color = "#377EB8", hjust = 0, size = 3.5, fontface = "italic") +
  annotate("text", x = 28, y = non_rb_by_pick$smooth_sharpe[28] + 0.08,
           label = "Avg non-RB at pick", color = "#FF9800", hjust = 1, size = 3.5, fontface = "bold") +
  scale_x_continuous(breaks = seq(1, 32, 2)) +
  labs(
    title = "Love's Fair Value: Expected RB Return vs Opportunity Cost",
    subtitle = "Blue line = expected Sharpe for a blue-chip RB prospect | Orange = avg non-RB alternative",
    x = "Pick Number", y = "Individual Player Sharpe"
  ) +
  theme_minimal(base_size = 13)
```

```{r love-fair-value-narrative, results='asis'}
# Find first pick where mean BC beats smoothed non-RB
crossover <- non_rb_by_pick %>% filter(smooth_sharpe <= bc_mean) %>% slice_head(n = 1)
crossover_med <- non_rb_by_pick %>% filter(smooth_sharpe <= bc_median) %>% slice_head(n = 1)

# Also compute: what % of BC RBs returned above the non-RB average at different pick ranges
early <- non_rb_by_pick %>% filter(pick <= 10) %>% summarise(avg = mean(smooth_sharpe))
mid <- non_rb_by_pick %>% filter(pick >= 11, pick <= 20) %>% summarise(avg = mean(smooth_sharpe))
late <- non_rb_by_pick %>% filter(pick >= 21) %>% summarise(avg = mean(smooth_sharpe))

pct_beat_early <- mean(bc_rb_sharpes$player_sharpe > early$avg)
pct_beat_mid <- mean(bc_rb_sharpes$player_sharpe > mid$avg)
pct_beat_late <- mean(bc_rb_sharpes$player_sharpe > late$avg)

cat(paste0(
  "The blue line is the expected Sharpe for a blue-chip RB prospect — the average outcome across all ",
  "historical RBs who scored in the top 10% of NGS Combined Score. The orange line is the average ",
  "non-RB return at each pick.\n\n",
  "**The honest answer:** the expected value of a blue-chip RB (",
  round(bc_mean, 2), ") never clearly exceeds the average non-RB alternative in the first round. ",
  "The opportunity cost is real at every pick.\n\n",
  "But expected value isn't the whole story. Among historical blue-chip RBs:\n\n",
  "- **", round(pct_beat_early * 100), "%** would have beaten the average non-RB pick in the top 10\n",
  "- **", round(pct_beat_mid * 100), "%** would have beaten the average non-RB pick at 11-20\n",
  "- **", round(pct_beat_late * 100), "%** would have beaten the average non-RB pick at 21-32\n\n",
  "That's the gamble. You're not betting on the average outcome — you're betting that Love is in the ",
  round(pct_beat_mid * 100), "% who beat the alternative, not the ",
  round((1 - pct_beat_mid) * 100), "% who don't. ",
  "The data can't tell you which side Love falls on. But it can tell you the odds — and where those odds start to favor the bet."
))
```

### Which Teams Actually Need an RB?

Our quality gate for most positions requires a premium second contract. But that's circular for RB — the Sharpe model already told us RBs don't get paid, so of course they never clear the gate. We need a different definition of "need" for running back.

For RB specifically, we define "filled" using **snap-based production only**: does the team have a back who played above the starter snap threshold (75% of positional baseline) in the prior season? No contract requirement. If the team had a productive back on the field, the position is filled regardless of what the market pays him.

```{r rb-need-snap-based, results='asis'}
# Redefine RB need as snap-based only (no contract gate)
# starter_is_hit can be NA (no starter found), treat NA as not filled
rb_snap_needs <- roster_needs %>%
  filter(pos_group == "RB") %>%
  mutate(
    snap_filled = coalesce(starter_is_hit, FALSE)
  )

rb_snap_by_team <- rb_snap_needs %>%
  group_by(team) %>%
  summarise(
    need_rate = mean(!snap_filled),
    n_seasons = n(),
    .groups = "drop"
  )

avg_need <- mean(rb_snap_by_team$need_rate)
teams_with_need <- rb_snap_by_team %>% filter(need_rate >= 0.5) %>% nrow()
total_teams <- nrow(rb_snap_by_team)

# Total team-seasons
total_ts <- nrow(rb_snap_needs)
need_ts <- sum(!rb_snap_needs$snap_filled)

cat(paste0(
  "Based on this definition, what teams \"need\" an RB? Using snap-based production (no contract gate), ",
  "**", round(need_ts / total_ts * 100), "%** of team-seasons show RB as a need — meaning the team didn't have a ",
  "starter-level back the prior year. **", teams_with_need, " of ", total_teams,
  "** teams needed an RB at least half the time.\n\n",
  "That's still a lot, which tells us something real: RB is a high-turnover position. ",
  "Backs get hurt, decline fast, and cycle through rosters more than QBs or OTs. ",
  "The need is genuine — it's not an artifact of our quality gate.\n\n",
  "The question isn't *whether* to invest in RB — most teams need one in most years. ",
  "It's whether THIS investment, in THIS prospect, in THIS class, justifies accepting the negative positional Sharpe."
))
```

### Who Should Draft Love? The Roster Piece

The breakeven analysis tells us *where* in the first round Love makes sense (picks 6-10). The class depth tells us *why* this year is different (thin RB class, huge separation). But we haven't answered *which teams* should actually consider it. That requires the roster analysis from Argument 1.

```{r love-roster-fit, results='asis'}
# Which teams picking in the top 10 of 2025 need an RB (snap-based)?
# Use the most recent season in our data as proxy for 2025 needs
latest_season <- max(roster_needs$draft_season)
recent_rb_needs <- roster_needs %>%
  filter(pos_group == "RB", draft_season == latest_season) %>%
  mutate(snap_filled = coalesce(starter_is_hit, FALSE))

teams_needing_rb <- recent_rb_needs %>%
  filter(!snap_filled) %>%
  pull(team)

n_needing <- length(teams_needing_rb)
n_total <- nrow(recent_rb_needs)

cat(paste0(
  "In the most recent season of our data (", latest_season, "), **", n_needing, " of ", n_total,
  "** teams lacked a starter-level RB by snap production",
  if(n_needing > 0 & n_needing <= 15) paste0(": ", paste(teams_needing_rb, collapse = ", ")) else "", ".\n\n",
  "The case for Love is strongest when all three factors align: a team that genuinely needs ",
  "an RB, picking in the 6-10 range, where Love's elite profile and the thin class ",
  "justify accepting the negative positional Sharpe. Teams picking in the top 5 face too steep ",
  "an opportunity cost — even CMC-level outcomes barely break even there. Teams with a productive ",
  "back already on the roster should follow the model's advice and draft a premium position."
))
```

### Putting It Together

Everything we've built in this article converges on a single question. Three arguments, one prospect:

<div style="background-color: #f0f4f8; border-left: 4px solid #377EB8; padding: 15px 20px; margin: 20px 0; border-radius: 4px;">

**The Sharpe ratio says:** RB is the worst first-round investment. Sharpe of -0.615 in the top 10. Negative across every tier. Don't do it.

**The class signal says:** 2025 is a thin RB class. If you pass on Love, the drop-off is steep -- a 17-point gap to RB2.

**The NGS history says:** RBs with scores like Love's (94.3) have historically hit at rates far above the position average. The pre-draft signal maps to career outcomes. CMC and Gurley — the RB ceiling — had individual Sharpes above 2.0.

**The breakeven says:** At picks 1-5, even an elite RB barely clears the opportunity cost. At picks 6-10, the math works — if the prospect is truly separated. At 15+, an elite profile justifies the pick outright.

**The need says:** Most teams need an RB in most years. It's a high-turnover position. The question isn't whether to invest — it's where and in whom.

</div>

Does this mean a team *should* draft Love? Not at 3. Probably not at 5. But at 8-10, for a team that genuinely needs an RB? The Sharpe model says no. The prospect data says he's in the CMC/Gurley cluster. The class is thin. And history says the few RBs who clear the bar at that range more than justify the pick.

That's the honest answer. Not a prescription, but an acknowledgment that the model's average-case advice meets its limit when roster need, prospect separation, and class depth all point the other direction — and the pick falls in the range where elite RBs have historically delivered.

---

## What We Learned

We set out to stress-test the Sharpe ratio, not to destroy it. Three arguments, three perspectives. Here's what we found.

The Sharpe ratio is a powerful tool for general draft strategy. But it operates at the level of averages, and GMs operate at the level of specifics. The model can't account for three things: **roster context** (do you actually need this position?), **class variance** (is this a good year to draft it?), and **prospect separation** (is this prospect different from the rest?).

When those factors align -- a genuine need, a weak class at that position, and a prospect whose pre-draft profile is separated from the field -- the model's blanket advice meets a real edge case. Not a refutation, but a boundary condition.

The NGS data gave us something the retrospective analysis couldn't: a prospective signal. Historical RBs with elite NGS Combined Scores hit at rates dramatically above the position average. The correlation between pre-draft scores and career outcomes isn't perfect, but it's real enough to challenge the assumption that all prospects at a position are interchangeable.

None of this proves that any specific pick will work. The Sharpe ratio is right far more often than it's wrong. But when teams draft a low-Sharpe position and it works out, it's often because they had information the model didn't -- roster need, class quality, prospect separation. That's not irrationality. That's context.

---

## What Survived the Stress Test

The counter-arguments don't destroy the model -- they define its edges. Here's what held up:

- **QB is still the best top-10 investment on average.** Nothing we found changes the overall positional hierarchy. If you need a QB and there's a blue-chip one available, take him without hesitation.

- **Late 1st round is still the sweet spot for OT and EDGE.** The combination of high hit rates and strong second-contract value makes picks 15-32 the most efficient draft real estate for these positions.

- **Day 3 RBs are still fine most of the time.** The cheap FA replacement cost means you can fill the position without premium capital. When the class isn't deep and the prospect isn't blue-chip, this is still the right call.

- **The aggregate advice is right; context determines the exceptions.** The Sharpe ratio is a compass, not a GPS. It points you in the right direction. But the last mile requires scouting, roster analysis, and class evaluation that no single metric can capture.

- **RBs produce immediately on rookie contracts.** This actually reinforces the Sharpe framework in a subtle way: RBs deliver their best production during their cheapest years. The rookie contract surplus is real — but it also means teams get the most value from an RB *before* the second contract the model measures. The model captures this: it's not that RBs are bad, it's that the position's value is front-loaded into the years where every pick is cheap.

---

## Methodology Notes

**Data:** Player stats from Pro Football Reference via `nflreadr`. Contract data from OverTheCap via `nflreadr`. All values normalized to percentage of salary cap at time of signing. Analysis covers 2010-2022 draft classes (minimum 4 NFL seasons for fair evaluation).

**Hit definition (Article 1):** Snap-based quality measure. A "hit" is a player who plays above the positional snap threshold (75% of baseline snaps for their position). This measures whether a player became a starter.

**Blue-chip outcome measure:** For comparing blue-chip categories, we use Article 1's snap-based hit rate as the outcome (independent of career AV, which defines the career blue-chip group). [Career Approximate Value](https://www.pro-football-reference.com/about/approximate_value.htm) (AV) from Pro Football Reference is used for defining career blue chips (top 10% at position) but not as the outcome measure, to avoid circularity.

**Blue-chip classification:** Two definitions used. Retrospective: top 10% of career AV within each position group. Prospective: top 10% of NGS Combined Score within each position group. Players meeting both criteria had the highest elite outcome rates.

**NGS Combined Score:** NFL Next Gen Stats composite prospect score combining athletic testing, college production, and size metrics. Published pre-draft. Historical scores matched to draft records via normalized name and position group.

**Class depth:** A class is "deep" if its count of Good+ prospects (NGS Combined Score >= 70) exceeds the position's median Good+ count across all years. "Thin" if below the median.

**Roster needs:** A position is "filled" if the team had a starter who played above the hit threshold, earned a second contract above the free-agent replacement cost for the position, and was not on an expiring deal. A position is a "need" if it wasn't filled.

**Part 1:** Full methodology and interactive charts at [stranger9977.github.io/draft-sharpe-analysis](https://stranger9977.github.io/draft-sharpe-analysis/)

*Built with R, ggplot2, nflplotR, and gt. Data from nflverse and Pro Football Reference.*
