---
title: "What the Sharpe Ratio Can't See"
subtitle: "Stress-testing our NFL Draft model — and finding where it ends"
output:
  html_document:
    toc: true
    toc_float: true
    self_contained: true
knit: (function(input, ...) rmarkdown::render(input, output_dir = dirname(input), ...))
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, warning = FALSE, message = FALSE,
  fig.width = 10, fig.height = 7, dpi = 300,
  fig.path = "charts/v2_"
)
library(tidyverse)
library(arrow)
library(scales)
library(gt)
library(nflplotR)
library(nflreadr)

pos_colors <- c(
  "QB" = "#E41A1C", "RB" = "#377EB8", "WR" = "#4DAF4A", "TE" = "#984EA3",
  "OT" = "#FF7F00", "IOL" = "#FFFF33", "EDGE" = "#A65628", "DL" = "#F781BF",
  "LB" = "#999999", "CB" = "#66C2A5", "S" = "#FC8D62"
)

tier_order <- c("Top 10", "Late 1st", "Round 2", "Round 3", "Rounds 4-7")
key_positions <- c("QB", "RB", "WR", "TE", "OT", "IOL", "EDGE", "DL", "LB", "CB", "S")
```

```{r load-data}
# Article 1 data
df <- read_parquet("data/draft_sharpe_analysis.parquet")
sharpe <- read_csv("output/sharpe_ratios.csv", show_col_types = FALSE)
player_lookup <- read_csv("output/player_lookup.csv", show_col_types = FALSE)

sharpe <- sharpe %>% mutate(tier = factor(tier, levels = tier_order))
df <- df %>% mutate(tier = factor(tier, levels = tier_order))
player_lookup <- player_lookup %>% mutate(tier = factor(tier, levels = tier_order))
df_eligible <- df %>% filter(season <= 2022)

# V2 data
class_depth <- read_csv("output/v2/class_depth.csv", show_col_types = FALSE)
blue_chip <- read_csv("output/v2/blue_chip.csv", show_col_types = FALSE)
roster_needs <- read_csv("output/v2/roster_needs.csv", show_col_types = FALSE)
team_efficiency <- read_csv("output/v2/team_efficiency_adjusted.csv", show_col_types = FALSE)
r1_picks <- read_csv("output/v2/r1_picks_need_adjusted.csv", show_col_types = FALSE)
market_efficiency <- read_csv("output/v2/market_efficiency.csv", show_col_types = FALSE)
ngs_prospects <- read_csv("output/v2/ngs_prospects_2025.csv", show_col_types = FALSE)
ngs_historical <- read_csv("output/v2/ngs_historical_matched.csv", show_col_types = FALSE)
ngs_class_tiers <- read_csv("output/v2/ngs_class_tiers.csv", show_col_types = FALSE)

# Factor levels
blue_chip <- blue_chip %>% mutate(tier = factor(tier, levels = tier_order))
r1_picks <- r1_picks %>% mutate(tier = factor(tier, levels = tier_order))
```

## The Model Was Right. And Incomplete.

In [Part 1](https://stranger9977.github.io/draft-sharpe-analysis/), we built a quasi-Sharpe ratio for NFL draft picks. The math was clean. The conclusion was clear: **RB shows negative returns across all tiers -- never draft one in round 1.**

The model is right on average. But GMs don't draft on average.

They draft in specific years, with specific rosters, facing specific prospects. A model that can't account for context isn't wrong -- it's incomplete. This is Part 2: an honest stress test that defines where Article 1's advice ends and judgment begins. Not to destroy the model, but to find its edges.

The gap between the model's advice and what teams actually do isn't always inefficiency -- sometimes it's intelligence. We tested four assumptions behind Part 1:

```{r assumptions-table, results='asis'}
cat("
| Assumption We Made | What Could Break It |
|---|---|
| Teams should always draft the highest-Sharpe position | Roster needs make some positions irrelevant |
| All draft classes are interchangeable | Class depth varies wildly year to year |
| All prospects at a position are fungible | Blue-chip prospects break the averages |
")
```

All three will crack the model open.

---

## Quick Recap: What Part 1 Found

```{r recap-sharpe, results='asis'}
rb_top10 <- sharpe %>% filter(pos_group == "RB", tier == "Top 10")
qb_top10 <- sharpe %>% filter(pos_group == "QB", tier == "Top 10")
best_late1 <- sharpe %>% filter(tier == "Late 1st", !is.na(sharpe_elite)) %>%
  slice_max(sharpe_elite, n = 1)

cat(paste0(
  "The quasi-Sharpe ratio combined three things: rookie contract surplus (how much cheaper ",
  "a drafted starter is than a free agent), second-contract value (the market's verdict), ",
  "and elite probability (the chance of a top-decile outcome). We measured risk-adjusted ",
  "returns by position and draft tier from 2010--2022.\n\n",
  "**Key findings:** QB leads all positions in top-10 Sharpe (",
  round(qb_top10$sharpe_elite, 2), "). OT and EDGE peak in the late first round (",
  round(best_late1$sharpe_elite, 2), " for ", best_late1$pos_group,
  "). RB was the worst first-round investment: Sharpe of **",
  round(rb_top10$sharpe_elite, 2), "** in the top 10, negative across every tier.\n\n",
  "The advice was simple: if you want a running back, wait until Day 2 or sign a free agent. ",
  "The data said so.\n\n",
  "But data without context is just math."
))
```

### How We Measure Quality

Before we dig in, a quick note on definitions. Part 1 and Part 2 measure "quality" differently, and that's deliberate.

**Part 1's "hit"** is snap-based: a player who plays above the positional snap threshold (75% of baseline snaps) earns starter status. Snaps measure quality -- if you're on the field, you're contributing. This is how Article 1 builds its Sharpe ratio.

**Part 2's "elite outcome"** uses [career Approximate Value](https://www.pro-football-reference.com/about/approximate_value.htm) (AV) from Pro Football Reference -- a cumulative measure of a player's total career contribution. We define an elite outcome as finishing in the **top 10% of career AV at their position** (90th percentile). This is a much stricter bar than Article 1's snap-based hit, and it's what we use throughout this article when evaluating blue-chip prospects and their outcomes.

Why the distinction? Snaps tell you who became a starter. Career AV tells you who became *great*. When we're asking whether a generational prospect justifies breaking the model, we need the higher bar.

---

## Argument 1: Roster Needs — "The Bills Don't Need Another QB"

Article 1 implicitly assumed that every team should always draft the highest-Sharpe position available. But what if you already have a starter?

The Bills are the perfect test case. From 2014 through 2022, Buffalo always had someone playing quarterback -- EJ Manuel, Kyle Orton, Tyrod Taylor, Matt Barkley, then Josh Allen. A snap-only model would say "QB is filled every year." But having a warm body at quarterback isn't the same as having a good one.

We added a quality gate: a position is only "filled" when the starter meets three criteria -- plays above the snap threshold, earned a second contract above the free-agent replacement cost for the position, and isn't on an expiring deal. A bad QB who plays a lot doesn't count. Tyrod Taylor started 46 games for the Bills and cleared every snap threshold, but he never earned a QB-level second contract -- the market's way of saying he wasn't the long-term answer.

```{r bills-qb-timeline, fig.height=5, fig.width=10}
bills_qb <- roster_needs %>%
  filter(team == "BUF", pos_group == "QB") %>%
  mutate(
    starter_name = case_when(
      pfr_player_id == "ManuEJ00" ~ "EJ Manuel",
      pfr_player_id == "OrtoKy00" ~ "Kyle Orton",
      pfr_player_id == "TaylTy00" ~ "Tyrod Taylor",
      pfr_player_id == "BarkMa00" ~ "Matt Barkley",
      pfr_player_id == "AlleJo02" ~ "Josh Allen",
      TRUE ~ pfr_player_id
    ),
    fill_label = if_else(is_filled, "Filled", "Need"),
    quality_label = case_when(
      starter_is_good ~ "Quality Starter",
      starter_is_hit & !starter_is_good ~ "Playing, Not Good Enough",
      TRUE ~ "No Starter"
    )
  )

ggplot(bills_qb, aes(x = draft_season, y = 1)) +
  geom_segment(aes(xend = draft_season, y = 0.6, yend = 1), color = "gray70", linewidth = 0.8) +
  geom_point(aes(color = is_filled), size = 6) +
  geom_text(aes(label = starter_name), vjust = -1.5, size = 3.5, fontface = "bold") +
  geom_text(aes(label = fill_label), vjust = 2.5, size = 3, color = "gray50") +
  scale_color_manual(values = c("TRUE" = "#2e7d32", "FALSE" = "#c62828"),
                     labels = c("TRUE" = "Position Filled", "FALSE" = "Position Need")) +
  scale_x_continuous(breaks = 2014:2022) +
  coord_cartesian(ylim = c(0.3, 1.7)) +
  labs(
    title = "Buffalo Bills: QB Need vs Filled (2014-2022, Quality Gate)",
    subtitle = "Red = starter lacked quality despite playing | Green = quality starter in place",
    x = NULL, y = NULL, color = NULL
  ) +
  theme_minimal(base_size = 14) +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "top"
  )
```

This is the interesting part. The model says QB has the best top-10 Sharpe. The Bills had a QB on the roster every year. But the quality gate reveals the truth: the model would have been **right** to tell them to draft QB in years where EJ Manuel or Matt Barkley were the answer -- and right to leave the position alone once Josh Allen arrived. The model's general advice ("draft QB") was correct in spirit, even when the roster looked superficially filled.

### How Need-Adjustment Changes the Rankings

In Part 1, we ranked teams by "draft efficiency" -- the average Sharpe ratio of their first-round picks. A team that spent top-10 picks on high-Sharpe positions (QB, OT, EDGE) scored well; a team that spent them on low-Sharpe positions (RB, S) scored poorly. It's a clean metric, but it ignores one obvious thing: whether the team actually needed a player at that position.

The need-adjusted version gives partial credit to teams that drafted low-Sharpe positions to fill genuine roster holes, and penalizes teams that drafted high-Sharpe positions they already had covered. The chart shows every team's original rank vs their need-adjusted rank.

```{r team-efficiency-slope, fig.height=9, fig.width=10}
te_plot <- team_efficiency %>%
  mutate(
    direction = case_when(
      rank_change > 2 ~ "Improved",
      rank_change < -2 ~ "Declined",
      TRUE ~ "Stable"
    )
  )

ggplot(te_plot, aes(y = reorder(team, -rank_adjusted))) +
  geom_segment(
    aes(x = rank_original, xend = rank_adjusted,
        yend = reorder(team, -rank_adjusted),
        color = direction),
    linewidth = 1.2, alpha = 0.7
  ) +
  geom_nfl_logos(aes(x = rank_adjusted, team_abbr = team), width = 0.04) +
  geom_point(aes(x = rank_original), shape = 1, size = 3, color = "gray50") +
  scale_color_manual(
    values = c("Improved" = "#2e7d32", "Declined" = "#c62828", "Stable" = "gray60"),
    name = "After Need Adjustment"
  ) +
  scale_x_reverse(breaks = seq(0, 32, 4)) +
  labs(
    title = "Draft Efficiency: Original Rank vs Need-Adjusted Rank",
    subtitle = "Open circle = original rank | Logo = need-adjusted rank | Left is better",
    x = "Rank (lower = better)", y = NULL
  ) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "top")
```

```{r top-movers, results='asis'}
biggest_up <- team_efficiency %>% slice_max(rank_change, n = 1)
biggest_down <- team_efficiency %>% slice_min(rank_change, n = 1)

cat(paste0(
  "**Biggest climber:** ", biggest_up$team, " moved up ", biggest_up$rank_change,
  " spots — they addressed real needs with their first-round picks. ",
  "**Biggest faller:** ", biggest_down$team, " dropped ", abs(biggest_down$rank_change),
  " spots — they kept drafting positions they already had filled."
))
```

### Need-Based Picks vs Filled-Position Picks

Do picks that address roster needs actually perform differently? Here's the raw Sharpe distribution for first-round picks that filled a need vs picks at positions the team already had covered:

```{r need-vs-filled-box, fig.height=6, fig.width=10}
r1_context <- r1_picks %>%
  mutate(
    context = if_else(is_need, "Addressed a Need", "Position Already Filled")
  )

ggplot(r1_context, aes(x = sharpe_adjusted, y = context, fill = context)) +
  geom_boxplot(alpha = 0.7, outlier.shape = 21, outlier.alpha = 0.5, width = 0.5) +
  geom_jitter(height = 0.15, alpha = 0.3, size = 2) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
  scale_fill_manual(values = c("Addressed a Need" = "#2196F3", "Position Already Filled" = "#FF9800")) +
  labs(
    title = "Need-Adjusted Sharpe: Picks That Filled Needs vs Filled Positions",
    subtitle = "Dashed line = break-even | Each dot is a first-round pick (2014-2022)",
    x = "Need-Adjusted Sharpe Ratio", y = NULL, fill = NULL
  ) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none")
```

```{r need-vs-filled-stats, results='asis'}
need_stats <- r1_context %>%
  group_by(context) %>%
  summarise(
    mean_adj = mean(sharpe_adjusted, na.rm = TRUE),
    median_adj = median(sharpe_adjusted, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  )
need_row <- need_stats %>% filter(context == "Addressed a Need")
filled_row <- need_stats %>% filter(context == "Position Already Filled")

cat(paste0(
  "Picks addressing real needs average a need-adjusted Sharpe of **",
  round(need_row$mean_adj, 3), "** (n=", need_row$n,
  "), while picks at already-filled positions average **",
  round(filled_row$mean_adj, 3), "** (n=", filled_row$n,
  "). The gap isn't just theoretical — teams that ignore roster context ",
  "systematically underperform their draft capital."
))
```

**The takeaway:** The Sharpe ratio tells you which positions have the best *average* returns. But if you already have a franchise QB, that average is irrelevant to you. Roster context turns a general truth into specific advice.

---

## Argument 2: Draft Class Depth — "Not All RB Classes Are Created Equal"

The Sharpe ratio averages across 13 draft classes (2010-2022). But class quality varies wildly. If a GM could know whether this year's class is loaded or barren *before* the draft, the model's one-size-fits-all advice starts to crack.

We actually can measure this -- at least partially. NFL Next Gen Stats publishes composite prospect scores before the draft, combining athletic testing, college production, and size. These scores exist before a single snap is played in the NFL.

### Prospective Class Quality: What GMs Can Actually See

NGS uses four tiers: **Below Average** (50-60), **Average** (60-70), **Good** (70-90), and **Elite** (90+). Instead of looking at mean scores (which compress into a narrow band), we count how many prospects in each class grade as Good or better. A class with five Good+ prospects is fundamentally different from one with two, even if their averages are similar.

We'll start with RB -- the position at the center of the Sharpe debate -- then see how the pattern plays out across every position group.

```{r ngs-rb-depth-big, fig.height=6, fig.width=10}
ngs_rb_tiers <- ngs_class_tiers %>%
  filter(pos_group == "RB", season >= 2010, season <= 2025,
         ngs_tier %in% c("Elite", "Good")) %>%
  mutate(ngs_tier = factor(ngs_tier, levels = c("Elite", "Good")))

ggplot(ngs_rb_tiers, aes(x = factor(season), y = n, fill = ngs_tier)) +
  geom_col(width = 0.7, alpha = 0.85) +
  scale_fill_manual(
    values = c("Elite" = "#1565C0", "Good" = "#66BB6A"),
    name = "NGS Tier"
  ) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  labs(
    title = "RB Draft Class Depth: Good and Elite Prospects by Year",
    subtitle = "Only prospects scoring Good (70-90) or Elite (90+) on NGS Combined Score",
    x = NULL, y = "Number of Good+ Prospects"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "top"
  )
```

The variation is striking. Some RB classes are loaded with Good+ talent; others barely register. A GM drafting RB in a deep year is playing a fundamentally different game than one drafting in a thin year -- but the Sharpe ratio treats them identically.

Now the same view across every position:

```{r ngs-all-depth-small, fig.height=14, fig.width=12}
ngs_other_tiers <- ngs_class_tiers %>%
  filter(pos_group %in% key_positions, season >= 2010, season <= 2025,
         ngs_tier %in% c("Elite", "Good")) %>%
  mutate(
    ngs_tier = factor(ngs_tier, levels = c("Elite", "Good")),
    pos_group = factor(pos_group, levels = key_positions)
  )

ggplot(ngs_other_tiers, aes(x = factor(season), y = n, fill = ngs_tier)) +
  geom_col(width = 0.7, alpha = 0.85) +
  scale_fill_manual(
    values = c("Elite" = "#1565C0", "Good" = "#66BB6A"),
    name = "NGS Tier"
  ) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  facet_wrap(~pos_group, scales = "free_y", ncol = 3) +
  labs(
    title = "Draft Class Depth Across All Positions",
    subtitle = "Good (70-90) and Elite (90+) NGS prospects by year | Each position tells a different story",
    x = NULL, y = "Number of Good+ Prospects"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 7),
    legend.position = "top",
    strip.text = element_text(face = "bold", size = 12)
  )
```

```{r ngs-class-narrative, results='asis'}
ngs_rb_class <- class_depth %>%
  filter(pos_group == "RB", !is.na(ngs_n_good_plus))
best_class <- ngs_rb_class %>% slice_max(ngs_n_good_plus, n = 1)
worst_class <- ngs_rb_class %>% slice_min(ngs_n_good_plus, n = 1)

cat(paste0(
  "The **", best_class$season,
  "** RB class had **", best_class$ngs_n_good_plus,
  "** prospects grading Good or better — the deepest pool of pre-draft talent in our sample. ",
  "The **", worst_class$season, "** class had just **", worst_class$ngs_n_good_plus,
  "**. That's the difference between a class where you can find value in any round and one ",
  "where you're chasing a single standout.\n\n",
  "The same pattern plays out at every position. Some years WR is loaded and EDGE is thin. ",
  "Some years the OT class is deep while TE is barren. QB class depth is notoriously spiky -- ",
  "one or two Elite prospects can define the entire class. The Sharpe ratio can't see any of this — ",
  "it averages across all years as if every class is the same. GMs know better."
))
```

```{r class-2025-commentary, results='asis'}
class_2025 <- ngs_class_tiers %>%
  filter(season == 2025, ngs_tier %in% c("Elite", "Good")) %>%
  group_by(pos_group) %>%
  summarise(
    n_good_plus = sum(n),
    n_elite = sum(n[ngs_tier == "Elite"]),
    .groups = "drop"
  ) %>%
  arrange(desc(n_good_plus))

deep_2025 <- class_2025 %>% slice_head(n = 3)
thin_2025 <- class_2025 %>% slice_tail(n = 3)

cat(paste0(
  "### The 2025 Class at a Glance\n\n",
  "Applying the same framework to the 2025 draft class: **",
  deep_2025$pos_group[1], "** (",  deep_2025$n_good_plus[1], " Good+), **",
  deep_2025$pos_group[2], "** (", deep_2025$n_good_plus[2], "), and **",
  deep_2025$pos_group[3], "** (", deep_2025$n_good_plus[3],
  ") are the deepest classes this year. The thinnest: **",
  thin_2025$pos_group[nrow(thin_2025)], "** (",
  thin_2025$n_good_plus[nrow(thin_2025)], " Good+), **",
  thin_2025$pos_group[nrow(thin_2025)-1], "** (",
  thin_2025$n_good_plus[nrow(thin_2025)-1],
  "). This is exactly the kind of information the Sharpe ratio can't incorporate — ",
  "the best position to draft depends on which classes are loaded *this year*."
))
```

### Did the Pre-Draft Signal Match Reality?

The NGS scores tell us what the class looked like before the draft. But did the "deep" classes actually produce more hits? Here's the RB timeline with retrospective hit rates -- bar color is the pre-draft NGS classification, bar height is what actually happened.

```{r rb-class-timeline-big, fig.height=7, fig.width=11}
rb_classes <- class_depth %>%
  filter(pos_group == "RB")

rb_notable <- tribble(
  ~season, ~label,
  2010, "0 hits",
  2011, "Murray",
  2012, "Martin, Morris",
  2013, "Bell, Lacy",
  2014, "Hill, Hyde",
  2015, "Gordon, Gurley,\nDavid Johnson",
  2016, "Zeke, Henry",
  2017, "CMC, Kamara,\nCook, Hunt,\nMixon, Jones",
  2018, "Saquon, Chubb",
  2019, "Jacobs, Sanders",
  2020, "Taylor, Swift",
  2021, "Harris, Javonte",
  2022, "Walker, Hall"
)

rb_classes <- rb_classes %>%
  left_join(rb_notable, by = "season")

ggplot(rb_classes, aes(x = factor(season), y = hit_rate, fill = ngs_class_quality)) +
  geom_col(width = 0.7, alpha = 0.85) +
  geom_text(
    aes(label = label),
    vjust = -0.3, size = 3, lineheight = 0.85
  ) +
  scale_fill_manual(
    values = c("Deep" = "#2e7d32", "Average" = "#FFA726", "Thin" = "#c62828"),
    na.value = "gray50"
  ) +
  scale_y_continuous(
    labels = percent_format(),
    expand = expansion(mult = c(0, 0.25))
  ) +
  labs(
    title = "RB Draft Class Quality: A Rollercoaster",
    subtitle = "Retrospective hit rate by draft year | Bar color = pre-draft NGS class quality",
    x = NULL, y = "Hit Rate", fill = "Class Quality"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "top"
  )
```

```{r rb-timeline-narrative, results='asis'}
ngs_rb_class <- class_depth %>%
  filter(pos_group == "RB", !is.na(ngs_n_good_plus))
ngs_rb_cor <- cor(ngs_rb_class$ngs_n_good_plus, ngs_rb_class$hit_rate, use = "complete.obs")

cat(paste0(
  "The colors (pre-draft class depth) and heights (what actually happened) track each other. ",
  "The correlation between a class's Good+ prospect count and its eventual hit rate is **",
  round(ngs_rb_cor, 2), "**.\n\n",
  "The mismatches are revealing. 2010 had the highest mean NGS score but produced zero hits — ",
  "athletic profiles alone don't guarantee production. Meanwhile, 2017 was the best actual class ",
  "(23.3% hit rate) because six different backs all found NFL roles.\n\n",
  "The model says \"never draft RB.\" The class data says \"it depends on the year.\" ",
  "And the NGS tier breakdown gives GMs a real window into which kind of year it is — ",
  "before the first pick is made."
))
```

The same signal-vs-reality check across other positions:

```{r other-pos-timeline-small, fig.height=14, fig.width=12}
other_classes <- class_depth %>%
  filter(pos_group %in% key_positions, pos_group != "RB",
         !is.na(hit_rate), !is.na(ngs_class_quality))

ggplot(other_classes, aes(x = factor(season), y = hit_rate, fill = ngs_class_quality)) +
  geom_col(width = 0.7, alpha = 0.85) +
  scale_fill_manual(
    values = c("Deep" = "#2e7d32", "Average" = "#FFA726", "Thin" = "#c62828"),
    na.value = "gray50",
    name = "Pre-Draft Class Quality"
  ) +
  scale_y_continuous(labels = percent_format(), expand = expansion(mult = c(0, 0.15))) +
  facet_wrap(~pos_group, scales = "free_y", ncol = 2) +
  labs(
    title = "Pre-Draft Signal vs Reality: All Positions",
    subtitle = "Retrospective hit rate by year | Bar color = pre-draft NGS class quality (deep/average/thin)",
    x = NULL, y = "Hit Rate"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 7),
    legend.position = "top",
    strip.text = element_text(face = "bold", size = 12)
  )
```

```{r other-timeline-narrative, results='asis'}
# Compute correlation for each position
pos_cors <- class_depth %>%
  filter(pos_group %in% key_positions, !is.na(ngs_n_good_plus), !is.na(hit_rate)) %>%
  group_by(pos_group) %>%
  summarise(
    cor = cor(ngs_n_good_plus, hit_rate, use = "complete.obs"),
    .groups = "drop"
  ) %>%
  arrange(desc(cor))

best_cor <- pos_cors %>% slice_head(n = 1)
worst_cor <- pos_cors %>% slice_tail(n = 1)

cat(paste0(
  "The pre-draft signal isn't equally predictive everywhere. **", best_cor$pos_group,
  "** shows the strongest correlation (r = ", round(best_cor$cor, 2),
  ") between Good+ prospect count and eventual hit rate, while **", worst_cor$pos_group,
  "** shows the weakest (r = ", round(worst_cor$cor, 2),
  "). Positions where athleticism translates more directly — like RB and EDGE — tend to have stronger ",
  "pre-draft signals. Positions where scheme fit and development matter more — like IOL and S — ",
  "show weaker connections between prospect scores and career outcomes."
))
```

---

## Argument 3: Blue-Chip Prospects — "When the Talent Is Undeniable"

The Sharpe ratio averages over all RB prospects. But we've already seen that not all classes are equal. The same logic applies to individual prospects: the model can't distinguish between an average back and a generational one.

The challenge is defining "blue chip" without hindsight. [Career Approximate Value](https://www.pro-football-reference.com/about/approximate_value.htm) (AV) -- our retrospective measure -- tells us *who turned out* to be elite, but a GM can't use it on draft day. NGS Combined Scores can. They're published before the draft, combining athletic testing, college production, and size into a single number. It's the closest thing to a prospective blue-chip identifier that exists.

**Definitions for this section:**

- **Pre-draft blue chip:** Top 10% of NGS Combined Score at the position (knowable before the draft)
- **Career blue chip:** Top 10% of [career Approximate Value](https://www.pro-football-reference.com/about/approximate_value.htm) at the position (only knowable in hindsight)
- **Hit (outcome measure):** Article 1's snap-based starter threshold — a player who plays above 75% of baseline snaps at their position. This is our outcome metric for comparing blue-chip categories because it's independent of career AV (which defines the career blue-chip group).

### NGS Scores and Career Outcomes

```{r ngs-blue-chip-scatter, fig.height=7, fig.width=10}
ngs_with_outcomes <- blue_chip %>%
  filter(!is.na(cmbn_score), !is.na(is_hit), !is.na(car_av),
         pos_group %in% key_positions, season <= 2022)

ggplot(ngs_with_outcomes, aes(x = cmbn_score, y = car_av, color = pos_group)) +
  geom_point(alpha = 0.4, size = 2) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1, alpha = 0.7) +
  scale_color_manual(values = pos_colors) +
  labs(
    title = "Do Pre-Draft NGS Scores Predict Career Value?",
    subtitle = "Each dot = a drafted player with both NGS score and career value | By position",
    x = "NGS Combined Score (pre-draft)", y = "Career Approximate Value",
    color = "Position"
  ) +
  theme_minimal(base_size = 14)
```

```{r ngs-outcome-narrative, results='asis'}
rb_ngs_cor <- ngs_with_outcomes %>%
  filter(pos_group == "RB") %>%
  summarise(r = cor(cmbn_score, car_av, use = "complete.obs"), n = n())

all_ngs_cor <- ngs_with_outcomes %>%
  summarise(r = cor(cmbn_score, car_av, use = "complete.obs"), n = n())

cat(paste0(
  "For RBs, the correlation between pre-draft NGS Combined Score and career value is **",
  round(rb_ngs_cor$r, 2), "** (n=", rb_ngs_cor$n,
  "). Across all positions it's **", round(all_ngs_cor$r, 2),
  "** (n=", all_ngs_cor$n,
  "). The relationship isn't deterministic -- plenty of high-scoring prospects bust, and ",
  "late-round gems score modestly -- but the signal is there. Higher NGS scores map to ",
  "better career outcomes at a rate that's hard to dismiss as noise."
))
```

### Pre-Draft Blue Chips vs Career Blue Chips

We can define "blue chip" two ways: retrospectively (top 10% of career AV at the position) and prospectively (top 10% of NGS Combined Score at the position). How much do they overlap -- and does it differ by position?

First, RB:

```{r ngs-vs-career-bc-rb-big, fig.height=5, fig.width=10}
bc_overlap <- ngs_with_outcomes %>%
  mutate(
    bc_type = case_when(
      is_blue_chip & is_ngs_blue_chip ~ "Both",
      is_blue_chip & !is_ngs_blue_chip ~ "Career Only",
      !is_blue_chip & is_ngs_blue_chip ~ "Pre-Draft Only",
      TRUE ~ "Neither"
    )
  )

bc_by_pos <- bc_overlap %>%
  filter(bc_type != "Neither") %>%
  group_by(pos_group, bc_type) %>%
  summarise(
    n = n(),
    hit_rate = mean(is_hit, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    bc_type = factor(bc_type, levels = c("Both", "Pre-Draft Only", "Career Only")),
    pos_group = factor(pos_group, levels = key_positions)
  )

bc_rb <- bc_by_pos %>% filter(pos_group == "RB")

ggplot(bc_rb, aes(x = hit_rate, y = bc_type, fill = bc_type)) +
  geom_col(width = 0.6, alpha = 0.85) +
  geom_text(aes(label = paste0(round(hit_rate * 100), "% (n=", n, ")")),
            hjust = -0.05, size = 4, fontface = "bold") +
  scale_fill_manual(values = c(
    "Both" = "#2e7d32",
    "Pre-Draft Only" = "#377EB8",
    "Career Only" = "#FF9800"
  ), guide = "none") +
  scale_x_continuous(labels = percent_format(), limits = c(0, 1.15),
                     expand = expansion(mult = c(0, 0))) +
  labs(
    title = "RB: Hit Rate by Blue-Chip Definition",
    subtitle = "Hit = Article 1's snap-based starter threshold | Pre-draft BC = top 10% NGS | Career BC = top 10% career AV",
    x = "Hit Rate (snap-based starter)", y = NULL
  ) +
  theme_minimal(base_size = 14)
```

```{r ngs-bc-rb-narrative, results='asis'}
rb_both <- bc_overlap %>% filter(bc_type == "Both", pos_group == "RB")
rb_ngs_only <- bc_overlap %>% filter(bc_type == "Pre-Draft Only", pos_group == "RB")
rb_baseline <- ngs_with_outcomes %>% filter(pos_group == "RB") %>%
  summarise(rate = mean(is_hit, na.rm = TRUE)) %>% pull(rate)

cat(paste0(
  "For RBs identified as blue chips by **both** pre-draft and career metrics, the hit rate is **",
  round(mean(rb_both$is_hit, na.rm = TRUE) * 100, 1),
  "%**. Pre-draft-only blue chips (high NGS but didn't reach top 10% career AV) still hit at **",
  round(mean(rb_ngs_only$is_hit, na.rm = TRUE) * 100, 1),
  "%** — compared to a baseline of **", round(rb_baseline * 100, 1),
  "%** for all RBs with NGS scores. The pre-draft signal is meaningful.\n\n",
  "How does this compare across positions?"
))
```

```{r ngs-vs-career-bc-all-small, fig.height=14, fig.width=12}
bc_other <- bc_by_pos %>% filter(pos_group != "RB")

ggplot(bc_other, aes(x = hit_rate, y = bc_type, fill = bc_type)) +
  geom_col(width = 0.6, alpha = 0.85) +
  geom_text(aes(label = paste0(round(hit_rate * 100), "% (n=", n, ")")),
            hjust = -0.05, size = 2.8, fontface = "bold") +
  scale_fill_manual(values = c(
    "Both" = "#2e7d32",
    "Pre-Draft Only" = "#377EB8",
    "Career Only" = "#FF9800"
  ), guide = "none") +
  scale_x_continuous(labels = percent_format(), limits = c(0, 1.15),
                     expand = expansion(mult = c(0, 0))) +
  facet_wrap(~pos_group, ncol = 2) +
  labs(
    title = "Hit Rate by Blue-Chip Definition: All Positions",
    subtitle = "Hit = Article 1's snap-based starter | Pre-draft BC = top 10% NGS | Career BC = top 10% career AV",
    x = "Hit Rate (snap-based starter)", y = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(strip.text = element_text(face = "bold", size = 12))
```

```{r ngs-bc-all-narrative, results='asis'}
both_all <- bc_overlap %>% filter(bc_type == "Both")
ngs_only_all <- bc_overlap %>% filter(bc_type == "Pre-Draft Only")
baseline_all <- mean(ngs_with_outcomes$is_hit, na.rm = TRUE)

cat(paste0(
  "Across all positions, players who were blue chips by **both** definitions became starters at **",
  round(mean(both_all$is_hit, na.rm = TRUE) * 100, 1),
  "%**. Pre-draft-only blue chips hit at **",
  round(mean(ngs_only_all$is_hit, na.rm = TRUE) * 100, 1),
  "%** — compared to a baseline of **", round(baseline_all * 100, 1),
  "%** across all players with NGS scores.\n\n",
  "The pre-draft signal isn't equally useful everywhere — it's strongest at positions ",
  "where athletic profiles translate most directly to NFL production, and weakest where ",
  "scheme fit and coaching development matter more. But at every position, ",
  "the top pre-draft prospects outperform the field."
))
```

### The Historical NGS Elite

Who were the highest-graded pre-draft prospects at each position, and did they deliver? Starting with RB:

```{r ngs-elite-rb-big, fig.height=6, fig.width=10}
top_ngs_rb <- ngs_with_outcomes %>%
  filter(pos_group == "RB") %>%
  arrange(desc(cmbn_score)) %>%
  slice_head(n = 8) %>%
  mutate(
    player_label = paste0(pfr_player_name, " ('", str_sub(season, 3, 4), ")"),
    player_label = fct_reorder(player_label, cmbn_score),
    outcome = if_else(is_hit, "Hit", "Miss")
  )

ggplot(top_ngs_rb, aes(x = cmbn_score, y = player_label, fill = outcome)) +
  geom_col(width = 0.7, alpha = 0.85) +
  geom_text(aes(label = round(cmbn_score, 1)), hjust = -0.15, size = 4, fontface = "bold") +
  scale_fill_manual(values = c("Hit" = "#2e7d32", "Miss" = "#c62828")) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(
    title = "Top 8 RB Prospects by NGS Combined Score (2010-2022)",
    subtitle = "Pre-draft NGS score vs career outcome | Green = became a starter",
    x = "NGS Combined Score", y = NULL, fill = "Career\nOutcome"
  ) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "top")
```

```{r ngs-elite-rb-narrative, results='asis'}
rb_elite_stats <- top_ngs_rb %>%
  summarise(hit_rate = mean(is_hit, na.rm = TRUE), n = n())

cat(paste0(
  "The top 8 RB prospects by pre-draft NGS score hit at **",
  round(rb_elite_stats$hit_rate * 100, 1),
  "%** — far above the position's overall average. The names are what you'd expect: ",
  "Jonathan Taylor, Saquon Barkley, Breece Hall, Derrick Henry. These aren't random hits — ",
  "the pre-draft signal identified them before a single NFL snap.\n\n",
  "Now the same view across all positions:"
))
```

```{r ngs-elite-all-small, fig.height=18, fig.width=12}
top_ngs_all <- ngs_with_outcomes %>%
  filter(pos_group %in% key_positions) %>%
  group_by(pos_group) %>%
  arrange(desc(cmbn_score)) %>%
  slice_head(n = 6) %>%
  ungroup() %>%
  mutate(
    player_label = paste0(pfr_player_name, " ('", str_sub(season, 3, 4), ")"),
    player_label_unique = paste0(pos_group, "___", player_label),
    player_label_unique = fct_reorder(player_label_unique, cmbn_score),
    outcome = if_else(is_hit, "Hit", "Miss"),
    pos_group = factor(pos_group, levels = key_positions)
  )

ggplot(top_ngs_all, aes(x = cmbn_score, y = player_label_unique, fill = outcome)) +
  geom_col(width = 0.7, alpha = 0.85) +
  geom_text(aes(label = round(cmbn_score, 1)), hjust = -0.15, size = 2.8, fontface = "bold") +
  scale_fill_manual(values = c("Hit" = "#2e7d32", "Miss" = "#c62828")) +
  scale_y_discrete(labels = function(x) str_remove(x, "^.*___")) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.12))) +
  facet_wrap(~pos_group, scales = "free", ncol = 2) +
  labs(
    title = "Top 6 Historical Players by NGS Combined Score: All Positions",
    subtitle = "Pre-draft NGS score vs career outcome | Green = became a starter",
    x = "NGS Combined Score", y = NULL, fill = "Career\nOutcome"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "top",
    strip.text = element_text(face = "bold", size = 12)
  )
```

```{r ngs-elite-all-narrative, results='asis'}
elite_stats <- ngs_with_outcomes %>%
  filter(pos_group %in% key_positions) %>%
  group_by(pos_group) %>%
  arrange(desc(cmbn_score)) %>%
  slice_head(n = 6) %>%
  summarise(
    hit_rate = mean(is_hit, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  ) %>%
  arrange(desc(hit_rate))

best_elite <- elite_stats %>% slice_head(n = 1)
worst_elite <- elite_stats %>% slice_tail(n = 1)

cat(paste0(
  "**", best_elite$pos_group, "** leads the pack — the top 6 NGS scorers hit at ",
  round(best_elite$hit_rate * 100, 1),
  "%. **", worst_elite$pos_group, "** has the weakest pre-draft-to-career translation (",
  round(worst_elite$hit_rate * 100, 1),
  "%). The model's \"never draft RB\" ",
  "is really \"never draft an *average* RB.\" When the pre-draft signal says a prospect is ",
  "separated from the class, the historical hit rate is dramatically higher at every position."
))
```

---

## The Love Question Revisited

In Part 1, we asked: *Should you draft Jeremiah Love at 10?*

The model said no. RB has a deeply negative Sharpe in the top 10. But we've now shown that the answer depends on things the model can't see: roster needs, class depth, and prospect separation. Love is the case that brings all three together.

### Where Love Sits in the NGS Historical Record

We showed above that historical RBs with elite pre-draft NGS scores hit at dramatically higher rates than the position average. Where does Love's 2025 profile fit in that history?

```{r ngs-love-bar-big, fig.height=7, fig.width=10}
ngs_rb <- ngs_prospects %>%
  filter(pos_group == "RB", !is.na(cmbn_score)) %>%
  arrange(desc(cmbn_score)) %>%
  mutate(
    is_love = str_detect(player, "(?i)love"),
    player_label = paste0(player, " (", college, ")"),
    player_label = fct_reorder(player_label, cmbn_score)
  )

ggplot(ngs_rb, aes(x = cmbn_score, y = player_label, fill = is_love)) +
  geom_col(width = 0.7, alpha = 0.85) +
  geom_text(aes(label = round(cmbn_score, 1)), hjust = -0.15, size = 3.5, fontface = "bold") +
  scale_fill_manual(values = c("TRUE" = "#377EB8", "FALSE" = "gray60"), guide = "none") +
  scale_x_continuous(limits = c(0, 105), expand = expansion(mult = c(0, 0.05))) +
  labs(
    title = "2025 RB Prospects: NGS Combined Score",
    subtitle = "Love's 94.3 is tied for #1 overall (all positions) with Sonny Styles | ProdScr: 95.9 (highest of any prospect)",
    x = "NGS Combined Score", y = NULL
  ) +
  theme_minimal(base_size = 14)
```

```{r ngs-love-narrative, results='asis'}
love_row <- ngs_rb %>% filter(is_love) %>% slice(1)
rb2_row <- ngs_rb %>% filter(!is_love) %>% slice_max(cmbn_score, n = 1)
gap <- love_row$cmbn_score - rb2_row$cmbn_score

cat(paste0(
  "Love's NGS Combined Score of **", round(love_row$cmbn_score, 1),
  "** is tied for #1 overall across all positions. His Production Score of **95.9** is the highest ",
  "of any prospect in the 2025 class. The gap to RB2 (", rb2_row$player, " at ",
  round(rb2_row$cmbn_score, 1), ") is **", round(gap, 1), " points** -- a chasm by NGS standards.\n\n",
  "But Love isn't the only outlier in 2025. Here's a look at the top prospect at every position:"
))
```

```{r ngs-2025-all-positions, fig.height=14, fig.width=12}
# Top 5 prospects by position in 2025 class
top_2025_all <- ngs_prospects %>%
  filter(!is.na(cmbn_score), pos_group %in% key_positions) %>%
  group_by(pos_group) %>%
  arrange(desc(cmbn_score)) %>%
  slice_head(n = 5) %>%
  ungroup() %>%
  mutate(
    player_label = paste0(player, " (", college, ")"),
    player_label_unique = paste0(pos_group, "___", player_label),
    player_label_unique = fct_reorder(player_label_unique, cmbn_score),
    is_top = row_number() <= 1,
    pos_group = factor(pos_group, levels = key_positions)
  ) %>%
  # Re-compute is_top within each position
  group_by(pos_group) %>%
  mutate(is_top = cmbn_score == max(cmbn_score)) %>%
  ungroup()

ggplot(top_2025_all, aes(x = cmbn_score, y = player_label_unique, fill = is_top)) +
  geom_col(width = 0.7, alpha = 0.85) +
  geom_text(aes(label = round(cmbn_score, 1)), hjust = -0.15, size = 2.5, fontface = "bold") +
  scale_fill_manual(values = c("TRUE" = "#1565C0", "FALSE" = "gray60"), guide = "none") +
  scale_y_discrete(labels = function(x) str_remove(x, "^.*___")) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.12))) +
  facet_wrap(~pos_group, scales = "free", ncol = 3) +
  labs(
    title = "2025 Draft Class: Top 5 Prospects by NGS Combined Score, Every Position",
    subtitle = "Blue = top prospect at position | How separated is #1 from the rest?",
    x = "NGS Combined Score", y = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "none",
    strip.text = element_text(face = "bold", size = 12)
  )
```

```{r ngs-2025-all-narrative, results='asis'}
# Find positions with biggest separation
pos_sep <- ngs_prospects %>%
  filter(!is.na(cmbn_score), pos_group %in% key_positions) %>%
  group_by(pos_group) %>%
  arrange(desc(cmbn_score)) %>%
  summarise(
    top_player = first(player),
    top_score = first(cmbn_score),
    second_score = nth(cmbn_score, 2),
    gap = first(cmbn_score) - nth(cmbn_score, 2),
    .groups = "drop"
  ) %>%
  arrange(desc(gap))

biggest_gap <- pos_sep %>% slice_head(n = 1)
smallest_gap <- pos_sep %>% slice_tail(n = 1)

cat(paste0(
  "The 2025 class has fascinating separation patterns. **", biggest_gap$pos_group,
  "** has the biggest gap between the top prospect (",
  biggest_gap$top_player, " at ", round(biggest_gap$top_score, 1),
  ") and #2 — a **", round(biggest_gap$gap, 1),
  "-point** chasm. Meanwhile, **", smallest_gap$pos_group,
  "** is the tightest race, with just a **", round(smallest_gap$gap, 1),
  "-point** gap at the top. Where a prospect is clearly separated, the blue-chip argument ",
  "gets stronger. Where the class is tightly bunched, the model's average-case advice holds."
))
```

### How Good Does Love Need to Be?

This is the practical question. Article 1 gave us the Sharpe ratio for RB in the top 10: **-0.615**. That's the average across all RBs drafted in the top 10 since 2010. But averages hide a massive range. Here's every first-round RB in our data, plotted against the opportunity cost of not drafting the best available non-RB at the same pick.

```{r rb-breakeven, fig.height=8, fig.width=11}
# All first-round picks with individual Sharpe
all_r1 <- player_lookup %>%
  filter(pick <= 32, pos_group %in% key_positions)

# Compute opportunity cost at each pick: mean Sharpe of non-RB picks in the same tier
opp_cost_by_tier <- all_r1 %>%
  filter(pos_group != "RB") %>%
  mutate(tier = factor(tier, levels = tier_order)) %>%
  group_by(tier) %>%
  summarise(
    opp_cost_sharpe = mean(player_sharpe, na.rm = TRUE),
    opp_cost_return = mean(player_return, na.rm = TRUE),
    .groups = "drop"
  )

# RB first-rounders with their tier
rb_r1 <- all_r1 %>%
  filter(pos_group == "RB") %>%
  left_join(opp_cost_by_tier, by = "tier") %>%
  mutate(
    beat_opp = player_sharpe > opp_cost_sharpe,
    name_label = paste0(pfr_player_name, "\n('", str_sub(season, 3, 4), ", #", pick, ")")
  )

ggplot(rb_r1, aes(x = pick, y = player_sharpe)) +
  # Opportunity cost line
  geom_hline(data = opp_cost_by_tier %>% filter(tier %in% c("Top 10", "Late 1st")),
             aes(yintercept = opp_cost_sharpe, linetype = tier),
             color = "gray40", linewidth = 0.8) +
  # Zero line
  geom_hline(yintercept = 0, color = "gray80", linewidth = 0.5) +
  # Individual RBs
  geom_point(aes(color = beat_opp), size = 4, alpha = 0.8) +
  geom_text(aes(label = pfr_player_name), vjust = -1.2, size = 2.8, fontface = "bold") +
  # CMC and Gurley highlighted
  geom_point(data = rb_r1 %>% filter(pfr_player_name %in% c("Christian McCaffrey", "Todd Gurley")),
             color = "#377EB8", size = 5, shape = 1, stroke = 2) +
  scale_color_manual(values = c("TRUE" = "#2e7d32", "FALSE" = "#c62828"),
                     labels = c("TRUE" = "Beat opportunity cost", "FALSE" = "Below opportunity cost"),
                     name = "vs. Non-RB Alternative") +
  scale_linetype_manual(values = c("Top 10" = "dashed", "Late 1st" = "dotted"),
                        name = "Avg Non-RB Sharpe") +
  scale_x_continuous(breaks = seq(0, 32, 4)) +
  labs(
    title = "The RB Gamble: Individual Sharpe vs Opportunity Cost",
    subtitle = "Each dot = a first-round RB | Dashed line = avg non-RB Sharpe in that tier | Blue circles = CMC & Gurley",
    x = "Draft Pick", y = "Individual Player Sharpe"
  ) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "top")
```

```{r rb-breakeven-narrative, results='asis'}
# Key stats
rb_top10_sharpe <- sharpe %>% filter(pos_group == "RB", tier == "Top 10") %>% pull(sharpe_elite)
cmc <- rb_r1 %>% filter(pfr_player_name == "Christian McCaffrey")
gurley <- rb_r1 %>% filter(pfr_player_name == "Todd Gurley")
opp_top10 <- opp_cost_by_tier %>% filter(tier == "Top 10") %>% pull(opp_cost_sharpe)
rb_beat_pct <- mean(rb_r1$beat_opp, na.rm = TRUE)

cat(paste0(
  "Article 1's tier-level Sharpe for RB in the top 10 is **", round(rb_top10_sharpe, 3),
  "** — the worst of any position. But that number averages together Christian McCaffrey (",
  round(cmc$player_sharpe, 2), ") with Trent Richardson (0.06). The range is enormous.\n\n",
  "The dashed line shows the opportunity cost: the average Sharpe of non-RB picks in the top 10 (",
  round(opp_top10, 2), "). To justify the pick, an RB needs to land **above that line**. ",
  "Only **", round(rb_beat_pct * 100), "%** of first-round RBs did.\n\n",
  "**CMC at pick 8** (Sharpe: ", round(cmc$player_sharpe, 2),
  ") and **Gurley at pick 10** (Sharpe: ", round(gurley$player_sharpe, 2),
  ") are the RB ceiling — both comfortably beat the opportunity cost. ",
  "Love needs to be in that tier, not the Trent Richardson tier, to justify a top-10 pick.\n\n",
  "The NGS data says Love's profile looks like the CMC/Gurley cluster, not the Richardson cluster. ",
  "But the model's warning is clear: even among first-round RBs, most don't clear the bar."
))
```

### Sharpe at Pick 3: The Steepest Cost

```{r sharpe-at-3, results='asis'}
# What do other positions return at pick 3 specifically?
pick_3_5 <- player_lookup %>%
  filter(pick >= 2, pick <= 5) %>%
  group_by(pos_group) %>%
  summarise(
    mean_sharpe = mean(player_sharpe, na.rm = TRUE),
    mean_return = mean(player_return, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  ) %>%
  arrange(desc(mean_sharpe))

best_at_3 <- pick_3_5 %>% slice_head(n = 3)
rb_at_3 <- pick_3_5 %>% filter(pos_group == "RB")

cat(paste0(
  "At picks 2-5 specifically, the opportunity cost is highest. The best-returning positions: **",
  best_at_3$pos_group[1], "** (avg Sharpe ", round(best_at_3$mean_sharpe[1], 2),
  "), **", best_at_3$pos_group[2], "** (", round(best_at_3$mean_sharpe[2], 2),
  "), **", best_at_3$pos_group[3], "** (", round(best_at_3$mean_sharpe[3], 2),
  "). RB at picks 2-5 averages just **", round(rb_at_3$mean_sharpe, 2),
  "** — that's Saquon Barkley (#2), Trent Richardson (#3), Zeke (#4), and Fournette (#4).\n\n",
  "At pick 3, you're giving up the most draft capital. The only RB ever drafted at 3 in our window ",
  "was Trent Richardson, who returned almost nothing. Taking Love at 3 would require him to be ",
  "a **CMC-level outcome** just to match what an average EDGE or QB pick would have returned there.\n\n",
  "This is where Article 1's advice is strongest. The higher the pick, the more the RB needs to ",
  "overcome. The case for Love gets better as you slide down the board."
))
```

### Where in the First Round Does Love Make Sense?

Here's the real question: at what pick does an RB with Love's profile start to justify the cost? We can compute this by comparing the CMC/Gurley outcome (the realistic ceiling for an elite RB) against the average non-RB return at each pick range.

```{r love-pick-range, fig.height=6, fig.width=10}
# Compare CMC/Gurley-level return vs opportunity cost across the first round
pick_ranges <- player_lookup %>%
  filter(pick <= 32) %>%
  mutate(pick_range = case_when(
    pick <= 5 ~ "1-5",
    pick <= 10 ~ "6-10",
    pick <= 16 ~ "11-16",
    pick <= 24 ~ "17-24",
    pick <= 32 ~ "25-32"
  )) %>%
  mutate(pick_range = factor(pick_range, levels = c("1-5", "6-10", "11-16", "17-24", "25-32")))

# Average non-RB return by pick range
opp_by_range <- pick_ranges %>%
  filter(pos_group != "RB") %>%
  group_by(pick_range) %>%
  summarise(mean_sharpe = mean(player_sharpe, na.rm = TRUE), .groups = "drop") %>%
  mutate(type = "Avg Non-RB Pick")

# RB hits (CMC/Gurley tier — Sharpe > 2.0) by pick range
rb_ceiling <- pick_ranges %>%
  filter(pos_group == "RB", player_sharpe > 2.0) %>%
  group_by(pick_range) %>%
  summarise(mean_sharpe = mean(player_sharpe, na.rm = TRUE), .groups = "drop") %>%
  mutate(type = "Elite RB (CMC/Gurley tier)")

# All RB average
rb_avg <- pick_ranges %>%
  filter(pos_group == "RB") %>%
  group_by(pick_range) %>%
  summarise(mean_sharpe = mean(player_sharpe, na.rm = TRUE), .groups = "drop") %>%
  mutate(type = "Avg RB Pick")

compare <- bind_rows(opp_by_range, rb_ceiling, rb_avg) %>%
  mutate(type = factor(type, levels = c("Elite RB (CMC/Gurley tier)", "Avg Non-RB Pick", "Avg RB Pick")))

ggplot(compare, aes(x = pick_range, y = mean_sharpe, fill = type)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6, alpha = 0.85) +
  geom_hline(yintercept = 0, color = "gray60", linewidth = 0.5) +
  scale_fill_manual(values = c(
    "Elite RB (CMC/Gurley tier)" = "#377EB8",
    "Avg Non-RB Pick" = "#FF9800",
    "Avg RB Pick" = "#c62828"
  )) +
  labs(
    title = "Where Does an Elite RB Justify the Pick?",
    subtitle = "CMC/Gurley-tier RB Sharpe vs average non-RB alternative | By first-round pick range",
    x = "Pick Range", y = "Average Individual Sharpe", fill = NULL
  ) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "top")
```

```{r love-pick-narrative, results='asis'}
cat(paste0(
  "The blue bars are the CMC/Gurley tier — first-round RBs who hit at elite levels. ",
  "The orange bars are what you'd expect from the average non-RB pick in the same range. ",
  "The red bars are the average RB — the number Article 1 is built on.\n\n",
  "At picks 1-5, even an elite RB barely matches the average non-RB alternative. The opportunity cost is simply too high. ",
  "At picks 6-10, the elite RB **clears** the bar — this is where CMC (pick 8) and Gurley (pick 10) ",
  "actually lived, and both delivered positive returns above what the average alternative position offered. ",
  "In the late first, the math gets more favorable, but elite RB talent is rarely available that late.\n\n",
  "**The verdict: Love at pick 3-5 is a losing bet even if he hits.** The opportunity cost is too steep. ",
  "But Love at pick 6-10 — if he's truly in the CMC/Gurley tier — is where the Sharpe model's ",
  "advice starts to crack. And at picks 15+, an elite RB profile can justify the selection outright."
))
```

### Every Team Needs an RB (And That's the Point)

```{r all-team-narrative, results='asis'}
all_rb_needs <- roster_needs %>%
  filter(pos_group == "RB") %>%
  group_by(team) %>%
  summarise(need_rate = mean(!is_filled, na.rm = TRUE), .groups = "drop")

pct_always_need <- mean(all_rb_needs$need_rate == 1)
pct_mostly_need <- mean(all_rb_needs$need_rate >= 0.67)
lowest_need <- all_rb_needs %>% slice_min(need_rate, n = 1)

cat(paste0(
  "Here's the twist that reframes the entire conversation: when we apply our quality gate to RB across all 32 teams, ",
  "**", round(pct_always_need * 100), "%** of teams had RB as a need in *every single season* from 2014-2022. ",
  "**", round(pct_mostly_need * 100), "%** had it as a need at least two-thirds of the time. ",
  "Even the team with the lowest RB need rate — ", lowest_need$team, " at ",
  round(lowest_need$need_rate * 100), "% — still needed an RB more often than not.\n\n",
  "Why is RB *always* a need? Because our quality gate requires a starter who earned a **premium second contract**. ",
  "And RBs almost never get premium second contracts — that's exactly what Article 1's Sharpe ratio measured. ",
  "The position's low market value means almost no RB clears the quality gate.\n\n",
  "This is the Sharpe ratio's own logic turned back on itself. The model says RBs aren't worth premium draft capital. ",
  "The market agrees — it doesn't pay them. But that means the position is **perpetually unfilled** by the quality standard. ",
  "The question isn't *whether* to invest in RB — every team needs one. It's whether THIS investment, ",
  "in THIS prospect, in THIS class, justifies accepting the negative positional Sharpe.\n\n",
  "For Love, the case is: his NGS profile matches the CMC/Gurley tier. The class is thin — a 17-point gap to RB2. ",
  "And at picks 6-10, historical elite RBs have returned above the opportunity cost. ",
  "That doesn't guarantee he'll work out. But it means the model's blanket \"no\" is answering the wrong question."
))
```

### Putting It Together

Everything we've built in this article converges on a single question. Three arguments, one prospect:

<div style="background-color: #f0f4f8; border-left: 4px solid #377EB8; padding: 15px 20px; margin: 20px 0; border-radius: 4px;">

**The Sharpe ratio says:** RB is the worst first-round investment. Sharpe of -0.615 in the top 10. Negative across every tier. Don't do it.

**The class signal says:** 2025 is a thin RB class. If you pass on Love, the drop-off is steep -- a 17-point gap to RB2.

**The NGS history says:** RBs with scores like Love's (94.3) have historically hit at rates far above the position average. The pre-draft signal maps to career outcomes. CMC and Gurley — the RB ceiling — had individual Sharpes above 2.0.

**The breakeven says:** At picks 1-5, even an elite RB barely clears the opportunity cost. At picks 6-10, the math works — if the prospect is truly separated. At 15+, an elite profile justifies the pick outright.

**The need says:** Every team needs an RB, always. The question isn't whether to invest — it's where and in whom.

</div>

Does this mean a team *should* draft Love? Not at 3. Probably not at 5. But at 8-10, for a team that needs an RB? The Sharpe model says no. The prospect data says he's in the CMC/Gurley cluster. The class is thin. And history says the few RBs who clear the bar at that range more than justify the pick.

That's the honest answer. Not a prescription, but an acknowledgment that the model's average-case advice meets its limit when roster need, prospect separation, and class depth all point the other direction — and the pick falls in the range where elite RBs have historically delivered.

---

## What We Learned

We set out to stress-test the Sharpe ratio, not to destroy it. Three arguments, three perspectives. Here's what we found.

The Sharpe ratio is a powerful tool for general draft strategy. But it operates at the level of averages, and GMs operate at the level of specifics. The model can't account for three things: **roster context** (do you actually need this position?), **class variance** (is this a good year to draft it?), and **prospect separation** (is this prospect different from the rest?).

When those factors align -- a genuine need, a weak class at that position, and a prospect whose pre-draft profile is separated from the field -- the model's blanket advice meets a real edge case. Not a refutation, but a boundary condition.

The NGS data gave us something the retrospective analysis couldn't: a prospective signal. Historical RBs with elite NGS Combined Scores hit at rates dramatically above the position average. The correlation between pre-draft scores and career outcomes isn't perfect, but it's real enough to challenge the assumption that all prospects at a position are interchangeable.

None of this proves that any specific pick will work. The Sharpe ratio is right far more often than it's wrong. But when teams draft a low-Sharpe position and it works out, it's often because they had information the model didn't -- roster need, class quality, prospect separation. That's not irrationality. That's context.

---

## What Survived the Stress Test

The counter-arguments don't destroy the model -- they define its edges. Here's what held up:

- **QB is still the best top-10 investment on average.** Nothing we found changes the overall positional hierarchy. If you need a QB and there's a blue-chip one available, take him without hesitation.

- **Late 1st round is still the sweet spot for OT and EDGE.** The combination of high hit rates and strong second-contract value makes picks 15-32 the most efficient draft real estate for these positions.

- **Day 3 RBs are still fine most of the time.** The cheap FA replacement cost means you can fill the position without premium capital. When the class isn't deep and the prospect isn't blue-chip, this is still the right call.

- **The aggregate advice is right; context determines the exceptions.** The Sharpe ratio is a compass, not a GPS. It points you in the right direction. But the last mile requires scouting, roster analysis, and class evaluation that no single metric can capture.

- **RBs produce immediately on rookie contracts.** This actually reinforces the Sharpe framework in a subtle way: RBs deliver their best production during their cheapest years. The rookie contract surplus is real — but it also means teams get the most value from an RB *before* the second contract the model measures. The model captures this: it's not that RBs are bad, it's that the position's value is front-loaded into the years where every pick is cheap.

---

## Methodology Notes

**Data:** Player stats from Pro Football Reference via `nflreadr`. Contract data from OverTheCap via `nflreadr`. All values normalized to percentage of salary cap at time of signing. Analysis covers 2010-2022 draft classes (minimum 4 NFL seasons for fair evaluation).

**Hit definition (Article 1):** Snap-based quality measure. A "hit" is a player who plays above the positional snap threshold (75% of baseline snaps for their position). This measures whether a player became a starter.

**Blue-chip outcome measure:** For comparing blue-chip categories, we use Article 1's snap-based hit rate as the outcome (independent of career AV, which defines the career blue-chip group). [Career Approximate Value](https://www.pro-football-reference.com/about/approximate_value.htm) (AV) from Pro Football Reference is used for defining career blue chips (top 10% at position) but not as the outcome measure, to avoid circularity.

**Blue-chip classification:** Two definitions used. Retrospective: top 10% of career AV within each position group. Prospective: top 10% of NGS Combined Score within each position group. Players meeting both criteria had the highest elite outcome rates.

**NGS Combined Score:** NFL Next Gen Stats composite prospect score combining athletic testing, college production, and size metrics. Published pre-draft. Historical scores matched to draft records via normalized name and position group.

**Class depth:** A class is "deep" if its count of Good+ prospects (NGS Combined Score >= 70) exceeds the position's median Good+ count across all years. "Thin" if below the median.

**Roster needs:** A position is "filled" if the team had a starter who played above the hit threshold, earned a second contract above the free-agent replacement cost for the position, and was not on an expiring deal. A position is a "need" if it wasn't filled.

**Part 1:** Full methodology and interactive charts at [stranger9977.github.io/draft-sharpe-analysis](https://stranger9977.github.io/draft-sharpe-analysis/)

*Built with R, ggplot2, nflplotR, and gt. Data from nflverse and Pro Football Reference.*
