---
title: "What the Sharpe Ratio Can't See"
subtitle: "Stress-testing our NFL Draft model and finding where it falls short"
author: "Nick Gurol — [X/Twitter](https://x.com/run_the_sims) | [LinkedIn](https://www.linkedin.com/in/nick-gurol-87738a1b7/)"
output:
  html_document:
    toc: true
    toc_float: true
    self_contained: true
knit: (function(input, ...) rmarkdown::render(input, output_dir = dirname(input), ...))
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, warning = FALSE, message = FALSE,
  fig.width = 8, fig.height = 6, dpi = 150,
  fig.path = "charts/v2_",
  out.width = "100%"
)
library(tidyverse)
library(arrow)
library(scales)
library(gt)
library(nflplotR)
library(nflreadr)
library(ggrepel)

pos_colors <- c(
  "QB" = "#E41A1C", "RB" = "#377EB8", "WR" = "#4DAF4A", "TE" = "#984EA3",
  "OT" = "#FF7F00", "IOL" = "#FFFF33", "EDGE" = "#A65628", "DL" = "#F781BF",
  "LB" = "#999999", "CB" = "#66C2A5", "S" = "#FC8D62"
)

tier_order <- c("Top 10", "Late 1st", "Round 2", "Round 3", "Rounds 4-7")
key_positions <- c("QB", "RB", "WR", "TE", "OT", "IOL", "EDGE", "DL", "LB", "CB", "S")
```

```{r load-data}
# Article 1 data
df <- read_parquet("data/draft_sharpe_analysis.parquet")
sharpe <- read_csv("output/sharpe_ratios.csv", show_col_types = FALSE)
player_lookup <- read_csv("output/player_lookup.csv", show_col_types = FALSE)

sharpe <- sharpe %>% mutate(tier = factor(tier, levels = tier_order))
df <- df %>% mutate(tier = factor(tier, levels = tier_order))
player_lookup <- player_lookup %>% mutate(tier = factor(tier, levels = tier_order))
df_eligible <- df %>% filter(season <= 2022)

# V2 data
class_depth <- read_csv("output/v2/class_depth.csv", show_col_types = FALSE)
blue_chip <- read_csv("output/v2/blue_chip.csv", show_col_types = FALSE)
roster_needs <- read_csv("output/v2/roster_needs.csv", show_col_types = FALSE)
team_efficiency <- read_csv("output/v2/team_efficiency_adjusted.csv", show_col_types = FALSE)
r1_picks <- read_csv("output/v2/r1_picks_need_adjusted.csv", show_col_types = FALSE)
market_efficiency <- read_csv("output/v2/market_efficiency.csv", show_col_types = FALSE)
ngs_prospects <- read_csv("output/v2/ngs_prospects_2025.csv", show_col_types = FALSE) %>%
  # NGS lists all defensive backs as "DB" → mapped to CB. Fix known safeties.
  mutate(pos_group = case_when(
    player %in% c("Caleb Downs", "Dillon Thieneman", "Nick Emmanwori", "Malaki Starks") ~ "S",
    TRUE ~ pos_group
  ))
ngs_historical <- read_csv("output/v2/ngs_historical_matched.csv", show_col_types = FALSE)
ngs_class_tiers <- read_csv("output/v2/ngs_class_tiers.csv", show_col_types = FALSE)

# Factor levels
blue_chip <- blue_chip %>% mutate(tier = factor(tier, levels = tier_order))
r1_picks <- r1_picks %>% mutate(tier = factor(tier, levels = tier_order))
```

## TL;DR

In [Part 1](https://stranger9977.github.io/draft-sharpe-analysis/), we built a model that said **never draft an RB in round 1**. That was the takeaway from our initial analysis. But if we break down the key assumptions and approach this from a different angle, does it make sense to draft an RB early? We stress-test three assumptions: that roster context doesn't matter, that all draft classes are equal, and that snap-based quality tells the whole story. The answer: it depends on the team, the year, and the prospect.

---

## The Model Was Right. And Incomplete.

In [Part 1](https://stranger9977.github.io/draft-sharpe-analysis/), we built a quasi-Sharpe ratio for NFL draft picks. The math was clean. The conclusion was clear: **RB shows negative returns across all tiers -- never draft one in round 1.**

That was the takeaway from our initial analysis. But if we break down some key assumptions and approach this from a different angle, does it make sense to draft an RB early?

GMs don't draft on average. They draft in specific years, with specific rosters, facing specific prospects. A model that can't account for context isn't wrong -- it's incomplete. This is Part 2: an honest stress test that defines where Article 1's advice falls short.

The gap between the model's advice and what teams actually do isn't always inefficiency -- sometimes it's intelligence. We tested three assumptions behind Part 1:

```{r assumptions-table, results='asis'}
cat("
| Assumption | What Could Break It |
|---|---|
| Teams should always draft the highest-Sharpe position | Roster needs make some positions irrelevant for some teams |
| All draft classes are interchangeable | Class depth varies wildly year to year |
| Snaps measure quality — and all prospects are fungible | Career value tells a different story, and elite prospects break the averages |
")
```

---

## Quick Recap: What Part 1 Found

```{r recap-sharpe, results='asis'}
rb_top10 <- sharpe %>% filter(pos_group == "RB", tier == "Top 10")
qb_top10 <- sharpe %>% filter(pos_group == "QB", tier == "Top 10")
best_late1 <- sharpe %>% filter(tier == "Late 1st", !is.na(sharpe_elite)) %>%
  slice_max(sharpe_elite, n = 1)

cat(paste0(
  "The quasi-Sharpe ratio combined three things: rookie contract surplus (how much cheaper ",
  "a drafted starter is than a free agent), second-contract value (the market's verdict), ",
  "and elite probability (the chance of a top-decile outcome). We measured risk-adjusted ",
  "returns by position and draft tier from 2010--2022.\n\n",
  "**Key findings:** QB leads all positions in top-10 Sharpe (",
  round(qb_top10$sharpe_elite, 2), "). OT and EDGE peak in the late first round (",
  round(best_late1$sharpe_elite, 2), " for ", best_late1$pos_group,
  "). RB was the worst first-round investment: Sharpe of **",
  round(rb_top10$sharpe_elite, 2), "** in the top 10, negative across every tier.\n\n",
  "The advice was simple: if you want a running back, wait until Day 2 or sign a free agent. ",
  "The data said so.\n\n",
  "But data without context is just math. Let's break down the assumptions and see where they fall short."
))
```

### Weighted Career AV: Our Measuring Stick

This article leans heavily on a single metric, so it's worth understanding exactly what it is and why we chose it.

**[Weighted Career Approximate Value](https://www.pro-football-reference.com/about/approximate_value.htm)** (weighted career AV) is Pro Football Reference's attempt to boil an entire NFL career into one number. It works in two steps:

1. **Season-level AV** assigns each player a value for every season based on playing time, statistical production, and team success. A quarterback who throws 4,500 yards on a playoff team earns more AV than a backup who plays four games on a 4-win team.
2. **Career weighting** sums those seasons but gives extra credit to peak years. The formula weights the best season at 100%, the second-best at 95%, the third at 90%, and so on. A player who had five great years and five mediocre ones scores higher than a player who was merely okay for ten straight.

The result: career AV rewards sustained excellence. Derrick Henry's 86 career AV reflects a decade of elite rushing production. A replacement-level back who hung around for eight years as a committee member might accumulate 25. The gap between those numbers is real -- it maps to Pro Bowls, All-Pro selections, and the kind of production that wins games.

**Why not use Part 1's Sharpe ratio?** Part 1 measured "return" using second-contract value -- what the market paid a player after his rookie deal. That's a reasonable proxy for quality, but it has a blind spot: **the NFL market systematically underpays running backs**. A back who rushes for 1,200 yards a season for four years might walk in free agency or get a modest extension, while a league-average edge rusher commands $15M/year. The Sharpe ratio captures what the *market* values, not what the *player* produced.

Weighted career AV strips out the market layer entirely. When we ran Part 1's Sharpe formula with career AV as the return metric instead of snap-based contract returns, **RB jumped from dead last (negative Sharpe) to 6th out of 11 positions** with a positive Sharpe of 0.249. That gap is the market inefficiency at work: RBs produce more than the market pays them for.

For this article, we need a metric that measures production, not market sentiment. That's career AV.

We define an **elite outcome** as finishing in the **top 10% of weighted career AV at their position** (90th percentile). This is a much stricter bar than Part 1's snap-based "hit" -- it doesn't just ask whether a player became a starter, but whether they became *great*.

---

## Assumption 1: Teams Should Always Draft the Highest-Sharpe Position

Article 1 implicitly assumed that every team should always draft the highest-Sharpe position available. But what if you already have a starter?

The Bills are the perfect test case. From 2014 through 2022, Buffalo always had someone playing quarterback -- EJ Manuel, Kyle Orton, Tyrod Taylor, Matt Barkley, then Josh Allen. A snap-only model would say "QB is filled every year." But having a warm body at quarterback isn't the same as having a good one.

We added a quality gate: a position is only "filled" when the starter meets three criteria -- plays above the snap threshold, earned a second contract above the free-agent replacement cost for the position, and isn't on an expiring deal. A bad QB who plays a lot doesn't count. Tyrod Taylor started 46 games for the Bills and cleared every snap threshold, but he never earned a QB-level second contract -- the market's way of saying he wasn't the long-term answer.

```{r bills-qb-timeline, fig.height=4, fig.width=8}
bills_qb <- roster_needs %>%
  filter(team == "BUF", pos_group == "QB") %>%
  mutate(
    starter_name = case_when(
      pfr_player_id == "ManuEJ00" ~ "EJ Manuel",
      pfr_player_id == "OrtoKy00" ~ "Kyle Orton",
      pfr_player_id == "TaylTy00" ~ "Tyrod Taylor",
      pfr_player_id == "BarkMa00" ~ "Matt Barkley",
      pfr_player_id == "AlleJo02" ~ "Josh Allen",
      TRUE ~ pfr_player_id
    ),
    fill_label = if_else(is_filled, "Filled", "Need"),
    quality_label = case_when(
      starter_is_good ~ "Quality Starter",
      starter_is_hit & !starter_is_good ~ "Playing, Not Good Enough",
      TRUE ~ "No Starter"
    )
  )

ggplot(bills_qb, aes(x = draft_season, y = 1)) +
  geom_segment(aes(xend = draft_season, y = 0.6, yend = 1), color = "gray70", linewidth = 0.8) +
  geom_point(aes(color = is_filled), size = 6) +
  geom_text(aes(label = starter_name), vjust = -1.5, size = 3.5, fontface = "bold") +
  geom_text(aes(label = fill_label), vjust = 2.5, size = 3, color = "gray50") +
  scale_color_manual(values = c("TRUE" = "#2e7d32", "FALSE" = "#c62828"),
                     labels = c("TRUE" = "Position Filled", "FALSE" = "Position Need")) +
  scale_x_continuous(breaks = 2014:2022) +
  coord_cartesian(ylim = c(0.3, 1.7)) +
  labs(
    title = "Buffalo Bills: QB Need vs Filled (2014-2022, Quality Gate)",
    subtitle = "Red = starter lacked quality despite playing | Green = quality starter in place",
    x = NULL, y = NULL, color = NULL
  ) +
  theme_minimal(base_size = 14) +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "top"
  )
```

This is the interesting part. The model says QB has the best top-10 Sharpe. The Bills had a QB on the roster every year. But the quality gate reveals the truth: the model would have been **right** to tell them to draft QB in years where EJ Manuel or Matt Barkley were the answer -- and right to leave the position alone once Josh Allen arrived. The model's general advice ("draft QB") was correct in spirit, even when the roster looked superficially filled.

### How Need-Adjustment Changes the Rankings

In Part 1, we ranked teams by "draft efficiency" -- the average Sharpe ratio of their first-round picks. A team that spent top-10 picks on high-Sharpe positions (QB, OT, EDGE) scored well; a team that spent them on low-Sharpe positions (RB, S) scored poorly. It's a clean metric, but it ignores one obvious thing: whether the team actually needed a player at that position.

The need-adjusted version gives partial credit to teams that drafted low-Sharpe positions to fill genuine roster holes, and penalizes teams that drafted high-Sharpe positions they already had covered. The chart shows every team's original rank vs their need-adjusted rank.

```{r team-efficiency-slope, fig.height=10, fig.width=8}
te_plot <- team_efficiency %>%
  mutate(
    direction = case_when(
      rank_change > 2 ~ "Improved",
      rank_change < -2 ~ "Declined",
      TRUE ~ "Stable"
    )
  )

ggplot(te_plot, aes(y = reorder(team, -rank_adjusted))) +
  geom_segment(
    aes(x = rank_original, xend = rank_adjusted,
        yend = reorder(team, -rank_adjusted),
        color = direction),
    linewidth = 1.2, alpha = 0.7
  ) +
  geom_nfl_logos(aes(x = rank_adjusted, team_abbr = team), width = 0.04) +
  geom_point(aes(x = rank_original), shape = 1, size = 3, color = "gray50") +
  scale_color_manual(
    values = c("Improved" = "#2e7d32", "Declined" = "#c62828", "Stable" = "gray60"),
    name = "After Need Adjustment"
  ) +
  scale_x_reverse(breaks = seq(0, 32, 4)) +
  labs(
    title = "Draft Efficiency: Original Rank vs Need-Adjusted Rank",
    subtitle = "Open circle = original rank | Logo = need-adjusted rank | Left is better",
    x = "Rank (lower = better)", y = NULL
  ) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "top")
```

```{r top-movers, results='asis'}
biggest_up <- team_efficiency %>% slice_max(rank_change, n = 1) %>% slice_head(n = 1)
biggest_down <- team_efficiency %>% slice_min(rank_change, n = 1) %>% slice_head(n = 1)

cat(paste0(
  "**Biggest climber:** ", biggest_up$team, " moved up ", biggest_up$rank_change,
  " spots — they addressed real needs with their first-round picks. ",
  "**Biggest faller:** ", biggest_down$team, " dropped ", abs(biggest_down$rank_change),
  " spots — they kept drafting positions they already had filled."
))
```

### Need-Based Picks vs Filled-Position Picks

Here's the intuition: if a team drafts a position they actually need, the player steps into a real role. If they draft a position that's already filled, the player competes for scraps. Does that show up in the Sharpe numbers?

```{r need-vs-filled-box, fig.height=5, fig.width=8}
r1_context <- r1_picks %>%
  mutate(
    context = if_else(is_need, "Addressed a Need", "Position Already Filled")
  )

ggplot(r1_context, aes(x = sharpe_adjusted, y = context, fill = context)) +
  geom_boxplot(alpha = 0.7, outlier.shape = 21, outlier.alpha = 0.5, width = 0.5) +
  geom_jitter(height = 0.15, alpha = 0.3, size = 2) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
  scale_fill_manual(values = c("Addressed a Need" = "#2196F3", "Position Already Filled" = "#FF9800")) +
  labs(
    title = "Need-Adjusted Sharpe: Picks That Filled Needs vs Filled Positions",
    subtitle = "Dashed line = break-even | Each dot is a first-round pick (2014-2022)",
    x = "Need-Adjusted Sharpe Ratio", y = NULL, fill = NULL
  ) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none")
```

```{r need-vs-filled-stats, results='asis'}
need_stats <- r1_context %>%
  group_by(context) %>%
  summarise(
    mean_adj = mean(sharpe_adjusted, na.rm = TRUE),
    median_adj = median(sharpe_adjusted, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  )
need_row <- need_stats %>% filter(context == "Addressed a Need")
filled_row <- need_stats %>% filter(context == "Position Already Filled")

cat(paste0(
  "Picks addressing real needs average a need-adjusted Sharpe of **",
  round(need_row$mean_adj, 3), "** (n=", need_row$n,
  "), while picks at already-filled positions average **",
  round(filled_row$mean_adj, 3), "** (n=", filled_row$n,
  "). The gap isn't just theoretical — teams that ignore roster context ",
  "systematically underperform their draft capital."
))
```

**The takeaway:** Picks that fill genuine roster needs perform better. It makes sense — a player drafted into opportunity has a clearer path to snaps, production, and a second contract. The Sharpe ratio tells you which positions have the best *average* returns. But if you already have a franchise QB, that average is irrelevant to you. Roster context turns a general truth into specific advice.

---

## Assumption 2: All Draft Classes Are Interchangeable

The Sharpe ratio averages across 13 draft classes (2010-2022). But class quality varies wildly. If a GM could know whether this year's class is loaded or barren *before* the draft, the model's one-size-fits-all advice starts to crack.

We actually can measure this -- at least partially. NFL Next Gen Stats publishes composite prospect scores before the draft, combining athletic testing, college production, and size. These scores exist before a single snap is played in the NFL.

### Prospective Class Quality: What Teams Know Before the Draft

NGS uses four tiers: **Below Average** (50-60), **Average** (60-70), **Good** (70-90), and **Elite** (90+). Instead of looking at mean scores (which compress into a narrow band), we count how many prospects in each class grade as Good or better. A class with five Good+ prospects is fundamentally different from one with two, even if their averages are similar.

We'll start with RB -- the position at the center of the Sharpe debate -- then see how the pattern plays out across every position group.

```{r ngs-rb-depth-big, fig.height=5, fig.width=8}
ngs_rb_tiers <- ngs_class_tiers %>%
  filter(pos_group == "RB", season >= 2010, season <= 2026,
         ngs_tier %in% c("Elite", "Good")) %>%
  mutate(
    ngs_tier = factor(ngs_tier, levels = c("Elite", "Good")),
    draft_year = factor(as.integer(season), levels = 2010:2026)
  )

ggplot(ngs_rb_tiers, aes(x = draft_year, y = n, fill = ngs_tier)) +
  geom_col(width = 0.7, alpha = 0.85) +
  scale_fill_manual(
    values = c("Elite" = "#1565C0", "Good" = "#66BB6A"),
    name = "NGS Tier"
  ) +
  scale_x_discrete(drop = FALSE) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  labs(
    title = "RB Draft Class Depth: Good and Elite Prospects (2010-2026)",
    subtitle = "Only prospects scoring Good (70-90) or Elite (90+) on NGS Combined Score",
    x = NULL, y = "Number of Good+ Prospects"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "top"
  )
```

The variation is striking. Some RB classes are loaded with Good+ talent; others barely register. A GM drafting RB in a deep year is playing a fundamentally different game than one drafting in a thin year -- but the Sharpe ratio treats them identically.

Now the same view across every position:

```{r ngs-all-depth-small, fig.height=12, fig.width=8}
ngs_other_tiers <- ngs_class_tiers %>%
  filter(pos_group %in% key_positions, season >= 2010, season <= 2026,
         ngs_tier %in% c("Elite", "Good")) %>%
  mutate(
    ngs_tier = factor(ngs_tier, levels = c("Elite", "Good")),
    pos_group = factor(pos_group, levels = key_positions),
    draft_year = factor(as.integer(season), levels = 2010:2026)
  )

ggplot(ngs_other_tiers, aes(x = draft_year, y = n, fill = ngs_tier)) +
  geom_col(width = 0.7, alpha = 0.85) +
  scale_fill_manual(
    values = c("Elite" = "#1565C0", "Good" = "#66BB6A"),
    name = "NGS Tier"
  ) +
  scale_x_discrete(drop = FALSE) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  facet_wrap(~pos_group, scales = "free_y", ncol = 3, axes = "all") +
  labs(
    title = "Draft Class Depth Across All Positions (2010-2026)",
    subtitle = "Good (70-90) and Elite (90+) NGS prospects by year | Each position tells a different story",
    x = NULL, y = "Number of Good+ Prospects"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 6),
    legend.position = "top",
    strip.text = element_text(face = "bold", size = 12)
  )
```

```{r ngs-class-narrative, results='asis'}
ngs_rb_class <- class_depth %>%
  filter(pos_group == "RB", !is.na(ngs_n_good_plus))
best_class <- ngs_rb_class %>% slice_max(ngs_n_good_plus, n = 1)
worst_class <- ngs_rb_class %>% slice_min(ngs_n_good_plus, n = 1)

cat(paste0(
  "The **", best_class$season,
  "** RB class had **", best_class$ngs_n_good_plus,
  "** prospects grading Good or better — the deepest pool of pre-draft talent in our sample. ",
  "The **", worst_class$season, "** class had just **", worst_class$ngs_n_good_plus,
  "**. That's the difference between a class where you can find value in any round and one ",
  "where you're chasing a single standout.\n\n",
  "The same pattern plays out at every position. Some years WR is loaded and EDGE is thin. ",
  "Some years the OT class is deep while TE is barren. QB class depth is notoriously spiky -- ",
  "one or two Elite prospects can define the entire class. The Sharpe ratio can't see any of this — ",
  "it averages across all years as if every class is the same. GMs know better."
))
```

```{r class-2025-commentary, results='asis'}
class_2025 <- ngs_class_tiers %>%
  filter(season == 2026, ngs_tier %in% c("Elite", "Good")) %>%
  group_by(pos_group) %>%
  summarise(
    n_good_plus = sum(n),
    n_elite = sum(n[ngs_tier == "Elite"]),
    .groups = "drop"
  ) %>%
  arrange(desc(n_good_plus))

deep_2025 <- class_2025 %>% slice_head(n = 3)
thin_2025 <- class_2025 %>% slice_tail(n = 3)

rb_2025 <- class_2025 %>% filter(pos_group == "RB")
qb_2025 <- class_2025 %>% filter(pos_group == "QB")

cat(paste0(
  "### The 2026 Class at a Glance\n\n",
  "Applying the same framework to the 2026 draft class: **",
  deep_2025$pos_group[1], "** (",  deep_2025$n_good_plus[1], " Good+), **",
  deep_2025$pos_group[2], "** (", deep_2025$n_good_plus[2], "), and **",
  deep_2025$pos_group[3], "** (", deep_2025$n_good_plus[3],
  ") are the deepest classes this year. The thinnest: **RB** (",
  if(nrow(rb_2025) > 0) rb_2025$n_good_plus else "few", " Good+) and **QB** (",
  if(nrow(qb_2025) > 0) qb_2025$n_good_plus else "few", " Good+).\n\n",
  "The RB class stands out as especially thin. Only ",
  if(nrow(rb_2025) > 0) rb_2025$n_good_plus else "a handful of",
  " RB prospects grade as Good or better — and Love is the only one in the Elite tier. ",
  "That 17-point gap to RB2 isn't just a Love story; it's a class story. If you pass on Love, ",
  "the next RB option is a significant downgrade.\n\n",
  "QB is thin too, but for QB the bar is different: Good isn't good enough. ",
  "QB is the one position where only elite prospects (Elite tier) reliably return value in the top 10 — ",
  "a Good QB prospect at pick 5 is a much bigger gamble than a Good OT or EDGE at the same spot. ",
  "For most other positions, Good is deep enough to find value.\n\n",
  "On the other end, **IOL** is loaded — ",
  if(nrow(deep_2025) > 0 && deep_2025$pos_group[1] == "IOL") paste0(deep_2025$n_good_plus[1], " Good+ prospects") else "the deepest position group",
  " give teams plenty of quality options throughout the draft. ",
  "If you need an interior lineman in 2026, the class will come to you.\n\n",
  "This is supply and demand, and the Sharpe ratio doesn't account for it. ",
  "When a class is deep (IOL, CB, WR this year), supply is high — quality players will be available in later rounds, ",
  "so the marginal cost of waiting is low. The Sharpe ratio's advice to 'draft premium positions' works well here ",
  "because you're not giving up much by passing on round 1.\n\n",
  "But when a class is thin (RB, QB this year), supply is scarce. The drop-off from prospect #1 to #2 is steep, ",
  "and waiting means the position becomes dramatically harder to fill. In a loaded class, you can find value at pick 50. ",
  "In a thin class, the value might only exist at pick 10. The Sharpe ratio averages across all classes — ",
  "it can't tell you that *this year's* supply curve looks completely different from last year's.\n\n",
  "That's the blind spot. The model says 'don't draft RB early' based on historical averages. ",
  "But the supply/demand equation changes every year, and in 2026, the RB market is tight."
))
```

### Did the Pre-Draft Signal Match Reality?

The NGS scores tell us what the class looked like before the draft. But did the "deep" classes actually produce more hits? Here's the RB timeline with retrospective hit rates — the percentage of ALL drafted RBs that season who became starters (75%+ of positional snap baseline, from Article 1). Bar color is the pre-draft NGS classification, bar height is what actually happened.

```{r rb-class-timeline-big, fig.height=6, fig.width=8}
rb_classes <- class_depth %>%
  filter(pos_group == "RB")

rb_notable <- tribble(
  ~season, ~label,
  2010, "0 hits",
  2011, "Murray",
  2012, "Martin, Morris",
  2013, "Bell, Lacy",
  2014, "Hill, Hyde",
  2015, "Gordon, Gurley,\nDavid Johnson",
  2016, "Zeke, Henry",
  2017, "CMC, Kamara,\nCook, Hunt,\nMixon, Jones",
  2018, "Saquon, Chubb",
  2019, "Jacobs, Sanders",
  2020, "Taylor, Swift",
  2021, "Harris, Javonte",
  2022, "Walker, Hall"
)

rb_classes <- rb_classes %>%
  left_join(rb_notable, by = "season")

ggplot(rb_classes, aes(x = factor(season), y = hit_rate, fill = ngs_class_quality)) +
  geom_col(width = 0.7, alpha = 0.85) +
  geom_text(
    aes(label = label),
    vjust = -0.3, size = 3, lineheight = 0.85
  ) +
  scale_fill_manual(
    values = c("Deep" = "#2e7d32", "Average" = "#FFA726", "Thin" = "#c62828"),
    na.value = "gray50"
  ) +
  scale_y_continuous(
    labels = percent_format(),
    expand = expansion(mult = c(0, 0.25))
  ) +
  labs(
    title = "RB Draft Class Quality: A Rollercoaster",
    subtitle = "Retrospective hit rate by draft year | Bar color = pre-draft NGS class quality",
    x = NULL, y = "Hit Rate", fill = "Class Quality"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "top"
  )
```

```{r rb-timeline-narrative, results='asis'}
ngs_rb_class <- class_depth %>%
  filter(pos_group == "RB", !is.na(ngs_n_good_plus))
ngs_rb_cor <- cor(ngs_rb_class$ngs_n_good_plus, ngs_rb_class$hit_rate, use = "complete.obs")

cat(paste0(
  "The colors (pre-draft class depth) and heights (what actually happened) track each other. ",
  "The correlation between a class's Good+ prospect count and its eventual hit rate is **",
  round(ngs_rb_cor, 2), "**.\n\n",
  "The mismatches are revealing. 2010 had the highest mean NGS score but produced zero hits — ",
  "athletic profiles alone don't guarantee production. Meanwhile, 2017 was the best actual class ",
  "(23.3% hit rate) because six different backs all found NFL roles.\n\n",
  "The model says \"never draft RB.\" The class data says \"it depends on the year.\" ",
  "And the NGS tier breakdown gives GMs a real window into which kind of year it is — ",
  "before the first pick is made."
))
```

The same signal-vs-reality check across other positions:

```{r other-pos-timeline-small, fig.height=12, fig.width=8}
other_classes <- class_depth %>%
  filter(pos_group %in% key_positions, pos_group != "RB",
         !is.na(hit_rate), !is.na(ngs_class_quality)) %>%
  mutate(draft_year = factor(as.integer(season), levels = sort(unique(as.integer(season)))))

ggplot(other_classes, aes(x = draft_year, y = hit_rate, fill = ngs_class_quality)) +
  geom_col(width = 0.7, alpha = 0.85) +
  scale_fill_manual(
    values = c("Deep" = "#2e7d32", "Average" = "#FFA726", "Thin" = "#c62828"),
    name = "Pre-Draft Class Quality"
  ) +
  scale_y_continuous(labels = percent_format(), expand = expansion(mult = c(0, 0.15))) +
  facet_wrap(~pos_group, scales = "free_y", ncol = 2, axes = "all") +
  scale_x_discrete(drop = FALSE) +
  labs(
    title = "Pre-Draft Signal vs Reality: All Positions",
    subtitle = "Retrospective hit rate by year | Bar color = pre-draft NGS class quality (deep/average/thin)",
    x = NULL, y = "Hit Rate"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 6),
    legend.position = "top",
    strip.text = element_text(face = "bold", size = 12),
    panel.spacing.y = unit(1, "lines")
  )
```

```{r other-timeline-narrative, results='asis'}
# Compute correlation for each position
pos_cors <- class_depth %>%
  filter(pos_group %in% key_positions, !is.na(ngs_n_good_plus), !is.na(hit_rate)) %>%
  group_by(pos_group) %>%
  summarise(
    cor = cor(ngs_n_good_plus, hit_rate, use = "complete.obs"),
    .groups = "drop"
  ) %>%
  arrange(desc(cor))

best_cor <- pos_cors %>% slice_head(n = 1)
worst_cor <- pos_cors %>% slice_tail(n = 1)

cat(paste0(
  "The pre-draft signal isn't equally predictive everywhere. **", best_cor$pos_group,
  "** shows the strongest correlation (r = ", round(best_cor$cor, 2),
  ") between Good+ prospect count and eventual hit rate, while **", worst_cor$pos_group,
  "** shows the weakest (r = ", round(worst_cor$cor, 2),
  "). Positions where athleticism translates more directly — like RB and EDGE — tend to have stronger ",
  "pre-draft signals. Positions where scheme fit and development matter more — like IOL and S — ",
  "show weaker connections between prospect scores and career outcomes."
))
```

---

## Assumption 3: Snaps Measure Quality, and All Prospects Are Fungible

The Sharpe ratio averages over all RB prospects. But we've already seen that not all classes are equal. The same logic applies to individual prospects: the model can't distinguish between an average back and a generational one.

The challenge is identifying elite prospects without hindsight. [Career Approximate Value](https://www.pro-football-reference.com/about/approximate_value.htm) (AV) -- our retrospective measure -- tells us *who turned out* to be elite, but a GM can't use it on draft day. NGS Combined Scores can. They're published before the draft, combining athletic testing, college production, and size into a single number. It's the closest thing to a prospective elite-prospect identifier that exists.

**Definitions for this section:**

- **Pre-draft elite:** Top 10% of NGS Combined Score at the position (knowable before the draft)
- **Career elite:** Top 10% of [career Approximate Value](https://www.pro-football-reference.com/about/approximate_value.htm) at the position (only knowable in hindsight)
- **Hit (outcome measure):** Article 1's snap-based starter threshold — a player who plays above 75% of baseline snaps at their position. This is our outcome metric for comparing elite categories because it's independent of weighted career AV (which defines the career elite group).

### NGS Scores and Career Outcomes

```{r ngs-data-prep}
# Prepare NGS + career outcome dataset (used throughout this section)
ngs_with_outcomes <- blue_chip %>%
  filter(!is.na(cmbn_score), !is.na(is_hit), !is.na(car_av),
         pos_group %in% key_positions, season <= 2022)
```

### Pre-Draft Elite vs Career Elite

We can define "elite" two ways: retrospectively (top 10% of weighted career AV at the position) and prospectively (top 10% of NGS Combined Score at the position). How much do they overlap -- and does it differ by position?

If high NGS scores actually predict career success, the top-scoring prospects should cluster toward the top of the career value axis. Here's every RB with an NGS score — the pre-draft elite prospects (top 10% NGS) are highlighted:

```{r ngs-bc-scatter-rb, fig.height=6, fig.width=8}
rb_scatter <- ngs_with_outcomes %>%
  filter(pos_group == "RB") %>%
  mutate(is_bc = is_ngs_blue_chip)

rb_lm <- lm(car_av ~ cmbn_score, data = rb_scatter)
rb_r2 <- summary(rb_lm)$r.squared

# Label elite prospects + top career AV non-elite prospects
rb_labeled <- rb_scatter %>% filter(is_bc | car_av >= quantile(car_av, 0.9))

ggplot(rb_scatter, aes(x = cmbn_score, y = car_av)) +
  geom_point(data = rb_scatter %>% filter(!is_bc), color = "gray70", alpha = 0.4, size = 2) +
  geom_point(data = rb_scatter %>% filter(is_bc), color = "#377EB8", size = 3.5, alpha = 0.8) +
  geom_text_repel(data = rb_labeled,
            aes(label = pfr_player_name, color = is_bc),
            size = 2.8, fontface = "bold", show.legend = FALSE,
            max.overlaps = 20, seed = 42, box.padding = 0.4,
            min.segment.length = 0.2, segment.color = "gray70") +
  geom_smooth(method = "lm", se = TRUE, color = "#377EB8", linewidth = 0.8, alpha = 0.2) +
  annotate("text",
           x = max(rb_scatter$cmbn_score, na.rm = TRUE) - 2,
           y = max(rb_scatter$car_av, na.rm = TRUE) - 5,
           label = paste0("R² = ", round(rb_r2, 3)),
           color = "#377EB8", size = 4, fontface = "bold", hjust = 1) +
  scale_color_manual(values = c("TRUE" = "#377EB8", "FALSE" = "gray50")) +
  labs(
    title = "RB: Do Pre-Draft Scores Predict Career Value?",
    subtitle = "Blue = pre-draft elite (top 10% NGS) | Gray = everyone else",
    x = "NGS Combined Score (pre-draft)", y = "Weighted Career AV"
  ) +
  theme_minimal(base_size = 14)
```

```{r ngs-bc-rb-narrative, results='asis'}
rb_bc <- rb_scatter %>% filter(is_bc)
rb_nonbc <- rb_scatter %>% filter(!is_bc)

cat(paste0(
  "The pre-draft elite prospects (blue dots) cluster toward the upper right — high NGS scores ",
  "mapping to high career value. Pre-draft elite RBs averaged **",
  round(mean(rb_bc$car_av, na.rm = TRUE), 1),
  "** career AV, vs **", round(mean(rb_nonbc$car_av, na.rm = TRUE), 1),
  "** for everyone else. They became starters at **",
  round(mean(rb_bc$is_hit, na.rm = TRUE) * 100, 1),
  "%** vs **", round(mean(rb_nonbc$is_hit, na.rm = TRUE) * 100, 1),
  "%**.\n\n",
  "Not every pre-draft elite prospect panned out — Jahvid Best and LaMichael James scored elite on NGS but busted. ",
  "But the signal is clear: the top of the NGS distribution produces starters at dramatically higher rates."
))
```

The same view across all positions:

```{r ngs-bc-scatter-all, fig.height=14, fig.width=8}
all_scatter <- ngs_with_outcomes %>%
  filter(pos_group %in% key_positions) %>%
  mutate(
    is_bc = is_ngs_blue_chip,
    pos_group = factor(pos_group, levels = key_positions)
  )

# Label the top elite prospects in each position
bc_labels <- all_scatter %>%
  filter(is_bc) %>%
  group_by(pos_group) %>%
  slice_max(car_av, n = 3) %>%
  ungroup()

# Label interesting gray dots: non-elite prospects with top career AV (the model misses)
gray_labels <- all_scatter %>%
  filter(!is_bc) %>%
  group_by(pos_group) %>%
  slice_max(car_av, n = 2) %>%
  ungroup()

ggplot(all_scatter, aes(x = cmbn_score, y = car_av)) +
  geom_point(data = all_scatter %>% filter(!is_bc), color = "gray70", alpha = 0.3, size = 1.5) +
  geom_point(data = all_scatter %>% filter(is_bc), color = "#377EB8", size = 2.5, alpha = 0.7) +
  geom_text_repel(data = bc_labels, aes(label = pfr_player_name),
            size = 2.2, color = "#377EB8", fontface = "bold",
            max.overlaps = 15, seed = 42, box.padding = 0.3,
            min.segment.length = 0.2, segment.color = "gray70") +
  geom_text_repel(data = gray_labels, aes(label = pfr_player_name),
            size = 2, color = "gray40", fontface = "italic",
            max.overlaps = 10, seed = 42, box.padding = 0.3,
            min.segment.length = 0.2, segment.color = "gray80") +
  geom_smooth(method = "lm", se = FALSE, color = "#377EB8", linewidth = 0.7, alpha = 0.5) +
  facet_wrap(~pos_group, scales = "free", ncol = 3) +
  labs(
    title = "Pre-Draft Score vs Career Value: All Positions",
    subtitle = "Blue = pre-draft elite (top 10% NGS) | Top 3 elite prospects labeled per position",
    x = "NGS Combined Score", y = "Weighted Career AV"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    strip.text = element_text(face = "bold", size = 11)
  )
```

```{r ngs-bc-all-narrative, results='asis'}
bc_all <- all_scatter %>% filter(is_bc)
nonbc_all <- all_scatter %>% filter(!is_bc)

cat(paste0(
  "Across all positions, pre-draft elite prospects became starters at **",
  round(mean(bc_all$is_hit, na.rm = TRUE) * 100, 1),
  "%** vs **", round(mean(nonbc_all$is_hit, na.rm = TRUE) * 100, 1),
  "%** for everyone else. The trend line slopes upward at every position — the pre-draft signal ",
  "isn't perfect, but it meaningfully separates prospects who become starters from those who don't."
))
```

### The Historical NGS Elite

Who were the highest-graded pre-draft prospects at each position, and did they deliver? Here we use the stricter bar: **top 10% of weighted career AV** at their position — not just a starter, but a *great* player. Starting with RB:

```{r ngs-elite-rb-big, fig.height=5, fig.width=8}
# Compute career AV elite threshold per position
av_thresholds <- ngs_with_outcomes %>%
  group_by(pos_group) %>%
  summarise(av_90 = quantile(car_av, 0.9, na.rm = TRUE), .groups = "drop")

ngs_with_av_bc <- ngs_with_outcomes %>%
  left_join(av_thresholds, by = "pos_group") %>%
  mutate(is_career_bc = car_av >= av_90)

top_ngs_rb <- ngs_with_av_bc %>%
  filter(pos_group == "RB") %>%
  arrange(desc(cmbn_score)) %>%
  slice_head(n = 8) %>%
  mutate(
    player_label = paste0(pfr_player_name, " ('", str_sub(season, 3, 4), ")"),
    player_label = fct_reorder(player_label, cmbn_score),
    outcome = if_else(is_career_bc, "Top 10% Weighted Career AV", "Below Threshold")
  )

ggplot(top_ngs_rb, aes(x = cmbn_score, y = player_label, fill = outcome)) +
  geom_col(width = 0.7, alpha = 0.85) +
  geom_text(aes(label = round(cmbn_score, 1)), hjust = -0.15, size = 4, fontface = "bold") +
  scale_fill_manual(values = c("Top 10% Weighted Career AV" = "#1565C0", "Below Threshold" = "#EF5350")) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(
    title = "Top 8 RB Prospects by NGS Combined Score (2010-2022)",
    subtitle = "Pre-draft NGS score vs outcome | Blue = top 10% weighted career AV",
    x = "NGS Combined Score", y = NULL, fill = "Career\nOutcome"
  ) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "top")
```

```{r ngs-elite-rb-narrative, results='asis'}
rb_elite_stats <- top_ngs_rb %>%
  summarise(elite_rate = mean(is_career_bc, na.rm = TRUE), n = n())

cat(paste0(
  "The top 8 RB prospects by pre-draft NGS score finished in the top 10% of career AV at **",
  round(rb_elite_stats$elite_rate * 100, 1),
  "%** — far above the position's overall average. The names are what you'd expect: ",
  "Jonathan Taylor, Saquon Barkley, Derrick Henry, Ezekiel Elliott. These aren't random outcomes — ",
  "the pre-draft signal identified them before a single NFL snap.\n\n",
  "Now the same view across all positions:"
))
```

```{r ngs-elite-all-small, fig.height=14, fig.width=8}
top_ngs_all <- ngs_with_av_bc %>%
  filter(pos_group %in% key_positions) %>%
  group_by(pos_group) %>%
  arrange(desc(cmbn_score)) %>%
  slice_head(n = 6) %>%
  ungroup() %>%
  mutate(
    player_label = paste0(pfr_player_name, " ('", str_sub(season, 3, 4), ")"),
    player_label_unique = paste0(pos_group, "___", player_label),
    player_label_unique = fct_reorder(player_label_unique, cmbn_score),
    outcome = if_else(is_career_bc, "Top 10% Weighted Career AV", "Below Threshold"),
    pos_group = factor(pos_group, levels = key_positions)
  )

ggplot(top_ngs_all, aes(x = cmbn_score, y = player_label_unique, fill = outcome)) +
  geom_col(width = 0.7, alpha = 0.85) +
  geom_text(aes(label = round(cmbn_score, 1)), hjust = -0.15, size = 2.8, fontface = "bold") +
  scale_fill_manual(values = c("Top 10% Weighted Career AV" = "#1565C0", "Below Threshold" = "#EF5350")) +
  scale_y_discrete(labels = function(x) str_remove(x, "^.*___")) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.12))) +
  facet_wrap(~pos_group, scales = "free", ncol = 2) +
  labs(
    title = "Top 6 Historical Players by NGS Combined Score: All Positions",
    subtitle = "Pre-draft NGS score vs career outcome | Blue = top 10% weighted career AV",
    x = "NGS Combined Score", y = NULL, fill = "Career\nOutcome"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "top",
    strip.text = element_text(face = "bold", size = 11)
  )
```

```{r ngs-elite-all-narrative, results='asis'}
elite_stats <- ngs_with_av_bc %>%
  filter(pos_group %in% key_positions) %>%
  group_by(pos_group) %>%
  arrange(desc(cmbn_score)) %>%
  slice_head(n = 6) %>%
  summarise(
    elite_rate = mean(is_career_bc, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  ) %>%
  arrange(desc(elite_rate))

best_elite <- elite_stats %>% slice_head(n = 1)
worst_elite <- elite_stats %>% slice_tail(n = 1)

cat(paste0(
  "**", best_elite$pos_group, "** leads the pack — the top 6 NGS scorers reached top-10% weighted career AV at",
  round(best_elite$elite_rate * 100, 1),
  "%. **", worst_elite$pos_group, "** has the weakest pre-draft-to-career translation (",
  round(worst_elite$elite_rate * 100, 1),
  "%). The model's \"never draft RB\" ",
  "is really \"never draft an *average* RB.\" When the pre-draft signal says a prospect is ",
  "separated from the class, they become *great* at dramatically higher rates."
))
```

---

## The Love Question Revisited

In Part 1, we asked: *Should you draft Jeremiah Love at 10?*

The model said no. RB has a deeply negative Sharpe in the top 10. But we've now shown that the answer depends on things the model can't see: roster needs, class depth, and prospect separation. Love is the case that brings all three together.

### The 2026 RB Class

We showed above that historical RBs with elite pre-draft NGS scores become great at dramatically higher rates than the position average. Where does Love's 2026 class fit in?

```{r ngs-love-bar-big, fig.height=7, fig.width=8}
ngs_rb <- ngs_prospects %>%
  filter(pos_group == "RB", !is.na(cmbn_score)) %>%
  arrange(desc(cmbn_score)) %>%
  mutate(
    is_love = str_detect(player, "(?i)love"),
    ngs_tier = case_when(
      cmbn_score >= 90 ~ "Elite",
      cmbn_score >= 70 ~ "Good",
      cmbn_score >= 60 ~ "Average",
      TRUE ~ "Below Avg"
    ),
    ngs_tier = factor(ngs_tier, levels = c("Elite", "Good", "Average", "Below Avg")),
    player_label = paste0(player, " (", college, ")"),
    player_label = fct_reorder(player_label, cmbn_score)
  )

ggplot(ngs_rb, aes(x = cmbn_score, y = player_label, fill = ngs_tier, color = ngs_tier)) +
  geom_col(width = 0.7, alpha = 0.85, show.legend = TRUE) +
  geom_point(aes(x = cmbn_score + 1.5), size = 8, shape = 21, stroke = 2, show.legend = FALSE) +
  geom_text(aes(x = cmbn_score + 1.5, label = round(cmbn_score, 1)),
            color = "white", size = 2.3, fontface = "bold", show.legend = FALSE) +
  scale_fill_manual(
    values = c("Elite" = "#1565C0", "Good" = "#66BB6A", "Average" = "#FFA726", "Below Avg" = "#EF5350"),
    name = "NGS Tier"
  ) +
  scale_color_manual(
    values = c("Elite" = "#1565C0", "Good" = "#66BB6A", "Average" = "#FFA726", "Below Avg" = "#EF5350"),
    guide = "none"
  ) +
  scale_x_continuous(breaks = seq(55, 95, 5), expand = expansion(mult = c(0, 0.12))) +
  coord_cartesian(xlim = c(55, NA), clip = "off") +
  labs(
    title = "2026 RB Prospects: NGS Combined Score",
    subtitle = "Love's 94.3 is tied for #1 overall (all positions) with Sonny Styles",
    x = "NGS Combined Score", y = NULL
  ) +
  theme_minimal(base_size = 14) +
  theme(panel.grid.major.y = element_blank())
```

```{r ngs-love-narrative, results='asis'}
love_row <- ngs_rb %>% filter(is_love) %>% slice(1)
rb2_row <- ngs_rb %>% filter(!is_love) %>% slice_max(cmbn_score, n = 1)
gap <- love_row$cmbn_score - rb2_row$cmbn_score

cat(paste0(
  "Love's NGS Combined Score of **", round(love_row$cmbn_score, 1),
  "** is tied for #1 overall across all positions. His Production Score of **95.9** is the highest ",
  "of any prospect in the 2026 class. The gap to RB2 (", rb2_row$player, " at ",
  round(rb2_row$cmbn_score, 1), ") is **", round(gap, 1), " points** -- a chasm by NGS standards."
))
```

### Where Love Ranks in the Historical Record

How does Love's 94.3 compare to every RB prospect since 2010? Here's every RB with an NGS Combined Score, with Love highlighted:

```{r love-historical-rank, fig.height=8, fig.width=8}
# Combine historical (2010-2022), recent (2023-2025), and 2026 RB NGS scores
historical_rb_ngs <- ngs_with_outcomes %>%
  filter(pos_group == "RB") %>%
  select(player_name = pfr_player_name, season, cmbn_score) %>%
  mutate(source = "Historical")

# 2023-2025 classes from ngs_historical_matched
recent_rb_ngs <- ngs_historical %>%
  filter(pos_group == "RB", season %in% 2023:2025, !is.na(cmbn_score)) %>%
  select(player_name = player, season, cmbn_score) %>%
  mutate(source = "Recent (2023-2025)")

prospect_rb_ngs <- ngs_prospects %>%
  filter(pos_group == "RB", !is.na(cmbn_score)) %>%
  select(player_name = player, cmbn_score) %>%
  mutate(season = 2026, source = "2026 Prospect")

all_rb_ngs <- bind_rows(historical_rb_ngs, recent_rb_ngs, prospect_rb_ngs) %>%
  arrange(desc(cmbn_score)) %>%
  mutate(
    rank = row_number(),
    is_love = str_detect(player_name, "(?i)love") & source == "2026 Prospect"
  )

love_rank <- all_rb_ngs %>% filter(is_love) %>% pull(rank)
total_rb <- nrow(all_rb_ngs)
love_pctile <- round((1 - love_rank / total_rb) * 100, 1)

# Show top 20 as a horizontal bar, colored by NGS tier
top20 <- all_rb_ngs %>% slice_head(n = 20) %>%
  mutate(
    label = paste0(player_name, " ('", str_sub(season, 3, 4), ")"),
    label = fct_reorder(label, cmbn_score),
    ngs_tier = case_when(
      cmbn_score >= 90 ~ "Elite",
      cmbn_score >= 70 ~ "Good",
      cmbn_score >= 60 ~ "Average",
      TRUE ~ "Below Avg"
    ),
    ngs_tier = factor(ngs_tier, levels = c("Elite", "Good", "Average", "Below Avg"))
  )

ggplot(top20, aes(x = cmbn_score, y = label, fill = ngs_tier, color = ngs_tier)) +
  geom_col(width = 0.7, alpha = 0.85, show.legend = TRUE) +
  geom_point(aes(x = cmbn_score + 0.8, fill = ngs_tier), size = 7, shape = 21, stroke = 1.5, show.legend = FALSE) +
  geom_text(aes(x = cmbn_score + 0.8, label = round(cmbn_score, 1)),
            color = "white", size = 2.2, fontface = "bold", show.legend = FALSE) +
  scale_fill_manual(
    values = c("Elite" = "#1565C0", "Good" = "#66BB6A", "Average" = "#FFA726", "Below Avg" = "#EF5350"),
    name = "NGS Tier"
  ) +
  scale_color_manual(
    values = c("Elite" = "#1565C0", "Good" = "#66BB6A", "Average" = "#FFA726", "Below Avg" = "#EF5350"),
    guide = "none"
  ) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.12))) +
  coord_cartesian(xlim = c(75, NA), clip = "off") +
  labs(
    title = "Top 20 RB Prospects by NGS Score (2010-2026)",
    subtitle = paste0("Love ranks #", love_rank, " out of ", total_rb, " RBs — ", love_pctile, "th percentile"),
    x = "NGS Combined Score", y = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "top", panel.grid.major.y = element_blank())
```

```{r love-historical-narrative, results='asis'}
cat(paste0(
  "Love's 94.3 ranks **#", love_rank, "** among all ", total_rb,
  " RB prospects in the NGS era. That puts him in the **", love_pctile,
  "th percentile** — in the company of Jonathan Taylor, Saquon Barkley, Breece Hall, and Bijan Robinson. ",
  "The pre-draft signal is as strong as it gets.\n\n",
  "But Love isn't the only outlier in 2026. Here's a look at the top prospect at every position:"
))
```

```{r ngs-2025-all-positions, fig.height=14, fig.width=8}
# Top 5 prospects by position in 2026 class
# Manual overrides: include Ty Simpson (QB) and Carnell Tate (WR) in top 5
override_players <- c("Ty Simpson", "Carnell Tate")
top_2025_all <- ngs_prospects %>%
  filter(!is.na(cmbn_score), pos_group %in% key_positions) %>%
  group_by(pos_group) %>%
  arrange(desc(cmbn_score)) %>%
  slice_head(n = 5) %>%
  ungroup()

# Add overrides: replace 5th player in their position if not already present
for(p in override_players) {
  override_row <- ngs_prospects %>% filter(player == p, !is.na(cmbn_score))
  if(nrow(override_row) > 0 && !p %in% top_2025_all$player) {
    pg <- override_row$pos_group[1]
    pos_rows <- top_2025_all %>% filter(pos_group == pg) %>% arrange(desc(cmbn_score))
    top_2025_all <- top_2025_all %>% filter(!(pos_group == pg & player == pos_rows$player[5]))
    top_2025_all <- bind_rows(top_2025_all, override_row)
  }
}

top_2025_all <- top_2025_all %>%
  mutate(
    ngs_tier = case_when(
      cmbn_score >= 90 ~ "Elite",
      cmbn_score >= 70 ~ "Good",
      cmbn_score >= 60 ~ "Average",
      TRUE ~ "Below Avg"
    ),
    ngs_tier = factor(ngs_tier, levels = c("Elite", "Good", "Average", "Below Avg")),
    player_label = paste0(player, " (", college, ")"),
    player_label_unique = paste0(pos_group, "___", player_label),
    player_label_unique = fct_reorder(player_label_unique, cmbn_score),
    pos_group = factor(pos_group, levels = key_positions)
  )

ggplot(top_2025_all, aes(x = cmbn_score, y = player_label_unique, fill = ngs_tier, color = ngs_tier)) +
  geom_col(width = 0.7, alpha = 0.85, show.legend = TRUE) +
  geom_point(aes(x = cmbn_score + 1.5), size = 8, shape = 21, stroke = 2, show.legend = FALSE) +
  geom_text(aes(x = cmbn_score + 1.5, label = round(cmbn_score, 1)),
            color = "white", size = 2.3, fontface = "bold", show.legend = FALSE) +
  scale_fill_manual(
    values = c("Elite" = "#1565C0", "Good" = "#66BB6A", "Average" = "#FFA726", "Below Avg" = "#EF5350"),
    name = "NGS Tier"
  ) +
  scale_color_manual(
    values = c("Elite" = "#1565C0", "Good" = "#66BB6A", "Average" = "#FFA726", "Below Avg" = "#EF5350"),
    guide = "none"
  ) +
  scale_y_discrete(labels = function(x) str_remove(x, "^.*___")) +
  scale_x_continuous(breaks = seq(60, 100, 10), expand = expansion(mult = c(0, 0.12))) +
  coord_cartesian(xlim = c(58, NA), clip = "off") +
  facet_wrap(~pos_group, scales = "free_y", ncol = 2) +
  labs(
    title = "2026 Draft Class: Top 5 Prospects by NGS Combined Score",
    subtitle = "Elite (90+) | Good (70-90) | Average (60-70) | Below Avg (<60)",
    x = "NGS Combined Score", y = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "top",
    strip.text = element_text(face = "bold", size = 11),
    panel.grid.major.y = element_blank(),
    plot.margin = margin(5, 25, 5, 5)
  )
```

```{r ngs-2025-all-narrative, results='asis'}
# Find positions with biggest separation
pos_sep <- ngs_prospects %>%
  filter(!is.na(cmbn_score), pos_group %in% key_positions) %>%
  group_by(pos_group) %>%
  arrange(desc(cmbn_score)) %>%
  summarise(
    top_player = first(player),
    top_score = first(cmbn_score),
    second_score = nth(cmbn_score, 2),
    gap = first(cmbn_score) - nth(cmbn_score, 2),
    .groups = "drop"
  ) %>%
  arrange(desc(gap))

biggest_gap <- pos_sep %>% slice_head(n = 1)
smallest_gap <- pos_sep %>% slice_tail(n = 1)

# Count elite (90+) prospects at premium positions
premium_pos <- c("QB", "EDGE", "OT", "WR")
elite_premium <- ngs_prospects %>%
  filter(!is.na(cmbn_score), pos_group %in% premium_pos, cmbn_score >= 90) %>%
  nrow()

# Count all elite prospects
all_elite <- ngs_prospects %>%
  filter(!is.na(cmbn_score), cmbn_score >= 90)
elite_non_premium <- all_elite %>% filter(!pos_group %in% premium_pos)

elite_note <- if(elite_premium == 0) {
  paste0(
    "\n\nNotably, **no prospect at a premium position (QB, EDGE, OT, WR) cracks the Elite tier (90+) this year**. ",
    "The ", nrow(all_elite), " Elite-graded prospects in 2026 are all at non-premium positions: ",
    paste0(elite_non_premium$player, " (", elite_non_premium$pos_group, ", ", round(elite_non_premium$cmbn_score, 1), ")", collapse = ", "),
    ". That flips the usual script: ",
    "the model says take premium positions early because the talent is richer there, ",
    "but this year the richest talent at the top of the draft is at positions like RB and TE."
  )
} else ""

cat(paste0(
  "The 2026 class has fascinating separation patterns. **", biggest_gap$pos_group,
  "** has the biggest gap between the top prospect (",
  biggest_gap$top_player, " at ", round(biggest_gap$top_score, 1),
  ") and #2 -- a **", round(biggest_gap$gap, 1),
  "-point** chasm. Meanwhile, **", smallest_gap$pos_group,
  "** is the tightest race, with just a **", round(smallest_gap$gap, 1),
  "-point** gap at the top. Where a prospect is clearly separated, the elite-prospect argument ",
  "gets stronger. Where the class is tightly bunched, the model's average-case advice holds.",
  elite_note
))
```

### What Makes an Elite Prospect? A Pre-Draft Model

The scatter plots show a pattern. Can we quantify it? We built a logistic regression predicting whether a drafted player becomes elite (top 10% career AV at their position) using only information available before the draft.

```{r p-elite-model}
# Modeling dataset: 2010-2022 with NGS scores and career outcomes
model_data <- blue_chip %>%
  filter(season >= 2010, season <= 2022,
         !is.na(cmbn_score), !is.na(car_av), !is.na(is_hit),
         pos_group %in% key_positions) %>%
  mutate(
    is_elite = as.integer(is_blue_chip),
    pos_group = factor(pos_group)
  )

# Out-of-time split: train on 2010-2019, test on 2020-2022
train_data <- model_data %>% filter(season <= 2019)
test_data <- model_data %>% filter(season >= 2020)

# Logistic regression: P(elite) from pre-draft features
# Use log(pick) — draft value declines non-linearly
elite_glm <- glm(is_elite ~ cmbn_score + athlete_score + prod_score + log(pick) + pos_group,
                  data = train_data, family = binomial)

elite_summary <- broom::tidy(elite_glm) %>%
  mutate(
    odds_ratio = exp(estimate),
    p_star = case_when(p.value < 0.001 ~ "***", p.value < 0.01 ~ "**",
                       p.value < 0.05 ~ "*", TRUE ~ "")
  )

# Out-of-time test performance
test_data$pred_elite <- predict(elite_glm, newdata = test_data, type = "response")

# AUC via concordance
pos_preds <- test_data$pred_elite[test_data$is_elite == 1]
neg_preds <- test_data$pred_elite[test_data$is_elite == 0]
if (length(pos_preds) > 0 && length(neg_preds) > 0) {
  test_auc <- mean(outer(pos_preds, neg_preds, ">") + 0.5 * outer(pos_preds, neg_preds, "=="))
} else {
  test_auc <- NA
}

# Brier score
test_brier <- mean((test_data$pred_elite - test_data$is_elite)^2)
```

```{r p-elite-table, results='asis'}
elite_summary %>%
  filter(term != "(Intercept)") %>%
  select(term, odds_ratio, p.value, p_star) %>%
  mutate(
    term = str_replace(term, "pos_group", "") %>%
      str_replace("cmbn_score", "NGS Combined Score") %>%
      str_replace("athlete_score", "NGS Athlete Score") %>%
      str_replace("prod_score", "NGS Production Score") %>%
      str_replace("log\\(pick\\)", "Log(Draft Pick)"),
    odds_ratio = round(odds_ratio, 3),
    p.value = format.pval(p.value, digits = 2)
  ) %>%
  gt() %>%
  cols_label(term = "Feature", odds_ratio = "Odds Ratio",
             p.value = "p-value", p_star = "") %>%
  tab_header(title = md("**What Predicts an Elite Career?**"),
             subtitle = paste0("Logistic regression | Train: 2010-2019 | Test AUC: ",
                               round(test_auc, 3), " | Test Brier: ", round(test_brier, 3))) %>%
  tab_options(table.font.size = px(12))
```

```{r p-elite-love}
# Love's features
love_row <- data.frame(
  cmbn_score = 94.3,
  athlete_score = ngs_prospects %>% filter(player == "Jeremiah Love") %>% pull(athlete_score),
  prod_score = ngs_prospects %>% filter(player == "Jeremiah Love") %>% pull(prod_score),
  pick = 10,
  pos_group = factor("RB", levels = levels(model_data$pos_group))
)

love_p_elite <- predict(elite_glm, newdata = love_row, type = "response")

# P(elite) at each pick
love_by_pick <- map_dfr(1:32, function(p) {
  row <- love_row
  row$pick <- p
  tibble(pick = p, p_elite = predict(elite_glm, newdata = row, type = "response"))
})
```

```{r p-elite-love-chart, fig.height=5, fig.width=8}
ggplot(love_by_pick, aes(x = pick, y = p_elite)) +
  geom_line(color = "#377EB8", linewidth = 1.2) +
  geom_point(data = love_by_pick %>% filter(pick == 10),
             color = "#377EB8", size = 4) +
  geom_text(data = love_by_pick %>% filter(pick == 10),
            aes(label = paste0(round(p_elite * 100, 1), "%")),
            vjust = -1.5, color = "#377EB8", fontface = "bold", size = 4) +
  scale_y_continuous(labels = percent_format()) +
  scale_x_continuous(breaks = seq(1, 32, 2)) +
  labs(
    title = "Love's Predicted Probability of an Elite Career",
    subtitle = "Logistic model: NGS scores + log(draft pick) + position | Trained on 2010-2019",
    x = "Draft Pick", y = "P(Elite)"
  ) +
  theme_minimal(base_size = 13)
```

```{r p-elite-narrative, results='asis'}
cmbn_or <- elite_summary %>% filter(term == "cmbn_score") %>% pull(odds_ratio)
cat(paste0(
  "The logistic model — trained on 2010-2019 and validated on the 2020-2022 draft classes ",
  "(AUC: ", round(test_auc, 3), ") — confirms what the scatter plots suggested: NGS Combined Score and draft capital ",
  "are the strongest predictors of elite career outcomes. Each additional point of NGS Combined Score ",
  "multiplies the odds of an elite career by **", round(cmbn_or, 2),
  "x**. Each pick later in the draft reduces them.\n\n",
  "At pick 10, Love's predicted probability of an elite career is **",
  round(love_by_pick$p_elite[love_by_pick$pick == 10] * 100, 1),
  "%**. That's ", round(love_by_pick$p_elite[love_by_pick$pick == 10] / mean(model_data$is_elite), 1),
  "x the base rate across all drafted players with NGS scores."
))
```

### Predicting Career Value: Two Approaches

The logistic model tells us *whether* a player becomes elite. But how much career value can we expect? We trained two models on the same pre-draft features — a linear regression for interpretability and a gradient-boosted model (XGBoost) to capture non-linear effects.

```{r career-av-models}
library(xgboost)

# Linear model — uses log(pick) for non-linear draft value decay
av_lm <- lm(car_av ~ cmbn_score + athlete_score + prod_score + log(pick) + pos_group,
             data = train_data)
av_lm_summary <- broom::tidy(av_lm)

# Linear test performance (out-of-time: 2020-2022)
lm_test_pred <- predict(av_lm, newdata = test_data)
lm_test_rmse <- sqrt(mean((test_data$car_av - lm_test_pred)^2))
lm_test_r2 <- 1 - sum((test_data$car_av - lm_test_pred)^2) /
  sum((test_data$car_av - mean(train_data$car_av))^2)
av_lm_r2 <- summary(av_lm)$r.squared

# XGBoost — use log(pick) in feature matrix
train_xgb <- train_data %>% mutate(log_pick = log(pick))
test_xgb <- test_data %>% mutate(log_pick = log(pick))

xgb_train_features <- model.matrix(
  ~ cmbn_score + athlete_score + prod_score + log_pick + pos_group - 1,
  data = train_xgb)
xgb_test_features <- model.matrix(
  ~ cmbn_score + athlete_score + prod_score + log_pick + pos_group - 1,
  data = test_xgb)
xgb_train_label <- train_data$car_av
xgb_test_label <- test_data$car_av

set.seed(42)
xgb_model <- xgboost(
  data = xgb_train_features,
  label = xgb_train_label,
  nrounds = 200,
  max_depth = 4,
  eta = 0.1,
  subsample = 0.8,
  colsample_bytree = 0.8,
  objective = "reg:squarederror",
  verbose = 0
)

# XGBoost test performance (out-of-time)
xgb_test_pred <- predict(xgb_model, xgb_test_features)
xgb_test_rmse <- sqrt(mean((xgb_test_label - xgb_test_pred)^2))
xgb_test_r2 <- 1 - sum((xgb_test_label - xgb_test_pred)^2) /
  sum((xgb_test_label - mean(xgb_train_label))^2)

# XGBoost train R² for reference
xgb_train_pred <- predict(xgb_model, xgb_train_features)
xgb_train_r2 <- 1 - sum((xgb_train_label - xgb_train_pred)^2) /
  sum((xgb_train_label - mean(xgb_train_label))^2)
```

```{r career-av-comparison, results='asis'}
tibble(
  Model = c("Linear Regression", "XGBoost (tree-based)"),
  `Train R²` = c(round(av_lm_r2, 3), round(xgb_train_r2, 3)),
  `Test R² (2020-22)` = c(round(lm_test_r2, 3), round(xgb_test_r2, 3)),
  `Test RMSE` = c(round(lm_test_rmse, 1), round(xgb_test_rmse, 1))
) %>%
  gt() %>%
  tab_header(title = md("**Predicting Career AV from Pre-Draft Features**"),
             subtitle = "Train: 2010-2019 | Test: 2020-2022 (out-of-time holdout)") %>%
  tab_options(table.font.size = px(12))
```

```{r career-av-importance, fig.height=4, fig.width=8}
importance <- xgb.importance(model = xgb_model, feature_names = colnames(xgb_train_features))

importance_clean <- importance %>%
  mutate(Feature = Feature %>%
           str_replace("pos_group", "Pos: ") %>%
           str_replace("cmbn_score", "NGS Combined") %>%
           str_replace("athlete_score", "NGS Athlete") %>%
           str_replace("prod_score", "NGS Production") %>%
           str_replace("log_pick", "Log(Draft Pick)")) %>%
  slice_head(n = 10)

ggplot(importance_clean, aes(x = Gain, y = fct_reorder(Feature, Gain))) +
  geom_col(fill = "#377EB8", alpha = 0.85) +
  labs(title = "What Drives Career Value? (XGBoost Feature Importance)",
       subtitle = "Top 10 features by information gain",
       x = "Gain", y = NULL) +
  theme_minimal(base_size = 13)
```

```{r career-av-love, fig.height=5, fig.width=8}
love_base <- data.frame(
  cmbn_score = 94.3,
  athlete_score = ngs_prospects %>% filter(player == "Jeremiah Love") %>% pull(athlete_score),
  prod_score = ngs_prospects %>% filter(player == "Jeremiah Love") %>% pull(prod_score),
  pos_group = factor("RB", levels = levels(model_data$pos_group))
)

love_av_by_pick <- map_dfr(1:32, function(p) {
  row <- love_base %>% mutate(pick = p, log_pick = log(p))
  xgb_row <- model.matrix(~ cmbn_score + athlete_score + prod_score + log_pick + pos_group - 1, data = row)
  tibble(
    pick = p,
    lm_pred = predict(av_lm, newdata = row),
    xgb_pred = predict(xgb_model, xgb_row)
  )
})

ggplot(love_av_by_pick, aes(x = pick)) +
  geom_line(aes(y = lm_pred, color = "Linear"), linewidth = 1) +
  geom_line(aes(y = xgb_pred, color = "XGBoost"), linewidth = 1) +
  scale_color_manual(values = c("Linear" = "#999999", "XGBoost" = "#377EB8"), name = "Model") +
  scale_x_continuous(breaks = seq(1, 32, 2)) +
  labs(
    title = "Love's Predicted Career AV by Draft Position",
    subtitle = "Both models: NGS scores + log(pick) + position | Trained 2010-2019",
    x = "Draft Pick", y = "Predicted Weighted Career AV"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "top")
```

```{r career-av-narrative, results='asis'}
love_at_10_xgb <- love_av_by_pick$xgb_pred[love_av_by_pick$pick == 10]
love_at_10_lm <- love_av_by_pick$lm_pred[love_av_by_pick$pick == 10]

cat(paste0(
  "Two models, same story — both trained on 2010-2019 and tested on the 2020-2022 classes they've never seen. ",
  "The linear regression (test R² = ", round(lm_test_r2, 3),
  ") captures the broad strokes: higher NGS scores and earlier picks produce more career value. ",
  "The gradient-boosted model (XGBoost) picks up non-linear effects — the diminishing returns of draft ",
  "capital after the top 10, the interaction between position and athletic profile — and generalizes ",
  if (xgb_test_r2 > lm_test_r2) "better" else "similarly",
  " to unseen data (test RMSE: ", round(xgb_test_rmse, 1), " vs ", round(lm_test_rmse, 1), " for linear).\n\n",
  "Both agree on Love: at pick 10, the linear model projects **", round(love_at_10_lm, 0),
  "** weighted career AV; XGBoost projects **", round(love_at_10_xgb, 0), "**. ",
  "For context, the median elite RB historically produced ",
  round(median(blue_chip$car_av[blue_chip$pos_group == "RB" & blue_chip$is_ngs_blue_chip == TRUE & !is.na(blue_chip$car_av)]), 0),
  " career AV."
))
```

### Did First-Round RBs Actually Outproduce the Alternative?

Article 1 gave us the Sharpe ratio for RB in the top 10: **-0.615**. That's the average across all RBs drafted in the top 10 since 2010. But averages hide a massive range. Some first-round RBs were franchise cornerstones. Others were busts. The question isn't whether RBs are worth it *on average* -- it's how often they outproduce the player a team could have taken instead.

Here's every first-round RB in our data, plotted against the non-RB taken with the very next pick. The **circle** is the RB. The **yellow X** is whoever was drafted immediately after them. Green means the RB produced more career value; red means the next pick was better.

```{r rb-breakeven, fig.height=9, fig.width=8}
# All first-round picks with career AV from blue_chip
all_r1_av <- blue_chip %>%
  filter(pick <= 32, pos_group %in% key_positions, season >= 2010) %>%
  select(pfr_player_name, pos_group, season, pick, car_av, player_sharpe)

# For each RB first rounder, find the very next non-RB drafted
rb_r1 <- all_r1_av %>% filter(pos_group == "RB")

rb_with_alts <- map_dfr(1:nrow(rb_r1), function(i) {
  row <- rb_r1[i, ]
  candidate <- all_r1_av %>%
    filter(season == row$season, pos_group != "RB", pick > row$pick) %>%
    arrange(pick) %>%
    slice_head(n = 1)
  if (nrow(candidate) > 0) {
    row %>% mutate(
      alt_name = candidate$pfr_player_name,
      alt_pos = candidate$pos_group,
      alt_pick = candidate$pick,
      alt_av = candidate$car_av
    )
  } else {
    row %>% mutate(
      alt_name = NA_character_, alt_pos = NA_character_,
      alt_pick = NA_real_, alt_av = NA_real_
    )
  }
})

rb_with_alts <- rb_with_alts %>%
  filter(!is.na(alt_av)) %>%
  mutate(
    beat_alt = car_av > alt_av,
    rb_label = paste0(pfr_player_name, " (#", pick, ", '", str_sub(season, 3, 4), ")"),
    alt_label = paste0(alt_name, " (", alt_pos, ", #", alt_pick, ")"),
    row_label = paste0(rb_label, "  vs  ", alt_label),
    row_label = fct_reorder(row_label, pick)
  )

ggplot(rb_with_alts, aes(y = row_label)) +
  geom_vline(xintercept = 0, color = "gray80", linewidth = 0.5) +
  # Connecting segment
  geom_segment(aes(x = car_av, xend = alt_av,
                   yend = row_label, color = beat_alt),
               linewidth = 1.5, alpha = 0.4) +
  # RB dot
  geom_point(aes(x = car_av, color = beat_alt), size = 4) +
  # Alternative dot
  geom_point(aes(x = alt_av), shape = 4, size = 3, color = "#FF9800", stroke = 1.5) +
  scale_color_manual(values = c("TRUE" = "#2e7d32", "FALSE" = "#c62828"),
                     labels = c("TRUE" = "RB beat the alternative", "FALSE" = "Next pick was better"),
                     name = NULL) +
  labs(
    title = "The RB Gamble: Who They Took\nvs Who Went Next",
    subtitle = "Circle = RB | X = next non-RB pick | Green = RB won",
    x = "Weighted Career AV", y = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "top",
    axis.text.y = element_text(size = 9)
  )
```

```{r rb-breakeven-narrative, results='asis'}
rb_top10_sharpe <- sharpe %>% filter(pos_group == "RB", tier == "Top 10") %>% pull(sharpe_elite)
cmc <- rb_with_alts %>% filter(pfr_player_name == "Christian McCaffrey")
gurley <- rb_with_alts %>% filter(pfr_player_name == "Todd Gurley")
trich <- rb_with_alts %>% filter(pfr_player_name == "Trent Richardson")
rb_beat_pct <- mean(rb_with_alts$beat_alt, na.rm = TRUE)

cat(paste0(
  "Article 1's tier-level Sharpe for RB in the top 10 is **", round(rb_top10_sharpe, 3),
  "** — the worst of any position. The weighted career AV comparison shows why.\n\n",
  "**The wins:** CMC (", round(cmc$car_av), " career AV",
  ") vs ", cmc$alt_name, " (", cmc$alt_pos, ", ",
  round(cmc$alt_av), "). Gurley (",
  round(gurley$car_av), ") vs ", gurley$alt_name, " (", gurley$alt_pos, ", ",
  round(gurley$alt_av), ").\n\n",
  "**The losses:** Trent Richardson at pick 3 (",
  round(trich$car_av), " career AV) — the next pick was **", trich$alt_name,
  "** (", trich$alt_pos, ", ", round(trich$alt_av), " career AV).\n\n",
  "Only **", round(rb_beat_pct * 100), "%** of first-round RBs outproduced the very next non-RB drafted. ",
  "Love needs to be in the CMC/Gurley tier, not the Richardson tier, to justify the pick."
))
```

### The Other 27%: Non-Premium First-Round Picks

The Sharpe ratio analysis tells us what positions teams *should* draft in round one. But what actually happened when teams drafted non-premium positions? Since 2010, a meaningful share of first-round picks went to RB, IOL, DL, LB, CB, S, and TE — the positions the model flags as inefficient. Here's how those picks actually performed.

```{r non-premium-picks}
premium_pos <- c("QB", "EDGE", "OT", "WR")

non_prem_r1 <- blue_chip %>%
  filter(pick <= 32, season >= 2010, season <= 2022,
         !is.na(car_av), pos_group %in% key_positions) %>%
  mutate(is_premium = pos_group %in% premium_pos)

# Weighted avg Sharpe by position (across R1 tiers only)
pos_sharpe <- sharpe %>%
  filter(tier %in% c("Top 10", "Late 1st")) %>%
  group_by(pos_group) %>%
  summarise(avg_sharpe = round(weighted.mean(sharpe_elite, n, na.rm = TRUE), 3), .groups = "drop")

non_prem_summary <- non_prem_r1 %>%
  filter(!is_premium) %>%
  group_by(pos_group) %>%
  summarise(
    n = n(),
    avg_pick = round(mean(pick), 1),
    avg_car_av = round(mean(car_av, na.rm = TRUE), 1),
    hit_rate = round(mean(is_hit, na.rm = TRUE) * 100, 1),
    elite_rate = round(mean(is_blue_chip, na.rm = TRUE) * 100, 1),
    .groups = "drop"
  ) %>%
  left_join(pos_sharpe, by = "pos_group") %>%
  arrange(desc(n))

prem_avg_av <- non_prem_r1 %>% filter(is_premium) %>% pull(car_av) %>% mean(na.rm = TRUE)
nonprem_avg_av <- non_prem_r1 %>% filter(!is_premium) %>% pull(car_av) %>% mean(na.rm = TRUE)
n_nonprem <- sum(!non_prem_r1$is_premium)
n_total <- nrow(non_prem_r1)
```

```{r non-premium-table}
non_prem_summary %>%
  gt() %>%
  cols_label(
    pos_group = "Position", n = "R1 Picks", avg_pick = "Avg Pick",
    avg_car_av = "Avg Career AV", hit_rate = "Hit Rate %",
    elite_rate = "Elite Rate %", avg_sharpe = "Avg Sharpe (R1)"
  ) %>%
  tab_header(
    title = md("**The Other 27%: Non-Premium First-Round Picks (2010-2022)**"),
    subtitle = "Premium = QB, EDGE, OT, WR. How did everyone else do?"
  ) %>%
  tab_style(
    style = cell_fill(color = "#e3f2fd"),
    locations = cells_body(rows = pos_group == "RB")
  ) %>%
  tab_options(table.font.size = px(12))
```

```{r non-premium-narrative, results='asis'}
rb_row <- non_prem_summary %>% filter(pos_group == "RB")
cat(paste0(
  "Since 2010, **", n_nonprem, " of ", n_total, " first-round picks (",
  round(n_nonprem/n_total*100), "%)** went to non-premium positions — the picks ",
  "that Article 1's Sharpe ratio says teams shouldn't make. Those picks averaged **",
  round(nonprem_avg_av, 0), "** career AV, vs **", round(prem_avg_av, 0),
  "** for premium positions.\n\n",
  "RB specifically: **", rb_row$n, " first-round RBs**, averaging pick ",
  rb_row$avg_pick, " and **", rb_row$avg_car_av, "** career AV. Their hit rate (",
  rb_row$hit_rate, "%) trails the premium positions, but their elite rate (",
  rb_row$elite_rate, "%) tells a different story — when first-round RBs hit, they hit big.\n\n",
  "The model says these were all bad picks. The career outcomes say it's more complicated than that."
))
```

### Love's Fair Value in This Draft

The dumbbell chart shows the historical gamble. But what pick makes the gamble worth it? The Sharpe ratio's opportunity cost assumes every class is average, with equal depth at every position. In 2026, that assumption breaks down. **RB supply is scarce** (Love is the only Elite-tier RB, with a 17-point gap to RB2) while **non-RB supply is deep** (IOL, CB, and WR are loaded with Good+ prospects — and even EDGE and OT have enough depth to find value in round 2). When non-RB positions are deep, you can find that EDGE or OT later. When RB is thin, you can't. The effective opportunity cost of drafting Love is lower than the historical average suggests because what you're "giving up" will still be there later.

```{r love-fair-value, fig.height=6, fig.width=8}
# Elite RB career AV (all NGS top-10% RBs)
bc_rb_av <- blue_chip %>%
  filter(pos_group == "RB", is_ngs_blue_chip == TRUE, !is.na(car_av))

bc_median <- median(bc_rb_av$car_av)
bc_q25 <- quantile(bc_rb_av$car_av, 0.25)
bc_q75 <- quantile(bc_rb_av$car_av, 0.75)
bc_min <- min(bc_rb_av$car_av)
bc_max <- max(bc_rb_av$car_av)

# Use blue_chip for career AV (has all drafted players)
all_r1_av <- blue_chip %>%
  filter(pick <= 32, pos_group %in% key_positions, !is.na(car_av), season >= 2010)

# Single non-RB line (all positions combined)
non_rb_line <- all_r1_av %>%
  filter(pos_group != "RB") %>%
  group_by(pick) %>%
  summarise(avg_av = mean(car_av, na.rm = TRUE), n = n(), .groups = "drop") %>%
  arrange(pick)

# 5-pick rolling smooth
smooth_line <- function(df, col = "avg_av") {
  df <- df %>% arrange(pick) %>% filter(pick <= 32)
  vals <- numeric(nrow(df))
  for(i in seq_along(df$pick)) {
    p <- df$pick[i]
    window_picks <- (p - 2):(p + 2)
    window_data <- df %>% filter(pick %in% window_picks)
    vals[i] <- mean(window_data[[col]], na.rm = TRUE)
  }
  df$smooth_av <- vals
  df
}

non_rb_smooth <- smooth_line(non_rb_line)

# Find where non-RB line first drops below elite RB median (the crossover point)
# This marks the end of the dead zone - where non-RB talent no longer outproduces elite RBs
above <- non_rb_smooth$smooth_av > bc_median

# Find the first contiguous stretch where non-RB > median
if(any(above)) {
  dead_zone_start <- non_rb_smooth$pick[which(above)[1]]
  # Find where the first run of TRUE ends
  first_false <- which(!above & non_rb_smooth$pick > dead_zone_start)
  if(length(first_false) > 0) {
    dead_zone_end <- non_rb_smooth$pick[first_false[1] - 1]
  } else {
    dead_zone_end <- non_rb_smooth$pick[max(which(above))]
  }
} else {
  dead_zone_start <- 0
  dead_zone_end <- 0
}

# Fair value = picks AFTER the dead zone
fair_value_start <- dead_zone_end + 1

ggplot() +
  # Dead zone (where non-RB > elite RB median)
  annotate("rect", xmin = dead_zone_start - 0.5, xmax = dead_zone_end + 0.5,
           ymin = -Inf, ymax = Inf, fill = "#c62828", alpha = 0.06) +
  annotate("text", x = (dead_zone_start + dead_zone_end) / 2, y = bc_median + 25,
           label = paste0("Dead zone (", dead_zone_start, "-", dead_zone_end, ")"),
           color = "#c62828", hjust = 0.5, size = 3.5, fontface = "bold") +
  # Fair value zone
  annotate("rect", xmin = dead_zone_end + 0.5, xmax = 32.5,
           ymin = -Inf, ymax = Inf, fill = "#2e7d32", alpha = 0.04) +
  annotate("text", x = (dead_zone_end + 32) / 2, y = bc_median + 25,
           label = paste0("Fair value (", fair_value_start, "+)"),
           color = "#2e7d32", hjust = 0.5, size = 3.5, fontface = "bold") +
  # Non-RB line
  geom_line(data = non_rb_smooth, aes(x = pick, y = smooth_av),
            color = "#c62828", linewidth = 1.2) +
  # Elite RB median
  geom_hline(yintercept = bc_median, color = "#377EB8", linewidth = 1.2, linetype = "solid") +
  # Labels
  annotate("text", x = 32, y = bc_median + 2,
           label = paste0("Elite RB median: ", round(bc_median, 0)),
           color = "#377EB8", hjust = 1, size = 3.2, fontface = "bold") +
  annotate("text", x = 4, y = non_rb_smooth$smooth_av[non_rb_smooth$pick == 4] + 4,
           label = "Non-RB avg", color = "#c62828", hjust = 0.5, size = 3.5, fontface = "bold") +
  scale_x_continuous(breaks = seq(1, 32, 2)) +
  coord_cartesian(ylim = c(max(bc_min - 5, 0), bc_median + 30)) +
  labs(
    title = "Love's Fair Value: Where the Gamble Makes Sense",
    subtitle = "Red = non-RB positions (all) | Blue = Elite RB median",
    x = "Pick Number", y = "Weighted Career AV"
  ) +
  theme_minimal(base_size = 13)
```

```{r love-fair-value-narrative, results='asis'}
cat(paste0(
  "The blue line is the median weighted career AV across all historical elite RBs (top 10% NGS Combined Score): **",
  round(bc_median, 0), "**. The red line is the average career AV of all non-RB first-rounders at each pick, smoothed over a 5-pick rolling window.\n\n",
  "The dead zone is **picks ", dead_zone_start, "-", dead_zone_end,
  "** -- and that number should jump off the page. Part 1 opened with the question: *Should you draft Jeremiah Love at 10?* ",
  "The career AV data just answered it. The crossover point -- where elite RB value starts to exceed the average non-RB pick -- ",
  "is **pick ", fair_value_start, "**. Part 1's question and the career AV breaking point land on the exact same number.\n\n",
  "In the dead zone, the average non-RB pick outproduces even the median elite RB. The opportunity cost is too high. ",
  "**After pick ", dead_zone_end, "**, non-RB talent thins out and the calculus flips.\n\n",
  "Now layer in what we've built:\n\n",
  "- **Roster needs:** If a team already has its premium positions filled, the opportunity cost drops further.\n",
  "- **Class depth:** In 2026, the RB class is thin (Love is the only Elite). Pass on him and there's no comparable RB later. ",
  "Meanwhile, IOL and CB are loaded -- those positions will still be available in round 2+.\n",
  "- **Prospect separation:** Love's NGS profile puts him in the top tier of historical elite RBs, so his expected value is likely *above* the median line, not on it.\n",
  "- **No elite premium prospects:** In a typical year, picks ", dead_zone_start, "-", dead_zone_end,
  " are rich with elite EDGE, OT, and WR talent. In 2026, not a single premium-position prospect grades Elite (90+) on NGS. ",
  "The opportunity cost of taking Love isn't as high as the historical average suggests -- because the alternatives aren't as good as usual.\n\n",
  "Love at 10 isn't just defensible. It might be the pick where the career AV math starts to agree with the gamble."
))
```

### Which Teams Actually Need an RB?

Remember the Bills? We opened this article with a quality gate: Buffalo had a QB on the roster every year, but the gate showed that EJ Manuel and Matt Barkley weren't good enough. Having a body at the position isn't the same as having the position filled.

The same logic applies to running back. RB is a high-turnover position. Backs get hurt, decline fast, and cycle through rosters more than QBs or OTs. A team can enter the offseason with an "RB1" on the depth chart and still have a genuine need if that back is a committee player, a career backup, or an aging veteran without a long-term future.

Here's what every team's RB room looks like heading into the 2026 draft, based on current depth charts.

```{r rb-room-gt-2026}
# Current rosters from ourlads depth charts (primary source)
ourlads <- read_csv("output/v2/ourlads_rb_depth_2026.csv", show_col_types = FALSE)

# 2026 R1 draft order
draft_2026 <- read_csv("output/v2/draft_order_2026_r1.csv", show_col_types = FALSE)

# Teams that need an RB based on current roster quality
# "Need" = no established bellcow, committee situation, or unproven room
need_teams <- c(
  "ARI",  # Allgeier + aging Conner, no long-term answer
  "WAS",  # Rachaad White + Jerome Ford — committee backs, no bellcow
  "NYG",  # Skattebo is a rookie, Tracy unproven — no established starter
  "TEN",  # Pollard aging, Spears hasn't seized the role
  "MIN",  # Aaron Jones is 31, Mason is a rental — no long-term answer
  "CAR",  # Chuba Hubbard is a career backup, Brooks coming off injury
  "JAX",  # Tuten/Allen — completely rebuilt room after Etienne trade, unproven
  "NE",   # RBBC — Stevenson is a committee lead, no difference-maker
  "SEA"   # Lost Kenneth Walker via trade, Charbonnet untested as lead
)

rb_table <- ourlads %>%
  left_join(draft_2026, by = "team") %>%
  mutate(
    status = if_else(team %in% need_teams, "Need", "Filled")
  ) %>%
  arrange(desc(status == "Need"), pick) %>%
  select(team, pick, rb1, rb2, rb3, status)

rb_table %>%
  gt() %>%
  cols_label(
    team = "Team",
    pick = "R1 Pick",
    rb1 = "RB1",
    rb2 = "RB2",
    rb3 = "RB3",
    status = "Status"
  ) %>%
  sub_missing(missing_text = "—") %>%
  tab_style(
    style = cell_fill(color = "#ffebee"),
    locations = cells_body(rows = status == "Need")
  ) %>%
  tab_style(
    style = cell_fill(color = "#e8f5e9"),
    locations = cells_body(rows = status == "Filled")
  ) %>%
  tab_header(
    title = md("**2026 RB Room Snapshot**"),
    subtitle = "Current depth charts via OurLads | Sorted by need then pick"
  ) %>%
  tab_options(
    table.font.size = px(12),
    data_row.padding = px(3)
  )

# Save teams needing RB for downstream
teams_needing_rb_2026 <- rb_table %>% filter(status == "Need") %>% pull(team)
```

### Putting It All Together: Who Should Draft Love?

Everything we've built in this article converges on a single question. Which teams are in the best position to draft Jeremiah Love?

```{r love-verdict, results='asis'}
# Combine: teams that need RB + their draft pick + fair value zone
love_fits <- rb_table %>%
  filter(str_detect(status, "Need"), !is.na(pick)) %>%
  mutate(
    in_fair_zone = pick > dead_zone_end,
    zone_label = if_else(in_fair_zone, "Fair value zone", "Dead zone")
  ) %>%
  arrange(desc(in_fair_zone), pick)

# Best fits: need RB AND in fair value zone
best_fits <- love_fits %>% filter(in_fair_zone)
dead_zone_fits <- love_fits %>% filter(!in_fair_zone)

cat(paste0(
  "We score each first-round team on three factors from this article:\n\n",
  "1. **Roster need:** Does the team lack a starter-level RB? (from the table above)\n",
  "2. **Fair value:** Is their pick outside the dead zone (picks ", dead_zone_start, "-", dead_zone_end,
  ") where premium position talent is richest?\n",
  "3. **Class context:** 2026 is a thin RB class with Love as the only Elite prospect — ",
  "this factor applies equally to every team.\n\n"
))

if(nrow(best_fits) > 0) {
  cat("**Best fits (need RB + fair value pick):**\n\n")
  for(i in 1:nrow(best_fits)) {
    cat(paste0("- **", best_fits$team[i], " (pick ", best_fits$pick[i], "):** ", best_fits$zone_label[i],
               ". RB room: ", best_fits$rb1[i], ", ", best_fits$rb2[i], ", ", best_fits$rb3[i], "\n"))
  }
  cat("\n")
}

if(nrow(dead_zone_fits) > 0) {
  cat("**Need RB but in the dead zone:**\n\n")
  for(i in 1:nrow(dead_zone_fits)) {
    r <- dead_zone_fits[i, ]
    cat(paste0("- **", r$team, " (pick ", r$pick, "):** Premium position talent is too rich here. ",
               "The model says take EDGE/OT/WR and address RB later.\n"))
  }
  cat("\n")
}

# Teams that DON'T need RB
filled_r1 <- rb_table %>%
  filter(!str_detect(status, "Need"), !is.na(pick))
cat(paste0(
  "**Everyone else:** ", nrow(filled_r1), " first-round teams already have a productive RB. ",
  "For them, the Sharpe model's advice stands — draft a premium position.\n\n"
))

cat(paste0(
  "---\n\n",
  "**The bottom line:** The model says the dead zone ends at pick ", dead_zone_end,
  ". Every team above that line pays a premium in opportunity cost that even an elite RB can't offset. ",
  "Below it, the math flips.\n\n",
  if(nrow(best_fits) > 0) paste0(
    "The best fits are **", paste(paste0(best_fits$team, " (", best_fits$pick, ")"), collapse = ", "),
    "** -- teams that need a back *and* pick in the fair value zone. ",
    "But keep an eye on the dead-zone teams too. If Love slides past pick ", dead_zone_end,
    ", a team like **MIN (18)** -- aging Aaron Jones, no long-term answer, picking squarely in the fair value zone -- ",
    "has every reason to trade up. That's the scenario where all three counter-arguments align at once: ",
    "need, class scarcity, and prospect separation. The model built in Part 1 can't see it. This article can."
  ) else
  "In 2026, no team perfectly combines need and fair-value pick position, which means Love could slide -- or a team trades up to get him."
))
```

---

## What We Learned

We built the Sharpe ratio to answer a simple question: which positions are worth drafting in round 1? The answer was clear -- QB, EDGE, OT, and WR. RB was dead last.

Then we stress-tested it. And the model held up -- mostly. The positional hierarchy is real. But we found three blind spots that turn "never draft RB in round 1" from a universal law into a conditional one:

1. **Roster context matters.** The Bills taught us that having a body at QB isn't the same as having Josh Allen. Nine teams enter 2026 without a long-term answer at RB. For them, the opportunity cost of *not* drafting one is higher than the model assumes.

2. **Class depth varies wildly.** The 2026 RB class is thin -- Love is the only Elite prospect. Pass on him and there's no comparable alternative in round 2 or 3. Meanwhile, IOL, CB, and WR are loaded. The positions the model says to take instead will still be available later. The RB won't be.

3. **Not all prospects are average.** Love's 94.3 NGS Combined Score ranks 5th among all RB prospects since 2010. Historical elite RBs with scores like his became great at dramatically higher rates than the position average. The model averages over all RBs. Love isn't average.

And then the number that ties it together: the fair value crossover lands at **pick 10** -- the exact pick from Part 1's opening question. The model's own math says that's where the gamble starts to make sense.

The wildest scenario? Minnesota at pick 18, with Aaron Jones turning 32 and no long-term answer behind him. If Love is still on the board at 10 -- or slides past it -- the Vikings have every reason to trade up. They have the need, the pick falls in the fair value zone, and the 2026 class gives them no alternative. Every counter-argument in this article points at MIN.

---

## What Survived the Stress Test

The counter-arguments didn't destroy the model. They sharpened it. Here's what still holds:

- **QB is still king.** Nothing we found changes the top of the positional hierarchy. If you need a franchise QB and one is available, take him.

- **The dead zone is real.** Picks 1-9 are loaded with non-RB talent. Even in 2026, with a weaker-than-usual premium class, the opportunity cost in the top 10 is steep. Love at 3 is a different conversation than Love at 10.

- **Day 2+ RBs are still the default.** When the class isn't thin and the top RB isn't separated, the model's advice stands. Most years, you can find a productive back in rounds 2-4.

- **The Sharpe ratio is a compass, not a GPS.** It points you in the right direction. But the last mile -- roster needs, class depth, prospect separation -- requires judgment that no single metric can capture. The model tells you what to do on average. GMs don't draft on average. They draft in a specific year, for a specific team, at a specific pick. Context is the difference between a good process and a great one.

---

## Methodology Notes

**Data:** Player stats from Pro Football Reference via `nflreadr`. Contract data from OverTheCap via `nflreadr`. All values normalized to percentage of salary cap at time of signing. Analysis covers 2010-2022 draft classes (minimum 4 NFL seasons for fair evaluation).

**Hit definition (Article 1):** Snap-based quality measure. A "hit" is a player who plays above the positional snap threshold (75% of baseline snaps for their position). This measures whether a player became a starter.

**Elite outcome measure:** For comparing elite categories, we use Article 1's snap-based hit rate as the outcome (independent of career AV, which defines the career elite group). [Weighted Career Approximate Value](https://www.pro-football-reference.com/about/approximate_value.htm) (weighted career AV) from Pro Football Reference — which weights toward peak seasons — is used for defining career elite prospects (top 10% at position) and for individual player comparisons in the dumbbell chart.

**Elite classification:** Two definitions used. Retrospective: top 10% of weighted career AV within each position group. Prospective: top 10% of NGS Combined Score within each position group. Players meeting both criteria had the highest elite outcome rates.

**NGS Combined Score:** NFL Next Gen Stats composite prospect score combining athletic testing, college production, and size metrics. Published pre-draft. Historical scores matched to draft records via normalized name and position group.

**Class depth:** A class is "deep" if its count of Good+ prospects (NGS Combined Score >= 70) exceeds the position's median Good+ count across all years. "Thin" if below the median.

**Roster needs:** A position is "filled" if the team had a starter who played above the hit threshold, earned a second contract above the free-agent replacement cost for the position, and was not on an expiring deal. A position is a "need" if it wasn't filled.

**Part 1:** Full methodology and interactive charts at [stranger9977.github.io/draft-sharpe-analysis](https://stranger9977.github.io/draft-sharpe-analysis/)

*Built with R, ggplot2, nflplotR, and gt. Data from nflverse and Pro Football Reference.*
