---
title: "When to Ignore the Sharpe Ratio"
subtitle: "Stress-testing our NFL Draft model — and finding when it's wrong"
output:
  html_document:
    toc: true
    toc_float: true
    self_contained: true
knit: (function(input, ...) rmarkdown::render(input, output_dir = dirname(input), ...))
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, warning = FALSE, message = FALSE,
  fig.width = 10, fig.height = 7, dpi = 300,
  fig.path = "charts/v2_"
)
library(tidyverse)
library(arrow)
library(scales)
library(gt)
library(nflplotR)
library(nflreadr)

pos_colors <- c(
  "QB" = "#E41A1C", "RB" = "#377EB8", "WR" = "#4DAF4A", "TE" = "#984EA3",
  "OT" = "#FF7F00", "IOL" = "#FFFF33", "EDGE" = "#A65628", "DL" = "#F781BF",
  "LB" = "#999999", "CB" = "#66C2A5", "S" = "#FC8D62"
)

tier_order <- c("Top 10", "Late 1st", "Round 2", "Round 3", "Rounds 4-7")
key_positions <- c("QB", "RB", "WR", "TE", "EDGE", "OT")
```

```{r load-data}
# Article 1 data
df <- read_parquet("data/draft_sharpe_analysis.parquet")
sharpe <- read_csv("output/sharpe_ratios.csv", show_col_types = FALSE)
player_lookup <- read_csv("output/player_lookup.csv", show_col_types = FALSE)

sharpe <- sharpe %>% mutate(tier = factor(tier, levels = tier_order))
df <- df %>% mutate(tier = factor(tier, levels = tier_order))
player_lookup <- player_lookup %>% mutate(tier = factor(tier, levels = tier_order))
df_eligible <- df %>% filter(season <= 2022)

# V2 data
class_depth <- read_csv("output/v2/class_depth.csv", show_col_types = FALSE)
blue_chip <- read_csv("output/v2/blue_chip.csv", show_col_types = FALSE)
roster_needs <- read_csv("output/v2/roster_needs.csv", show_col_types = FALSE)
team_efficiency <- read_csv("output/v2/team_efficiency_adjusted.csv", show_col_types = FALSE)
r1_picks <- read_csv("output/v2/r1_picks_need_adjusted.csv", show_col_types = FALSE)
market_efficiency <- read_csv("output/v2/market_efficiency.csv", show_col_types = FALSE)

# Factor levels
blue_chip <- blue_chip %>% mutate(tier = factor(tier, levels = tier_order))
r1_picks <- r1_picks %>% mutate(tier = factor(tier, levels = tier_order))
```

## The Model Was Right. And Useless.

In [Part 1](https://stranger9977.github.io/draft-sharpe-analysis/), we built a quasi-Sharpe ratio for NFL draft picks. The math was clean. The conclusion was clear: **RB shows negative returns across all tiers -- never draft one in round 1.**

We were wrong. Not about the math -- the averages hold. But averages hide the signal.

This is Part 2: finding when the model's advice should be ignored.

The Sharpe ratio tells you what to do *in general*. GMs don't draft in general. They draft in specific years, with specific rosters, facing specific prospects. A model that can't account for context isn't wrong -- it's incomplete.

We're going to stress-test Part 1 through four arguments:

```{r assumptions-table, results='asis'}
cat("
| # | Assumption We Made | What Could Break It |
|---|---|---|
| 1 | Teams should always draft the highest-Sharpe position | Roster needs make some positions irrelevant |
| 2 | All draft classes are interchangeable | Class depth varies wildly year to year |
| 3 | All prospects at a position are fungible | Blue-chip prospects break the averages |
| 4 | Second contracts measure real value | Market inefficiency could distort the signal |
")
```

Three of these will crack the model open. One will surprise us.

---

## Quick Recap: What Part 1 Found

```{r recap-sharpe, results='asis'}
rb_top10 <- sharpe %>% filter(pos_group == "RB", tier == "Top 10")
qb_top10 <- sharpe %>% filter(pos_group == "QB", tier == "Top 10")
best_late1 <- sharpe %>% filter(tier == "Late 1st", !is.na(sharpe_elite)) %>%
  slice_max(sharpe_elite, n = 1)

cat(paste0(
  "The quasi-Sharpe ratio combined three things: rookie contract surplus (how much cheaper ",
  "a drafted starter is than a free agent), second-contract value (the market's verdict), ",
  "and elite probability (the chance of a top-decile outcome). We measured risk-adjusted ",
  "returns by position and draft tier from 2010--2022.\n\n",
  "**Key findings:** QB leads all positions in top-10 Sharpe (",
  round(qb_top10$sharpe_elite, 2), "). OT and EDGE peak in the late first round (",
  round(best_late1$sharpe_elite, 2), " for ", best_late1$pos_group,
  "). RB was the worst first-round investment: Sharpe of **",
  round(rb_top10$sharpe_elite, 2), "** in the top 10, negative across every tier.\n\n",
  "The advice was simple: if you want a running back, wait until Day 2 or sign a free agent. ",
  "The data said so.\n\n",
  "But data without context is just math."
))
```

---

## Argument 1: Roster Needs — "The Bills Don't Need Another QB"

Article 1 implicitly assumed that every team should always draft the highest-Sharpe position available. But the Bills had a quarterback. From 2014 through 2022, Buffalo's QB room was filled every single year -- EJ Manuel, Kyle Orton, Tyrod Taylor, Matt Barkley, then Josh Allen for three straight seasons.

Telling the Bills to draft a QB because "QB has the best top-10 Sharpe" is like telling a homeowner to buy another house because real estate has good returns. Context matters.

We built a roster-need adjustment: if a team already has a starter at a position (is_filled = TRUE), drafting that position again gets penalized. If the position is a genuine need, the pick gets a bonus. Here's what the Bills' QB situation looked like:

```{r bills-qb-timeline, fig.height=5, fig.width=10}
bills_qb <- roster_needs %>%
  filter(team == "BUF", pos_group == "QB") %>%
  mutate(
    starter_name = case_when(
      pfr_player_id == "ManuEJ00" ~ "EJ Manuel",
      pfr_player_id == "OrtoKy00" ~ "Kyle Orton",
      pfr_player_id == "TaylTy00" ~ "Tyrod Taylor",
      pfr_player_id == "BarkMa00" ~ "Matt Barkley",
      pfr_player_id == "AlleJo02" ~ "Josh Allen",
      TRUE ~ pfr_player_id
    ),
    fill_label = if_else(is_filled, "Filled", "Need")
  )

ggplot(bills_qb, aes(x = draft_season, y = 1)) +
  geom_segment(aes(xend = draft_season, y = 0.6, yend = 1), color = "gray70", linewidth = 0.8) +
  geom_point(aes(color = is_filled), size = 6) +
  geom_text(aes(label = starter_name), vjust = -1.5, size = 3.5, fontface = "bold") +
  geom_text(aes(label = fill_label), vjust = 2.5, size = 3, color = "gray50") +
  scale_color_manual(values = c("TRUE" = "#2e7d32", "FALSE" = "#c62828"),
                     labels = c("TRUE" = "Position Filled", "FALSE" = "Position Need")) +
  scale_x_continuous(breaks = 2014:2022) +
  coord_cartesian(ylim = c(0.3, 1.7)) +
  labs(
    title = "Buffalo Bills: QB Was Filled Every Year (2014-2022)",
    subtitle = "Our model would have told the Bills to draft QB every year — even with Josh Allen on the roster",
    x = NULL, y = NULL, color = NULL
  ) +
  theme_minimal(base_size = 14) +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "top"
  )
```

QB was filled all nine years. The model's "always draft QB" advice would have been actively harmful for Buffalo.

### How Need-Adjustment Changes the Rankings

When we account for roster context, some teams look much smarter -- and others look much worse. The chart below shows every team's original draft efficiency rank vs their need-adjusted rank. Teams that drafted to fill needs climb; teams that doubled down on already-strong positions fall.

```{r team-efficiency-slope, fig.height=9, fig.width=10}
te_plot <- team_efficiency %>%
  mutate(
    direction = case_when(
      rank_change > 2 ~ "Improved",
      rank_change < -2 ~ "Declined",
      TRUE ~ "Stable"
    )
  )

ggplot(te_plot, aes(y = reorder(team, -rank_adjusted))) +
  geom_segment(
    aes(x = rank_original, xend = rank_adjusted,
        yend = reorder(team, -rank_adjusted),
        color = direction),
    linewidth = 1.2, alpha = 0.7
  ) +
  geom_nfl_logos(aes(x = rank_adjusted, team_abbr = team), width = 0.04) +
  geom_point(aes(x = rank_original), shape = 1, size = 3, color = "gray50") +
  scale_color_manual(
    values = c("Improved" = "#2e7d32", "Declined" = "#c62828", "Stable" = "gray60"),
    name = "After Need Adjustment"
  ) +
  scale_x_reverse(breaks = seq(0, 32, 4)) +
  labs(
    title = "Draft Efficiency: Original Rank vs Need-Adjusted Rank",
    subtitle = "Open circle = original rank | Logo = need-adjusted rank | Left is better",
    x = "Rank (lower = better)", y = NULL
  ) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "top")
```

```{r top-movers, results='asis'}
biggest_up <- team_efficiency %>% slice_max(rank_change, n = 1)
biggest_down <- team_efficiency %>% slice_min(rank_change, n = 1)

cat(paste0(
  "**Biggest climber:** ", biggest_up$team, " moved up ", biggest_up$rank_change,
  " spots — they addressed real needs with their first-round picks. ",
  "**Biggest faller:** ", biggest_down$team, " dropped ", abs(biggest_down$rank_change),
  " spots — they kept drafting positions they already had filled."
))
```

### Need-Based Picks vs Filled-Position Picks

Do picks that address roster needs actually perform differently? Here's the raw Sharpe distribution for first-round picks that filled a need vs picks at positions the team already had covered:

```{r need-vs-filled-box, fig.height=6, fig.width=10}
r1_context <- r1_picks %>%
  mutate(
    context = if_else(is_need, "Addressed a Need", "Position Already Filled")
  )

ggplot(r1_context, aes(x = sharpe_adjusted, y = context, fill = context)) +
  geom_boxplot(alpha = 0.7, outlier.shape = 21, outlier.alpha = 0.5, width = 0.5) +
  geom_jitter(height = 0.15, alpha = 0.3, size = 2) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
  scale_fill_manual(values = c("Addressed a Need" = "#2196F3", "Position Already Filled" = "#FF9800")) +
  labs(
    title = "Need-Adjusted Sharpe: Picks That Filled Needs vs Filled Positions",
    subtitle = "Dashed line = break-even | Each dot is a first-round pick (2014-2022)",
    x = "Need-Adjusted Sharpe Ratio", y = NULL, fill = NULL
  ) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none")
```

```{r need-vs-filled-stats, results='asis'}
need_stats <- r1_context %>%
  group_by(context) %>%
  summarise(
    mean_adj = mean(sharpe_adjusted, na.rm = TRUE),
    median_adj = median(sharpe_adjusted, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  )
need_row <- need_stats %>% filter(context == "Addressed a Need")
filled_row <- need_stats %>% filter(context == "Position Already Filled")

cat(paste0(
  "Picks addressing real needs average a need-adjusted Sharpe of **",
  round(need_row$mean_adj, 3), "** (n=", need_row$n,
  "), while picks at already-filled positions average **",
  round(filled_row$mean_adj, 3), "** (n=", filled_row$n,
  "). The gap isn't just theoretical — teams that ignore roster context ",
  "systematically underperform their draft capital."
))
```

**The takeaway:** The Sharpe ratio tells you which positions have the best *average* returns. But if you already have a franchise QB, that average is irrelevant to you. Roster context turns a general truth into specific advice.

---

## Argument 2: Draft Class Depth — "Not All RB Classes Are Created Equal"

The Sharpe ratio averages across 13 draft classes (2010-2022). But class quality varies wildly. The 2017 RB class was loaded -- McCaffrey, Kamara, Cook, Hunt, Mixon, Aaron Jones. The 2010 class produced **zero hits**. Telling a GM in 2017 and 2010 "don't draft RB" is giving the same advice for very different situations.

### The Hit Rate Heatmap

```{r class-depth-heatmap, fig.height=8, fig.width=12}
heatmap_data <- class_depth %>%
  filter(pos_group %in% key_positions) %>%
  mutate(pos_group = factor(pos_group, levels = rev(key_positions)))

ggplot(heatmap_data, aes(x = factor(season), y = pos_group, fill = hit_rate)) +
  geom_tile(color = "white", linewidth = 0.5) +
  geom_text(
    aes(label = paste0(round(hit_rate * 100), "%")),
    size = 3, color = "white", fontface = "bold"
  ) +
  scale_fill_gradient2(
    low = "#c62828", mid = "#f9a825", high = "#2e7d32",
    midpoint = 0.15, limits = c(0, max(heatmap_data$hit_rate, na.rm = TRUE)),
    labels = percent_format()
  ) +
  labs(
    title = "Draft Class Hit Rates Vary Wildly by Year and Position",
    subtitle = "Hit rate = % of drafted players who became starters | Key positions, 2010-2022",
    x = NULL, y = NULL, fill = "Hit Rate"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank()
  )
```

The variance is the story. Some position-year combinations are bright green (loaded classes), others deep red (barren). A model that averages across all of them misses the point entirely.

### Deep vs Thin Classes: Same Position, Different Returns

```{r deep-vs-thin-bar, fig.height=6, fig.width=10}
depth_comparison <- class_depth %>%
  filter(pos_group %in% key_positions, !is.na(class_quality)) %>%
  group_by(pos_group, class_quality) %>%
  summarise(
    mean_return = mean(mean_player_return, na.rm = TRUE),
    mean_hit_rate = mean(hit_rate, na.rm = TRUE),
    n_classes = n(),
    .groups = "drop"
  ) %>%
  mutate(pos_group = factor(pos_group, levels = key_positions))

ggplot(depth_comparison, aes(x = pos_group, y = mean_hit_rate, fill = class_quality)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6, alpha = 0.85) +
  geom_text(
    aes(label = paste0(round(mean_hit_rate * 100, 1), "%")),
    position = position_dodge(width = 0.7), vjust = -0.5, size = 3.5
  ) +
  scale_fill_manual(values = c("Deep" = "#2e7d32", "Thin" = "#c62828")) +
  scale_y_continuous(labels = percent_format(), expand = expansion(mult = c(0, 0.15))) +
  labs(
    title = "Deep Draft Classes Produce Hits at Much Higher Rates",
    subtitle = "Average hit rate in 'Deep' vs 'Thin' class years | 2010-2022",
    x = NULL, y = "Average Hit Rate", fill = "Class Quality"
  ) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "top")
```

```{r deep-thin-rb-gap, results='asis'}
rb_depth <- depth_comparison %>% filter(pos_group == "RB")
rb_deep <- rb_depth %>% filter(class_quality == "Deep")
rb_thin <- rb_depth %>% filter(class_quality == "Thin")

cat(paste0(
  "For RBs, deep classes produce hits at **", round(rb_deep$mean_hit_rate * 100, 1),
  "%** vs **", round(rb_thin$mean_hit_rate * 100, 1),
  "%** in thin classes. That's not a marginal difference -- it's the difference between ",
  "\"this position is draftable\" and \"stay away.\""
))
```

### The RB Class Timeline

```{r rb-class-timeline, fig.height=6, fig.width=10}
rb_classes <- class_depth %>%
  filter(pos_group == "RB")

# Notable players per year for annotations
rb_notable <- tribble(
  ~season, ~label,
  2010, "0 hits",
  2011, "Murray",
  2012, "Martin, Morris",
  2013, "Bell, Lacy",
  2014, "Hill, Hyde",
  2015, "Gordon, Gurley,\nDavid Johnson",
  2016, "Zeke, Henry",
  2017, "CMC, Kamara,\nCook, Hunt,\nMixon, Jones",
  2018, "Saquon, Chubb",
  2019, "Jacobs, Sanders",
  2020, "Taylor, Swift",
  2021, "Harris, Javonte",
  2022, "Walker, Hall"
)

rb_classes <- rb_classes %>%
  left_join(rb_notable, by = "season")

ggplot(rb_classes, aes(x = factor(season), y = hit_rate, fill = class_quality)) +
  geom_col(width = 0.7, alpha = 0.85) +
  geom_text(
    aes(label = label),
    vjust = -0.3, size = 2.8, lineheight = 0.85
  ) +
  scale_fill_manual(
    values = c("Deep" = "#2e7d32", "Thin" = "#c62828"),
    na.value = "gray50"
  ) +
  scale_y_continuous(
    labels = percent_format(),
    expand = expansion(mult = c(0, 0.25))
  ) +
  labs(
    title = "RB Draft Class Quality: A Rollercoaster",
    subtitle = "Hit rate by draft year | Annotated with notable players from each class",
    x = NULL, y = "Hit Rate", fill = "Class Quality"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "top"
  )
```

```{r rb-timeline-narrative, results='asis'}
cat(paste0(
  "2017 stands out: a **23.3% hit rate** -- the highest for RBs in our dataset. ",
  "McCaffrey, Kamara, Cook, Hunt, Mixon, and Aaron Jones all came from the same class. ",
  "Contrast that with 2010 (**0%**) or 2016 and 2018 -- thin classes where Ezekiel Elliott ",
  "and Saquon Barkley were drafted in the top 5. Zeke and Saquon were high picks *because* ",
  "they were the only standout RBs in weak classes, not because the position was a good bet that year.\n\n",
  "The model says \"never draft RB.\" The data says \"it depends on the year.\""
))
```

---

## Argument 3: Blue-Chip Prospects — "When the Talent Is Undeniable"

The Sharpe ratio averages over all RB prospects. But not all prospects are created equal. Blue-chip prospects -- defined here as players who end up in the top 10% of career value at their position -- are a different animal entirely.

```{r blue-chip-rates, results='asis'}
bc_summary <- blue_chip %>%
  filter(!is.na(is_hit)) %>%
  group_by(pos_group) %>%
  summarise(
    all_hit = mean(is_hit),
    blue_hit = mean(is_hit[is_blue_chip == TRUE]),
    n_blue = sum(is_blue_chip == TRUE),
    gap = mean(is_hit[is_blue_chip == TRUE]) - mean(is_hit),
    .groups = "drop"
  ) %>%
  arrange(desc(gap))

rb_bc <- bc_summary %>% filter(pos_group == "RB")

cat(paste0(
  "Blue-chip RBs hit at **", round(rb_bc$blue_hit * 100, 1),
  "%** -- compared to **", round(rb_bc$all_hit * 100, 1),
  "%** for all RBs. That's a ", round(rb_bc$gap * 100, 1),
  " percentage point gap, the largest of any position. ",
  "The question isn't \"should you draft an RB at 10\" but \"should you draft *this* RB at 10.\""
))
```

### Blue-Chip vs Overall Hit Rates

```{r blue-chip-bar, fig.height=7, fig.width=10}
bc_plot <- bc_summary %>%
  pivot_longer(cols = c(all_hit, blue_hit), names_to = "type", values_to = "hit_rate") %>%
  mutate(
    type = if_else(type == "blue_hit", "Blue Chip (Top 10%)", "All Players"),
    pos_group = fct_reorder(pos_group, hit_rate, .fun = max, .desc = TRUE)
  )

ggplot(bc_plot, aes(x = pos_group, y = hit_rate, fill = type)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6, alpha = 0.85) +
  geom_text(
    aes(label = paste0(round(hit_rate * 100, 1), "%")),
    position = position_dodge(width = 0.7), vjust = -0.5, size = 3.2
  ) +
  scale_fill_manual(values = c("Blue Chip (Top 10%)" = "#2e7d32", "All Players" = "#c62828")) +
  scale_y_continuous(labels = percent_format(), expand = expansion(mult = c(0, 0.15))) +
  labs(
    title = "Blue-Chip Prospects Break the Averages",
    subtitle = "Hit rate for top-10% career value players vs all drafted players | 2010-2022",
    x = NULL, y = "Hit Rate", fill = NULL
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "top",
    axis.text.x = element_text(angle = 30, hjust = 1)
  )
```

This is the money chart. Every position benefits from blue-chip talent, but the gap is biggest at RB. The model's "never draft RB" is really "never draft an *average* RB" -- a very different statement.

### Blue-Chip Player Returns by Draft Position

```{r blue-chip-loess, fig.height=7, fig.width=10}
bc_loess <- blue_chip %>%
  filter(
    is_blue_chip == TRUE,
    !is.na(player_return),
    pos_group %in% key_positions
  )

ggplot(bc_loess, aes(x = pick, y = player_return, color = pos_group)) +
  geom_point(alpha = 0.5, size = 2.5) +
  geom_smooth(method = "loess", se = FALSE, linewidth = 1.3, span = 0.9) +
  scale_color_manual(values = pos_colors) +
  scale_x_continuous(breaks = seq(0, 260, 30)) +
  scale_y_continuous(labels = percent_format()) +
  labs(
    title = "Blue-Chip Player Returns by Draft Position",
    subtitle = "LOESS smoothed curves | Only top-10% career value players at each position",
    x = "Overall Pick", y = "Player Return (Cap %)", color = "Position"
  ) +
  theme_minimal(base_size = 14)
```

```{r blue-chip-loess-note, results='asis'}
cat(paste0(
  "When you isolate blue-chip players, the picture changes. Blue-chip RBs drafted early ",
  "produce returns that compete with other positions -- the problem is that *most* RBs aren't ",
  "blue chips. The challenge for GMs isn't the position; it's prospect identification."
))
```

### The Decision Matrix: Class Depth x Prospect Quality

```{r decision-matrix, fig.height=8, fig.width=10}
# Build a decision matrix for R1 picks using class quality and blue chip status
r1_decision <- blue_chip %>%
  filter(
    !is.na(player_return),
    !is.na(class_quality),
    round == 1
  ) %>%
  left_join(
    class_depth %>%
      select(season, pos_group, hit_rate) %>%
      rename(class_hit_rate_direct = hit_rate),
    by = c("season", "pos_group")
  ) %>%
  mutate(
    quadrant = case_when(
      is_blue_chip & class_quality == "Thin" ~ "Override with Caution\n(Blue chip, thin class)",
      is_blue_chip & class_quality == "Deep" ~ "Take the Best Player\n(Blue chip, deep class)",
      !is_blue_chip & class_quality == "Deep" ~ "Follow the Model\n(Regular prospect, deep class)",
      !is_blue_chip & class_quality == "Thin" ~ "Follow the Model\n(Regular prospect, thin class)",
    ),
    outcome = if_else(is_hit, "Hit", "Miss")
  )

ggplot(r1_decision, aes(x = class_hit_rate_direct, y = car_av_pctl, color = outcome)) +
  geom_point(alpha = 0.6, size = 3) +
  geom_hline(yintercept = 0.9, linetype = "dashed", color = "gray40", alpha = 0.7) +
  geom_vline(
    xintercept = median(class_depth$hit_rate[class_depth$pos_group %in% key_positions], na.rm = TRUE),
    linetype = "dashed", color = "gray40", alpha = 0.7
  ) +
  scale_color_manual(values = c("Hit" = "#2e7d32", "Miss" = "#c62828")) +
  scale_x_continuous(labels = percent_format()) +
  annotate("text", x = 0.02, y = 0.95, label = "Override with\nCaution", fontface = "italic",
           color = "gray40", hjust = 0, size = 3.5) +
  annotate("text", x = 0.30, y = 0.95, label = "Take the\nBest Player", fontface = "italic",
           color = "gray40", hjust = 1, size = 3.5) +
  annotate("text", x = 0.02, y = 0.05, label = "Follow\nthe Model", fontface = "italic",
           color = "gray40", hjust = 0, size = 3.5) +
  annotate("text", x = 0.30, y = 0.05, label = "Follow\nthe Model", fontface = "italic",
           color = "gray40", hjust = 1, size = 3.5) +
  labs(
    title = "The Decision Matrix: When to Override the Sharpe Ratio",
    subtitle = "Each dot = a first-round pick | Dashed lines = blue-chip threshold (y) and median class depth (x)",
    x = "Class Hit Rate (depth of that year's class)", y = "Career AV Percentile (prospect quality)",
    color = "Outcome"
  ) +
  theme_minimal(base_size = 14)
```

The upper-left quadrant is where the model is most likely wrong: blue-chip prospects in thin classes. These are the Zekes and Saquons -- players drafted high because they're the clear best at their position in a weak class. The hit rate among blue chips (regardless of class depth) tells us the talent can overcome the positional headwinds.

---

## The Love Question Revisited

In Part 1, we asked: *Should you draft Jeremiah Love at 10?*

The model said no. RB has a deeply negative Sharpe in the top 10. But we've now shown that the answer depends on three things the model ignores:

<div style="background-color: #f0f4f8; border-left: 4px solid #377EB8; padding: 15px 20px; margin: 20px 0; border-radius: 4px;">

**1. Does your team need an RB?** If you already have a starter, the Sharpe ratio's general advice applies -- don't draft for surplus. But if you're running out James Conner on a one-year deal, the calculus changes.

**2. Is Love a blue chip?** Early 2025 scouting consensus has Love as the clear RB1 with significant separation from RB2. If he's truly a top-decile talent, his expected hit rate is closer to `r round(bc_summary %>% filter(pos_group == "RB") %>% pull(blue_hit) * 100, 1)`% than the `r round(bc_summary %>% filter(pos_group == "RB") %>% pull(all_hit) * 100, 1)`% the model uses.

**3. Is the 2025 class thin?** If Love is the only blue-chip RB and the drop-off to RB2 is steep, waiting for Day 2 means accepting a non-blue-chip prospect -- and those hit at the base rate the model correctly identifies as poor.

</div>

If all three answers are yes -- and the early signals suggest they are -- then the Sharpe ratio's advice to avoid RB **doesn't apply**. The model is averaging over situations that don't look like this one.

The irony: Article 1's math is what *proves* Article 1's conclusion is too simple. The same framework that shows RB is bad *on average* also shows that blue-chip RBs in thin classes are the exception.

---

## Argument 4: Market Inefficiency — "Actually, the Market Is Mostly Right"

This was supposed to be our strongest critique. Article 1 used second-contract value as a key component of the Sharpe ratio. If the market overvalues some positions and undervalues others, the whole framework falls apart -- maybe RBs just get underpaid relative to their production.

We tested this by comparing second-contract APY to post-contract performance (average snap percentage after the second deal). If the market is efficient, players who got paid more should perform better afterward. If it's not, we'd see weak correlations.

The data surprised us.

```{r market-scatter, fig.height=9, fig.width=10}
market_plot <- market_efficiency %>%
  filter(!is.na(post_contract_avg_snap), !is.na(second_apy_cap_pct),
         pos_group %in% key_positions) %>%
  mutate(pos_group = factor(pos_group, levels = key_positions))

# Compute correlations for labels
cors <- market_plot %>%
  group_by(pos_group) %>%
  summarise(
    cor = cor(second_apy_cap_pct, post_contract_avg_snap),
    n = n(),
    .groups = "drop"
  ) %>%
  mutate(label = paste0("r = ", round(cor, 2), " (n=", n, ")"))

ggplot(market_plot, aes(x = second_apy_cap_pct, y = post_contract_avg_snap)) +
  geom_point(alpha = 0.3, size = 1.5, color = "gray40") +
  geom_smooth(method = "lm", se = TRUE, color = "#377EB8", linewidth = 1) +
  geom_text(
    data = cors,
    aes(x = Inf, y = Inf, label = label),
    hjust = 1.1, vjust = 1.5, size = 3.5, fontface = "bold", color = "#377EB8",
    inherit.aes = FALSE
  ) +
  facet_wrap(~pos_group, scales = "free", ncol = 3) +
  scale_x_continuous(labels = percent_format()) +
  scale_y_continuous(labels = percent_format()) +
  labs(
    title = "Second Contract Value vs Post-Contract Performance",
    subtitle = "Does the market price players correctly? Higher correlation = more efficient market",
    x = "Second Contract APY (% of Cap)", y = "Post-Contract Avg Snap %"
  ) +
  theme_minimal(base_size = 14) +
  theme(strip.text = element_text(face = "bold", size = 12))
```

The correlations range from moderate to strong across all positions. The market isn't perfect, but it's not randomly pricing players either.

### Market Efficiency by Position

```{r market-correlation-bar, fig.height=6, fig.width=10}
all_cors <- market_efficiency %>%
  filter(!is.na(post_contract_avg_snap), !is.na(second_apy_cap_pct)) %>%
  group_by(pos_group) %>%
  summarise(
    cor = cor(second_apy_cap_pct, post_contract_avg_snap),
    n = n(),
    .groups = "drop"
  ) %>%
  mutate(pos_group = fct_reorder(pos_group, cor))

ggplot(all_cors, aes(x = cor, y = pos_group, fill = pos_group)) +
  geom_col(width = 0.6, alpha = 0.85) +
  geom_text(aes(label = round(cor, 3)), hjust = -0.15, size = 3.8, fontface = "bold") +
  scale_fill_manual(values = pos_colors) +
  scale_x_continuous(limits = c(0, 1), expand = expansion(mult = c(0, 0.1))) +
  labs(
    title = "Market Efficiency: Correlation Between Contract Value and Performance",
    subtitle = "Higher = market prices the position more accurately | Second contract APY vs post-contract snap %",
    x = "Correlation (r)", y = NULL
  ) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none")
```

```{r market-narrative, results='asis'}
qb_cor <- all_cors %>% filter(pos_group == "QB") %>% pull(cor)
dl_cor <- all_cors %>% filter(pos_group == "DL") %>% pull(cor)
rb_cor <- all_cors %>% filter(pos_group == "RB") %>% pull(cor)

cat(paste0(
  "**QB has the weakest market signal** (r = ", round(qb_cor, 3),
  ") — the market struggles most to price quarterbacks, which makes sense given the ",
  "position's boom-or-bust nature. **DL is the most efficiently priced** (r = ",
  round(dl_cor, 3), "). RB sits at **", round(rb_cor, 3),
  "** — one of the strongest correlations.\n\n",
  "This is the plot twist we didn't expect. We came in hoping to argue that RBs are ",
  "undervalued by the market — that the low Sharpe was an artifact of underpayment, not ",
  "real low value. But the data says the opposite: **the market prices RBs accurately.** ",
  "The low Sharpe ratio for RBs isn't a market distortion. It's real.\n\n",
  "But here's why this *strengthens* arguments 2 and 3: if the market is efficient, then ",
  "the low Sharpe reflects the true average value of the position. The key word is *average*. ",
  "Blue-chip RBs in thin classes are the exceptions to the average — and because the market ",
  "is efficient, we can trust that the blue chips who *do* get paid earned it."
))
```

---

## When to Follow vs Override the Sharpe Ratio

We've tested four arguments against the model. Three cracked it open; one reinforced it. Here's the decision framework that emerges:

<div style="background-color: #f0f4f8; border: 2px solid #377EB8; padding: 20px 25px; margin: 20px 0; border-radius: 8px; font-family: monospace; font-size: 14px; line-height: 2;">

<span style="font-family: sans-serif; font-weight: bold; font-size: 16px; color: #377EB8;">The Override Decision Tree</span>

**Is this position a real team need?**
&nbsp;&nbsp;&nbsp;&nbsp;No &rarr; **Follow the Sharpe ratio.** Draft the highest-value position available.

&nbsp;&nbsp;&nbsp;&nbsp;Yes &darr;

**Is this a blue-chip prospect (top 10% talent)?**
&nbsp;&nbsp;&nbsp;&nbsp;No &rarr; **Follow the Sharpe ratio.** Average prospects regress to average outcomes.

&nbsp;&nbsp;&nbsp;&nbsp;Yes &darr;

**Is the class thin at this position?**
&nbsp;&nbsp;&nbsp;&nbsp;No &rarr; **Follow the Sharpe ratio.** Comparable talent will be available later.

&nbsp;&nbsp;&nbsp;&nbsp;Yes &darr;

**OVERRIDE the Sharpe ratio. Take the blue chip.**

</div>

The framework is deliberately conservative. You need *all three* conditions to override: a real need, a blue-chip prospect, and a thin class. Miss any one, and the model's general advice holds.

Most of the time, the Sharpe ratio is right. The exceptions are narrow — but when they occur, they're the picks that define drafts.

---

## What We Still Believe

Not everything from Part 1 fell apart. Here's what survived the stress test:

- **QB is still the best top-10 investment on average.** Nothing we found changes the overall positional hierarchy. If you need a QB and there's a blue-chip one available, take him without hesitation.

- **Late 1st round is still the sweet spot for OT and EDGE.** The combination of high hit rates and strong second-contract value makes picks 15-32 the most efficient draft real estate for these positions.

- **Day 3 RBs are still fine most of the time.** The cheap FA replacement cost means you can fill the position without premium capital. When the class isn't deep and the prospect isn't blue-chip, this is still the right call.

- **The aggregate advice is right; context determines the exceptions.** The Sharpe ratio is a compass, not a GPS. It points you in the right direction. But the last mile requires scouting, roster analysis, and class evaluation that no single metric can capture.

- **The market is mostly efficient.** This validates the Sharpe framework rather than undermining it. Second-contract value reflects real performance, not market noise. When the model says a position has low returns, believe it — then check whether you're looking at the average or the exception.

---

## Methodology Notes

**Data:** Player stats from Pro Football Reference via `nflreadr`. Contract data from OverTheCap via `nflreadr`. All values normalized to percentage of salary cap at time of signing. Analysis covers 2010-2022 draft classes (minimum 4 NFL seasons for fair evaluation).

**Blue-chip classification:** Top 10% of career approximate value within each position group, across all drafts since 1980.

**Class depth:** A class is "deep" if its hit rate exceeds the position's median hit rate across all years in the sample. "Thin" otherwise.

**Roster needs:** A position is "filled" if the team had a starter playing above the hit threshold at that position in the draft year. A position is a "need" if it wasn't filled.

**Market efficiency:** Correlation between second-contract APY (% of cap) and post-contract average snap percentage. Higher correlation indicates the market accurately prices player value.

**Part 1:** Full methodology and interactive charts at [stranger9977.github.io/draft-sharpe-analysis](https://stranger9977.github.io/draft-sharpe-analysis/)

*Built with R, ggplot2, nflplotR, and gt. Data from nflverse and Pro Football Reference.*
